---
title: "Chloride Benchmarks"
author: "S. Berzins, A. Thompson"
date: "2026-03-31"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev = "pdf",           # Save figures as PDF
  fig.path = "C://Users//sberzin//OneDrive - Oregon//Desktop//figures"  # Directory where figures will be saved
)
```
# *Chloride*

# **1 - LOAD PACKAGES**

```{r load packages, warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(maps)
library(RColorBrewer)
library(htmlwidgets)
```

# **2 - IMPORT DATA**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r import stations}
chlor <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-02-10.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Chloride') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>% 
  mutate(Chloride = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

comids <- chlor %>% 
  select(COMID) # Get just the COMIDs for next step
```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

##### Learn about StreamCat predictors: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

##### View StreamCat updates: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-updates>

### OPTION 1: pull directly from StreamCat

#### Run function to query ALL StreamCat data for selected COMIDs by subcategory

```{r streamcat function}
# NOTE: StreamCat doesn't pull data for 'other' areas of interest when aoi='watershed, catchment, other'. 
# Instead, use aoi='watershed' and run separate line for aoi='other' since they don't pull unless on their own.
# ALSO: Do not separate metric lists onto a new line by hitting the enter button, as any metric after the break will be excluded from the pull.

# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture","Climate","Dams","Flow","IWI","Land Cover","Lithology","Mining_Toxics","Roads",
                                           "Urban","Wildfire","Others")){
type <- match.arg(type)

if(type == "Agriculture"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslphigh2001,pctagslpmid2001,pestic1997,sw_flux,waterinput,wdrw_LD,pctcrop2001,pcthay2001,cbnf,fert,manure,nani,nsurp,agkffact',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip2008,precip8110,precip9120,tmax8110,tmax9120,tmean2008,tmean8110,tmean9120,tmin8110,tmin9120,inorgnwetdep2008,nh42008,no32008,sn2008,precip_minus_evt,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens,damnidstor,damnrmstor,NABD_Dens,NABD_NIDStor,NABD_NrmStor', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev,runoff', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,habt,hyd,sed,temp', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctdecid2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturblo2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'al2o3,cao,compstrgth,fe2o3,hydrlcond,k2o,mgo,na2o,p2o5,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,sio2,clay,kffact,om,perm,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'huden2010,npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,wwtpminordens,pctimp2001,pctimpslphigh2001,pctimpslpmid2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctburnarea2002,pcthighsev2002,pctincvegresp2002,pctlowsev2002,pctmodsev2002,pctnonprocmask2002,pctundsev2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfulldepth,bankfullwidth,ici,iwi,mast2008,msst2008,mwst2008,thalwegdepth,wettedwidth,prg_bmmi0809', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

#### Pull ALL StreamCat data by subcategory and merge

```{r streamcat pull}
# # PULL DATA BY SUBCATEGORY/TYPE
# ag <- get_streamcat(comids = comids, type = "Agriculture")
# clim <- get_streamcat(comids = comids, type = "Climate")
# dams <- get_streamcat(comids = comids, type = "Dams")
# flow <- get_streamcat(comids = comids, type = "Flow")
# iwi <- get_streamcat(comids = comids, type = "IWI") 
# cover <- get_streamcat(comids = comids, type = "Land Cover")
# litho <- get_streamcat(comids = comids, type = "Lithology") 
# mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
# roads <- get_streamcat(comids = comids, type = "Roads")
# urb <- get_streamcat(comids = comids, type = "Urban")
# fire <- get_streamcat(comids = comids, type = "Wildfire")
# other <- get_streamcat(comids = comids, type = "Others")
# 
# # MERGE INTO ONE TABLE
# sc_all <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
# sc_all <- sc_all %>% reduce(full_join, by = 'COMID')
# 
# rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'roads', 'urb', 'fire', 'other'))
```

### OPTION 2: Read in SC data from pre-downloaded file.

```{r streamcat flat file}
sc_all <- read_excel("~/BioMonORDEQ/Benchmarks/Water Chemistry/SC_all_comids2026-04-14.xlsx")
sc_all <- inner_join(sc_all, comids)  # only keep CHLOR
```

## C. OTHER PREDICTORS - *OUTSIDE OF STREAMCAT*

#### Erodible geology, NHD slope, stream power

```{r other predictors}
# ERODIBLE VS RESISTENT GEOLOGY (KEPT ERODIBLE, SINCE RESISTENT IS INVERSE OF ERODIBLE)
erod_resist_strmpow <- sc_all %>%
  mutate(PCT_EROD = PCTALLUVCOASTWS+PCTCARBRESIDWS+PCTCOASTCRSWS+PCTCOLLUVSEDWS+PCTEOLCRSWS+PCTEOLFINEWS+PCTGLACLAKECRSWS+PCTGLACLAKEFINEWS+PCTGLACTILCLAYWS+PCTGLACTILCRSWS+PCTGLACTILLOAMWS+PCTSALLAKEWS+PCTNONCARBRESIDWS)  %>% 
  mutate(PCT_RESIST = PCTALKINTRUVOLWS+PCTEXTRUVOLWS+PCTHYDRICWS+PCTSILICICWS) 

# SLOPE FROM NHD (TAKES 1-2 MIN)
# Note: do NOT push the downloaded file to GitHub as it's too large and will result in issues. Commit the ignore instead.
erod_resist_strmpow <- ORDEQBioassessment::get_NHD_info(erod_resist_strmpow) 

# WATERSHED AREA AND STREAM POWER
erod_resist_strmpow <- erod_resist_strmpow %>% 
  mutate(STRMPOW_CAT = totdasqkm + PRECIP9120WS * slope) %>%  # WSAREASQKM from SC, instead pulling totdasqkm from get_NHD (they are the same)
  select(COMID, PCT_EROD, STRMPOW_CAT, NHD_pSLOPE, SITE_TYPE) # Drop PCT_RESIST since it's the inverse of PCT_ERODE

# JOIN WITH OTHER PARAMETERS NOT FROM STREAMCAT
chlor_vars <- sc_all %>% 
  inner_join(erod_resist_strmpow, by='COMID') %>% # Join StreamCat data with erod_resist_strmpow data calulated above
  inner_join(chlor, by = 'COMID') # Join those with columns from AWQMS.

# RE-ORDER COLUMNS 
chlor_vars <- chlor_vars %>% 
  relocate(MLocID:Chloride, .before = 'COMID') %>% 
  relocate(L2Eco:L3Eco, .after = SITE_TYPE) %>% 
  select(-c(HUC8, EcoRegion4)) 
```

#### Distance to coast
```{r dist to coast}
# Done in ArcGIS Pro. Distance is in km.
coast_dist <- read_excel("//deqlab1//GIS_WA//Personal_Working_Folders//SBerzins//Dist_to_coast_All_sites.xls") %>% select(-c("OBJECTID", "StationDes","Lat_DD", "Long_DD", "NEAR_FID")) %>% rename(Coast_dist = NEAR_DIST) 

# Join to chloride table
chlor_vars <- inner_join(coast_dist, chlor_vars, by = "MLocID") %>% 
  relocate(Coast_dist, .before = "PCT_EROD")
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r handle NAs}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(chlor_vars, function(x) sum(is.na(x))))

# THEN WRITE CODE FOR THEM IN ROUGH FIX SECTION BELOW.

# Note: Some predictors may have a large number of missing values (e.g., bankfull width, n = 51). Consider performing rough fix only for variables with ~10 or fewer missing values. This will avoid introducing error in the correlation matrix and variable selection phase.
```

#### PERFORM NA ROUGH FIX

```{r NA roughfix}
# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH ONLY A FEW MISSING VALUES (Leave metrics with >10 NAs [MAST2008, BMMI0809, WETTEDWITDTH, THALWEGDEPTH, BANKFULLWIDTH, BANKFULLDEPTH] until after the correlation table is built.)
chlor_vars$NSURPWS  <- na.roughfix(chlor_vars$NSURPWS) # 1 NA
chlor_vars$NANIWS  <- na.roughfix(chlor_vars$NANIWS) # 1 NA
chlor_vars$SW_FLUXWS  <- na.roughfix(chlor_vars$SW_FLUXWS) # 2 NA
chlor_vars$PCTIMPSLPHIGH2001WS  <- na.roughfix(chlor_vars$PCTIMPSLPHIGH2001WS) # 4 NA
chlor_vars$MWST2008  <- na.roughfix(chlor_vars$MWST2008) # 6 NA
```

## B. PREDICTOR VARIABLE SELECTION

#### Make Spearman correlation matrix to identify co-varied predictors (>0.8)

```{r spearman matrix, warning=FALSE}
# Subset numeric variables only
chlor_vars_num <- chlor_vars %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Chloride)) %>%  
  relocate(sort(names(.))) # sort columns alphabetically

# FULL CORRELATION MATRIX
spear <- cor(chlor_vars_num, method = "spearman", use = "pairwise.complete.obs")
spear <- as.data.frame(spear)
spear.rows <- tibble::rownames_to_column(spear)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix <- spear.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear', 'spear.rows'))
```

#### Export matrix and clean up in Excel

```{r export spearman matrix}
# Export once, then clean up in Excel.
#write_xlsx(matrix, path = paste0("~/BioMonORDEQ/Benchmarks/Water Chemistry/Chloride/Chloride_Spearmans_Matrix", Sys.Date(), ".xlsx"))
```

#### Calculate coefficient of variance (CV), proportion of 0s, and kernel density plots to guide variable selection decisions

```{r coefficient of variance}
# Calculate coefficient of variance (CV) for each predictor.
cv <- chlor_vars_num %>% 
  summarise(across(everything(), ~ cv(.x, na.rm = TRUE))) %>% 
  pivot_longer(1:141, names_to="Var", values_to="CV") # Change first argument to 1:# columns

# Calculate % of zeros for each predictor.
cv_pctzero <- chlor_vars_num %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE) / n() * 100)) %>% 
  pivot_longer(1:141, names_to="Var", values_to="Percent Zeros") %>%  # Change  first argument to 1:# columns
  inner_join(cv, by = "Var")

# Write output to Excel for variable selection meeting.
#write_xlsx(cv_pctzero, path = "~/BioMonORDEQ/Benchmarks/Water Chemistry/Chloride/Chloride_CV.xlsx")
rm(cv)

# Make kernel density plots for predictors of interest to compare distributions.
chlor_vars %>%
    select("PRECIP_MINUS_EVTWS", "PCTCONIF2001WS", "PESTIC1997WS", "PCTBL2001WS", "ROCKNWS", "WTDEPWS", "RDDENSWS", "RCKDEPWS", "KFFACTWS", "SEPTICWS") %>%  # Edit this line to include vars that you want to look at.
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  geom_rug()
```

#### Hold variable selection meeting

```{r var selection info}
# We did this as a team 4/3/2026.

# 1. Make a variable selection decisions sheet in Excel based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection decisions.xlsx) and save to model folder on GitHub.

# 2. Clean up correlation matrix Excel table manually:
  # a) Filter data and delete rows where cor = 1.
  # b) Sort row names alphabetically.
  # c) Remove composite indexes (CHEM, CONN, HABT, HYD, ICI, IWI, SED, and TEMP).
  # d) Make a new matrix predictor decisions sheet based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection correlations) and save to model folder on GitHub. 
  #     Copy row names into template (col A) and update formulas, column names into col G, and correlations into col L. 
  #     Add horizontal borders between groupings. 
  #     You may want to hide columns B and H (they are just for the formulas)

# 3. For each co-varied predictor(s), discuss ecological considerations and data distribution (CV, how well covers gradient, %0s).

# 4. Record which predictors to keep/remove and document rationale in the decisions sheet.

# 5. Remove predictors that have 100% zeros (no modeling power) and paste list of keepers below.
```

#### Update predictor list

```{r create predictor list}
# Use Excel formula (=TEXTJOIN(", ", TRUE, range:range)) to compile list from decisions sheet and paste below.
chlor_vars_selected <- chlor_vars %>% 
  select(c(MLocID:COMID,BFIWS, CANALDENSWS, CAOWS, CLAYWS, COMPSTRGTHWS, DAMDENSWS, ELEVWS, FERTWS, HYDRLCONDWS, KFFACTWS, MINEDENSWS, MSST2008, NA2OWS, NANIWS, NH42008WS, NHD_pSLOPE, NPDESDENSWS, NWS, OMWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTALKINTRUVOLWS, PCTALLUVCOASTWS, PCTBL2001WS, PCTBURNAREA2002WS, PCTCARBRESIDWS, PCTCOLLUVSEDWS, PCTCONIF2001WS, PCTCROP2001WS, PCTDECID2001WS, PCTEOLFINEWS, PCTEXTRUVOLWS, PCTFRSTLOSS2002WS, PCTGLACLAKECRSWS, PCTGLACLAKEFINEWS, PCTGLACTILCRSWS, PCTGRS2001WS, PCTHBWET2001WS, PCTICE2001WS, PCTIMP2001WS, PCTINCVEGRESP2002WS, PCTMXFST2001WS, PCTNONAGINTRODMANAGVEGWS, PCTOW2001WS, PCTSALLAKEWS, PCTSHRB2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PCTWATERWS, PCTWDWET2001WS, PESTIC1997WS, POPDEN2010WS, PRECIP_MINUS_EVTWS, RCKDEPWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, ROCKNWS, RUNOFFWS, SANDWS, SEPTICWS, STRMPOW_CAT, SUPERFUNDDENSWS, SWS, TMAX9120WS, TRIDENSWS, WETINDEXWS, WTDEPWS, WWTPALLDENSWS, WWTPMAJORDENSWS, Coast_dist,SITE_TYPE:L3Eco)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCHLOR=log(Chloride)) %>% 
  relocate(logCHLOR, .before = Chloride)
```

#### Perform final correlation check to ensure no co-varied predictors remain

```{r final correlation check, warning=FALSE}
# Subset numeric variables only
chlor_vars_num_selected <- chlor_vars_selected %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Chloride)) 

# FULL CORRELATION MATRIX
spear2 <- cor(chlor_vars_num_selected, method = "spearman", use = "pairwise.complete.obs")
spear2 <- as.data.frame(spear2)
spear2.rows <- tibble::rownames_to_column(spear2)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix2 <- spear2.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear2', 'spear2.rows'))

# ***OPEN NEWLY CREATED MATRIX AND CONTINUE ONLY IF NO CORRELATED VARIABLES REMAIN (ALL SHOULD BE 1)***
```

#### Clear out the environment

```{r clean up environment}
rm(list=c('cv_pctzero', 'erod_resist_strmpow', 'matrix', 'matrix2', 'comids', 'chlor_vars_num','chlor_vars_num_selected', 'coast_dist'))
```

#### Perform final NA check and perform rough fix as desired

```{r final NA check}
# FIND OUT WHICH COLUMNS HAVE NAs
sapply(chlor_vars_selected, function(x) sum(is.na(x)))

# DO ANY LAST NA ROUGH FIXES AS DESIRED (write code for specific predictors) 
# none for chloride
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r plot response variable}
# HISTOGRAM
hist(chlor_vars_selected$Chloride, main = "", xlab = "Chloride (mg/L)") #raw data
hist(chlor_vars_selected$logCHLOR, main = "", xlab = "log(Chloride)", breaks = 15)

# BOX PLOTS
boxplot(chlor_vars_selected$Chloride, ylab = "Chloride (mg/L)")
boxplot(chlor_vars_selected$logCHLOR, ylab = "log(Chloride)")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r prep response data}
# Select response data of interest
CHLOR <- chlor_vars_selected %>%
  select(-c(MLocID:COMID, Chloride, SampleStartDate)) %>%  #drop Chloride in favor of logCHLOR. Drop AWQMS columns too.
  as.data.frame() # Make sure CHLOR is a data frame, not a tibble

# Convert character columns to factors
CHLOR[sapply(CHLOR, is.character)] <- lapply(CHLOR[sapply(CHLOR, is.character)], as.factor)

# # Order factors
# CHLOR$L2Eco <- factor(CHLOR$L2Eco, levels = c("Marine West Coast Forest", "Western Cordillera", "Cold Deserts"))
# CHLOR$L3Eco <- factor(CHLOR$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r full model tuning}
#Adjust the parameters below based on tuning
full_tuning <- tune(logCHLOR~., data = CHLOR, seed = 1234)
full_tuning$optimal
```

#### Build full model

```{r full model}
full <- rfsrc(logCHLOR ~ ., data = CHLOR, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 160)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(CHLOR$logCHLOR, na.rm = TRUE))
print(full_per.var)

range_CHLOR <- max(exp(CHLOR$logCHLOR)) - min(exp(CHLOR$logCHLOR))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(CHLOR$logCHLOR))
full_rmse.scaled <- full_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(full_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(CHLOR$logCHLOR))
full_oob.rmse.scaled <- full_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = CHLOR$logCHLOR, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =CHLOR$logCHLOR, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR Full Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```

#### Leaflet map of full model

```{r leaflet map of full model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(CHLOR$ELEVWS))  # Elevation works.
# any(duplicated(CHLOR$PRECIP9120WS))

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, CHLOR) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

full_awqms <- chlor_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logCHLOR) 

# Join pred with awqms table, and makes sure the CHLOR results match (Make sure logCHLOR column is the same as logCHLOR.1)
full_map <- cbind(full_obs.predict,full_awqms)

# Then select to remove duplicate logCHLOR column and reorder.
full_map <- full_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Observed CHLOR: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$pred), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Predicted CHLOR: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Obs - Pred CHLOR: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$obs, position = "bottomright" , title = "Observed CHLOR" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = data$obs, position = "bottomright" , title = "Predicted CHLOR" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred CHLOR" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map$obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map$delta),max(full_map$delta)))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//full_model_map.html", selfcontained = TRUE, title = "Full model")
```

#### Partial dependence plots for full model

```{r partial dependence plots for full model}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c("ELEVWS", "TMAX9120WS", "CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS", "PCTCONIF2001WS", "SEPTICWS", "PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c("PCTMXFST2001WS", "Coast_dist", "HYDRLCONDWS", "MSST2008", "PCTURBOP2001WS", "CANALDENSWS", "OMWS", "RUNOFFWS"), partial = TRUE, plots.one.page = TRUE)

par(mar = c(10, 5, 2, 2))
partialplots3 <- plot.variable(full, xvar.names = c("L2Eco", "L3Eco"), partial = TRUE, las = 2, cex.axis = 0.7, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("SEPTICWS", "PCTFRSTLOSS2002WS", "PCTURBOP2001WS", "CANALDENSWS", "PCTNONAGINTRODMANAGVEGWS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Another way to do partial plots (slower but more customizable)
```{r more partial plots for full model}
# partial.obj <- partial(full, partial.xvar = "ELEVWS", partial.values = full$xvar$ELEVWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "ELEVWS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "TMAX9120WS", partial.values = full$xvar$TMAX9120WS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "TMAX9120WS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "CLAYWS", partial.values = full$xvar$CLAYWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "CLAYWS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "PRECIP_MINUS_EVTWS", partial.values = full$xvar$PRECIP_MINUS_EVTWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "PRECIP_MINUS_EVTWS", ylab = "ŷ")

```

#### Leaflet map of predictors  
```{r leaflet map of selected predictors}
map_predictors <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "ELEVWS", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$ELEVWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "ELEVWS: ", "</strong>", data$ELEVWS))  %>% 
      addCircleMarkers(data = data, group = "TMAX9120WS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$TMAX9120WS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "TMAX9120WS: ", "</strong>", data$TMAX9120WS))  %>% 
     addCircleMarkers(data = data, group = "CLAYWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$CLAYWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "CLAYWS: ", "</strong>", data$CLAYWS))  %>% 
     addCircleMarkers(data = data, group = "PRECIP_MINUS_EVTWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$PRECIP_MINUS_EVTWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "PRECIP_MINUS_EVTWS: ", "</strong>", data$PRECIP_MINUS_EVTWS))  %>% 
     addCircleMarkers(data = data, group = "PCTBL2001WS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal5(data$PCTBL2001WS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "PCTBL2001WS: ", "</strong>", data$PCTBL2001WS))  %>% 
    hideGroup(c("TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS")) %>%  # Don't check all layers by default
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("ELEVWS", "TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "ELEVWS", values = data$ELEVWS, position = "bottomright" , title = "ELEVWS" ) %>% 
    addLegend(pal = pal2,group = "TMAX9120WS",values = data$TMAX9120WS, position = "bottomright" , title = "TMAX9120WS" ) %>%  
    addLegend(pal = pal3, group = "CLAYWS", values = data$CLAYWS, position = "bottomleft" , title = "CLAYWS" ) %>% 
    addLegend(pal = pal4, group = "PRECIP_MINUS_EVTWS", values = data$PRECIP_MINUS_EVTWS, position = "bottomleft" , title = "PRECIP_MINUS_EVTWS" ) %>% 
    addLegend(pal = pal5, group = "PCTBL2001WS", values = data$PCTBL2001WS, position = "bottomleft" , title = "PCTBL2001WS" )
}

pal <- colorNumeric(palette = "Oranges", 
  domain = c(min(full_map$ELEVWS),max(full_map$ELEVWS)))
pal2 <- colorNumeric(palette = "RdBu", 
  domain = c(min(full_map$TMAX9120WS), max(full_map$TMAX9120WS)), reverse = TRUE)
pal3 <- colorNumeric(palette = "Reds", 
                     domain = c(min(full_map$CLAYWS), max(full_map$CLAYWS)))
pal4 <- colorNumeric(palette = "Blues", 
  domain = c(min(full_map$PRECIP_MINUS_EVTWS), max(full_map$PRECIP_MINUS_EVTWS)))
pal5 <- colorNumeric(palette = "Greys", 
  domain = c(min(full_map$PCTBL2001WS), max(full_map$PCTBL2001WS)))

map_predictors(full_map)

# Save map as html
# predictors_map <- map_predictors(full_map)
# saveWidget(predictors_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//predictors_map.html", selfcontained = TRUE, title = "Predictors map")
```

#### Residual plots
```{r full model residual plots}
# Plot residuals 
full_resid <- exp(CHLOR$logCHLOR) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = "Predicted CHLOR (mg/L)", ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(CHLOR$logCHLOR) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r adj full model}
# Reduce all human-influenced vars to all 0s.
CHLOR_no.hi <- CHLOR %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMDENSWS", "FERTWS", "MINEDENSWS", "NANIWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= CHLOR_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(CHLOR_no.hi$logCHLOR, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(CHLOR_no.hi$logCHLOR))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = CHLOR_no.hi$logCHLOR, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR Full Model predictions without human influenced vars",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x =0.9, y = 100, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(CHLOR_no.hi$logCHLOR) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = "Predicted CHLOR (mg/L)", ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map adjusting human vbls to 0

```{r leaflet map adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Join predictions to cal_val table
full_obs.predict_no.hi <- cbind(full_plotdf_no.hi, CHLOR_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join CHLOR.obs_predict_no.hi with CHLOR.awqms, and makes sure the results match.
full_map_no.hi <- cbind(full_awqms,full_obs.predict_no.hi)

# Then select to remove duplicate logCHLOR column and reorder.
full_map_no.hi <- full_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, max(full_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, max(full_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map_no.hi$delta), max(full_map_no.hi$delta)))
map_function(full_map_no.hi)

# Save map as html
# full_model_map_no_hi <- map_function(full_map_no.hi)
# saveWidget(full_model_map_no_hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//full_model_map_no_hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

## C. Model Reduction

#### Build model with Top 10 predictors

```{r reduced model tuning}
#Adjust the parameters below based on tuning 
red10_tuning <- tune(logCHLOR ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS + L2Eco + PRECIP_MINUS_EVTWS + PCTBL2001WS + PCTCONIF2001WS + SEPTICWS + PCTFRSTLOSS2002WS , data = CHLOR, seed = 1234)
red10_tuning$optimal
```

```{r reduced model}
red10 <- rfsrc(logCHLOR ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS + L2Eco + PRECIP_MINUS_EVTWS + PCTBL2001WS + PCTCONIF2001WS + SEPTICWS + PCTFRSTLOSS2002WS , data = CHLOR, mtry = red10_tuning$optimal[[2]], nodesize = red10_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red10 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red10_oob.mse <- tail(red10$err.rate,1)
red10_per.var <- 100 * (1 - red10_oob.mse/ var(CHLOR$logCHLOR, na.rm = TRUE))
print(red10_per.var)

#In-bag RMSE
red10_rmse<- rmse(exp(red10$predicted), exp(CHLOR$logCHLOR))
red10_rmse.scaled <- red10_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(red10_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(red10_rmse.scaled*100,2), "%"))

#OOB RMSE
red10_oob.rmse<- rmse(exp(red10$predicted.oob), exp(CHLOR$logCHLOR))
red10_oob.rmse.scaled <- red10_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(red10_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(red10_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red10$err.rate

# VIP
plot(red10, plots.one.page=FALSE)
red10_vimp <- sort(red10$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red10_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR Reduced (10 variables) model VIMP")

#print(vimp(red10)$importance)

# Make a df for plots
red10_pred.oob <- data.frame(obs = CHLOR$logCHLOR, pred = red10$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red10_pred.ib <- data.frame(obs =CHLOR$logCHLOR, pred = red10$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red10_plotdf <- rbind(red10_pred.oob, red10_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red10_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR Red10 Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.7, y = 60, label = paste0('OOB RMSE = ' , round(red10_oob.rmse,2), ' | scaled RMSE = ', round(red10_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.7, y = 77, label = paste0('IB RMSE = ' , round(red10_rmse,2), ' | scaled RMSE = ', round(red10_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.7, y = 100, label = paste0('% variance explained = ' , round(red10_per.var,2), '%'), size = 3) 
```

#### Leaflet map of red10 model

```{r leaflet map of reduced model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(CHLOR$ELEVWS))  # Elevation works.
# any(duplicated(CHLOR$PRECIP9120WS))

# Join predictions to StreamCat table
red10_obs.predict <- cbind(red10_plotdf, CHLOR) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

red10_awqms <- chlor_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logCHLOR) 

# Join obs.pred with awqms table, and makes sure the CHLOR results match (Make sure logCHLOR column is the same as logCHLOR.1)
red10_map <- cbind(red10_obs.predict,red10_awqms)

# Then select to remove duplicate logCHLOR column and reorder.
red10_map <- red10_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, red10_map$obs))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, red10_map$obs))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(red10_map$delta),max(red10_map$delta)))

map_function(red10_map)

# Save map as html
# red10_model_map <- map_function(red10_map)
# saveWidget(red10_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//red10_model_map.html", selfcontained = TRUE, title = "red10 model")
```

#### Overlay reduced and full models

```{r reduced and full model prediction plot}
full_red_map <- left_join(full_map, red10_map %>% select(MLocID, pred), by = "MLocID") %>% 
  mutate(full_pred = pred.x) %>% 
  mutate(red10_pred = pred.y) %>% 
  select(-c(pred.x,pred.y)) %>% 
  relocate(red10_pred, .after = obs) %>% 
  relocate(full_pred, .after = obs) %>% 
  select(MLocID:predictions) %>% 
  gather(key = "Model", value = "Prediction", full_pred, red10_pred)

ggplot(full_red_map, aes(x = Prediction, y = obs, color = Model)) +
  geom_point(size = 1.5, alpha = 0.5) + theme_bw(base_size = 15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Predicted CHLOR (mg/L)",
    y = "Observed CHLOR (mg/L)") + scale_x_log10() + scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```

#### Distributions for Human-influenced vars at all reference sites

```{r distributions of human influenced variables at all ref sites}
# Select only the human influenced predictors at reference sites from the reduced model.
CHLOR_ref_hi <- full_map %>% 
  filter(ReferenceSite == "REFERENCE") %>% 
  select(c(MLocID:logCHLOR, SEPTICWS, PCTFRSTLOSS2002WS, PCTURBOP2001WS, CANALDENSWS, PCTNONAGINTRODMANAGVEGWS))

# PCTFRSTLOSS2002WS
par(mfrow=c(1,2))
plot.variable(red10, xvar.names = c("PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(CHLOR_ref_hi$PCTFRSTLOSS2002WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(CHLOR_ref_hi$PCTFRSTLOSS2002WS, main = '')
quantile(CHLOR_ref_hi$PCTFRSTLOSS2002WS, probs = seq(.1, .9, by = .1))

# SEPTICWS
par(mfrow=c(1,2))
plot.variable(red10, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(CHLOR_ref_hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(CHLOR_ref_hi$SEPTICWS, main = '')
quantile(CHLOR_ref_hi$SEPTICWS, probs = seq(.1, .9, by = .1))

# BELOW not in reduced model

# # PCTURBOP2001WS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("PCTURBOP2001WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(CHLOR_ref_hi$PCTURBOP2001WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(CHLOR_ref_hi$PCTURBOP2001WS, main = '')
# quantile(CHLOR_ref_hi$PCTURBOP2001WS, probs = seq(.1, .9, by = .1))
# 
# # CANALDENSWS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("CANALDENSWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(CHLOR_ref_hi$CANALDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(CHLOR_ref_hi$CANALDENSWS, main = '')
# quantile(CHLOR_ref_hi$CANALDENSWS, probs = seq(.1, .9, by = .1))
# 
# # PCTNONAGINTRODMANAGVEGWS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("PCTNONAGINTRODMANAGVEGWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(CHLOR_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(CHLOR_ref_hi$PCTNONAGINTRODMANAGVEGWS, main = '')
# quantile(CHLOR_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = seq(.1, .9, by = .1))
```

#### Reduce all human-influenced variables [to 0]* and make predictions using reduced model.
##### * We may decide to adjust to a non-zero percentile based on reference quantiles in the chunk above.

```{r adj reduced model}
# Make predictions with adjusted human-influence to 0 
red10_no.hi <- predict(red10, newdata= CHLOR_no.hi, seed = 24)

# Mean squared error (untransformed)
red10_mse_no.hi <- tail(red10_no.hi$err.rate,1)
red10_per.var_no.hi <- 100 * (1 - red10_mse_no.hi/ var(CHLOR_no.hi$logCHLOR, na.rm = TRUE))

# RMSE
red10_rmse_no.hi<- rmse(exp(red10_no.hi$predicted), exp(CHLOR_no.hi$logCHLOR))
red10_rmse.scaled_no.hi <- red10_rmse_no.hi/range_CHLOR
print(paste0('OOB RMSE = ' , round(red10_rmse_no.hi,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(red10_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red10_plotdf_no.hi <- data.frame(obs = CHLOR_no.hi$logCHLOR, pred = red10_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red10_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR red10 Model predictions without human influenced vars",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.6, y = 90, label = paste0('OOB RMSE = ' , round(red10_rmse_no.hi,2), ' | scaled RMSE = ', round(red10_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red10 model, adjusted human-influenced vars to 0
red10_resid_no.hi <- exp(CHLOR_no.hi$logCHLOR) - exp(red10_no.hi$predicted) # Observed minus predicted oob
plot(red10_resid_no.hi ~ exp(red10_no.hi$predicted), xlab = "Predicted CHLOR (mg/L)", ylab = "Residuals (obs - pred)", main = "Red10 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map after adjusting human vbls for reduced10 model

```{r leaflet map after adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Join predictions to cal_val table
red10_obs.predict_no.hi <- cbind(red10_plotdf_no.hi, CHLOR_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join CHLOR.obs_predict_no.hi with CHLOR.awqms, and makes sure the results match.
red10_map_no.hi <- cbind(full_awqms,red10_obs.predict_no.hi)

# Then select to remove duplicate logCHLOR column and reorder.
red10_map_no.hi <- red10_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red10_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red10_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1* max(red10_map_no.hi$delta), max(red10_map_no.hi$delta)))

map_function(red10_map_no.hi)

# Save map as html
# red10_model_map_no.hi <- map_function(red10_map_no.hi)
# saveWidget(red10_model_map_no.hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//red10_model_map_no.hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```
#### Partial dependence plots for red10 model

```{r partial dependence plots for reduced model}
# Top importance vars
red10_partialplots <- plot.variable(red10, xvar.names = c("ELEVWS", "TMAX9120WS", "CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS", "PCTCONIF2001WS", "SEPTICWS", "PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)

par(mar = c(10, 5, 2, 2))
red10_partialplots2 <- plot.variable(red10, xvar.names = c("L2Eco", "L3Eco"), partial = TRUE, las = 2, cex.axis = 0.7, plots.one.page = TRUE)

# Human-influenced vars
red10_partialplots_hi <- plot.variable(red10, xvar.names = c("SEPTICWS", "PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Top 4 predictors (not using this one)

```{r reducing the model further tuning}
#Adjust the parameters below based on tuning
red4_tuning <- tune(logCHLOR ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = CHLOR, seed = 1234)
red4_tuning$optimal
```

```{r reducing the model further}
red4 <- rfsrc(logCHLOR ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = CHLOR, mtry = red4_tuning$optimal[[2]], nodesize = red4_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red4 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red4_oob.mse <- tail(red4$err.rate,1)
red4_per.var <- 100 * (1 - red4_oob.mse/ var(CHLOR$logCHLOR, na.rm = TRUE))
print(red4_per.var)

#In-bag RMSE
red4_rmse<- rmse(exp(red4$predicted), exp(CHLOR$logCHLOR))
red4_rmse.scaled <- red4_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(red4_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(red4_rmse.scaled*100,2), "%"))

#OOB RMSE
red4_oob.rmse<- rmse(exp(red4$predicted.oob), exp(CHLOR$logCHLOR))
red4_oob.rmse.scaled <- red4_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(red4_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(red4_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red4$err.rate

# VIP
plot(red4, plots.one.page=FALSE)
red4_vimp <- sort(red4$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red4_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR Reduced (4 variables) model VIMP")

#print(vimp(red4)$importance)

# Make a df for plots
red4_pred.oob <- data.frame(obs = CHLOR$logCHLOR, pred = red4$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red4_pred.ib <- data.frame(obs =CHLOR$logCHLOR, pred = red4$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red4_plotdf <- rbind(red4_pred.oob, red4_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red4_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR red4 Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.7, y = 60, label = paste0('OOB RMSE = ' , round(red4_oob.rmse,2), ' | scaled RMSE = ', round(red4_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.7, y = 75, label = paste0('IB RMSE = ' , round(red4_rmse,2), ' | scaled RMSE = ', round(red4_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.7, y = 100, label = paste0('% variance explained = ' , round(red4_per.var,2), '%'), size = 3) 
```

## E. Reference-Only Model

#### Subset the data

```{r subset reference only data}
CHLOR.ref <- chlor_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, Chloride, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMDENSWS", "FERTWS", "MINEDENSWS", "NANIWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")) %>% # Remove anthropogenic StreamCat vars
  as.data.frame()

# Convert character columns to factors
CHLOR.ref[sapply(CHLOR.ref, is.character)] <- lapply(CHLOR.ref[sapply(CHLOR.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r tuning ref only model}
#Adjust the parameters below based on tuning 
ref_tuning <- tune(logCHLOR~., CHLOR.ref, seed = 2)
ref_tuning$optimal

```

#### Build full ref-only model

```{r ref only model}
ref <- rfsrc(logCHLOR ~ ., data = CHLOR.ref, mtry = ref_tuning$optimal[[2]], nodesize = ref_tuning$optimal[[1]], importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(CHLOR.ref$logCHLOR, na.rm = TRUE))
print(ref_per.var)

ref_range_CHLOR <- max(exp(CHLOR.ref$logCHLOR)) - min(exp(CHLOR.ref$logCHLOR))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(CHLOR.ref$logCHLOR))
ref_rmse.scaled <- ref.rmse/ref_range_CHLOR
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(CHLOR.ref$logCHLOR))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_CHLOR
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = CHLOR.ref$logCHLOR, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =CHLOR.ref$logCHLOR, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR Full Reference-Only Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.75, y = 15, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.75, y = 19, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.75, y = 25, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(CHLOR.ref$logCHLOR) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = "Predicted CHLOR (mg/L)", ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Partial dependence plots for full ref-only model

```{r partial dependence plots for full ref only model}
# Top importance vars
partialplots <- plot.variable(ref, xvar.names = c("BFIWS", "Coast_dist", "PCTMXFST2001WS", "ELEVWS", "TMAX9120WS", "ROCKNWS", "MSST2008", "RUNOFFWS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of Ref-Only model

```{r leaflet map of ref only model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(CHLOR$ELEVWS))  # Elevation works.
# any(duplicated(CHLOR$PRECIP9120WS))

# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, CHLOR.ref) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

ref_awqms <- chlor_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  filter(ReferenceSite=='REFERENCE') %>% 
  select(MLocID:logCHLOR) 

# Join obs.predict and awqms tables, and makes sure the CHLOR results match (Make sure logCHLOR column is the same as logCHLOR.1)
ref_map <- cbind(ref_obs.predict, ref_awqms)

# Then select to remove duplicate logCHLOR column and reorder.
ref_map <- ref_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal2 <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(ref_map$delta), max(ref_map$delta)))

map_function(ref_map)

# Save map as html
# ref_model_map <- map_function(ref_map)
# saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## F. Make comparisons of predictions

#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r comopare reduced model (adjusted and unadjusted) by ecoregion}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(MLocID = red10_map$MLocID, StationDes = red10_map$StationDes, Obs = red10_map$obs, Pred_unadj = red10_map$pred, L3Eco = as.character(red10_map$L3Eco),Pred_adj = red10_map_no.hi$pred, Ref = red10_map$ReferenceSite, Pred_full = full_map$pred)

boxplot_ref<- ref_map %>%
  select(MLocID, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="MLocID")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_unadj, Pred_adj, Pred_ref, Pred_full, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_unadj", "Pred_full", "Obs"))
boxplot_df$L3Eco <- factor(boxplot_df$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  theme_bw(base_size=14) + ylab("CHLOR (mg/L)") +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5
  )
  

# Ignore warning for 540 rows removed - these are NAs in the Pred_ref column.
```

#### Combined Leaflet map of predictions

```{r combinded leaflet map of predictions}
leaflet_df <- left_join(boxplot_df, chlor %>%  select(MLocID, Project1, Lat_DD, Long_DD), by = "MLocID") %>% 
  pivot_wider(names_from = Model, values_from = Prediction) %>% 
  relocate(Obs, .before= Pred_unadj)

map_all <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$Obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Observed CHLOR: ", "</strong>", data$Obs))  %>% 
      addCircleMarkers(data = data, group = "Pred_unadj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$Pred_unadj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted CHLOR UNADJUSTED (reduced model): ", "</strong>", data$Pred_unadj))  %>% 
     addCircleMarkers(data = data, group = "Pred_adj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$Pred_adj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted CHLOR ADJUSTED: ", "</strong>", data$Pred_adj))  %>% 
     addCircleMarkers(data = data, group = "Pred_ref",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$Pred_ref), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted CHLOR REF ONLY: ", "</strong>", data$Pred_ref))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Pred_unadj", "Pred_adj", "Pred_ref"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$Obs, position = "bottomright" , title = "Observed CHLOR" ) %>% 
    addLegend(pal = pal2,group = "Pred_unadj",values = data$Obs, position = "bottomright" , title = "Predicted CHLOR (UNADJUSTED)" ) %>%  # Keep values=data$Obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Pred_adj", values = data$Pred_adj, position = "bottomleft" , title = "Predicted CHLOR (ADJUSTED)" ) %>% 
        addLegend(pal = pal4, group = "Pred_ref", values = data$Pred_adj, position = "bottomleft" , title = "Predicted CHLOR (REF ONLY)" )    # Keep values=data$Pred_adj here (instead of data$Ref) to keep the legend colors scales the same between ref expectations.
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal3 <- colorNumeric(palette = "Oranges", c(0, max(leaflet_df$Pred_unadj)))
pal4 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_unadj)), na.color = "transparent") 

map_all(leaflet_df)

# Save map as html
# map_all <- map_all(leaflet_df)
# saveWidget(map_all, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//Chloride//map_all.html", selfcontained = TRUE, title = "Map_All")
```
 
#### Unadjusted predictions vs. adjusted human inflence 

```{r unadjusted predictions vs adj human influence}
leaflet_df$Ref <- factor(leaflet_df$Ref, levels = c("REFERENCE", "MODERATELY DISTURBED", "MOST DISTURBED"))

ggplot(leaflet_df, aes(x = Pred_unadj, y = Pred_adj, color = Ref, shape = Ref)) +
  geom_point(size = 1.5) + theme_bw(base_size = 15) +
  labs(
    x = "Reduced model predicted CHLOR (mg/L)",
    y = "Predicted CHLOR (mg/L) adjusted\n for no human disturbance predictions",
    color = "Reference Status", shape = "Reference Status") + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```
## G. EXPERIMENTATION with the models
##### Take all this with a grain of salt!  (No chloride pun intended)

#### Remove L2 and L3 Ecoregion as predictors
```{r}
# Select response data of interest
CHLOR_noEco <- chlor_vars_selected %>%
  select(-c(MLocID:COMID, Chloride, SampleStartDate, L2Eco, L3Eco, ELEVWS)) %>%  #drop Chloride in favor of logCHLOR. Drop AWQMS columns too. Drop Ecoregions.
  as.data.frame() # Make sure CHLOR is a data frame, not a tibble

# Convert character columns to factors
CHLOR_noEco[sapply(CHLOR_noEco, is.character)] <- lapply(CHLOR_noEco[sapply(CHLOR_noEco, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning
tuning <- tune(logCHLOR~., data = CHLOR_noEco, seed = 123)
tuning$optimal

full_noEco <- rfsrc(logCHLOR ~ ., data = CHLOR_noEco, mtry = tuning$optimal[[2]], nodesize = tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_noEco # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_noEco_oob.mse <- tail(full_noEco$err.rate,1)
full_noEco_per.var <- 100 * (1 - full_noEco_oob.mse/ var(CHLOR_noEco$logCHLOR, na.rm = TRUE))
print(full_noEco_per.var)

range_CHLOR <- max(exp(CHLOR_noEco$logCHLOR)) - min(exp(CHLOR_noEco$logCHLOR))

#In-bag RMSE
full_noEco_rmse<- rmse(exp(full_noEco$predicted), exp(CHLOR_noEco$logCHLOR))
full_noEco_rmse.scaled <- full_noEco_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(full_noEco_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_noEco_rmse.scaled*100,2), "%"))

#OOB RMSE
full_noEco_oob.rmse<- rmse(exp(full_noEco$predicted.oob), exp(CHLOR_noEco$logCHLOR))
full_noEco_oob.rmse.scaled <- full_noEco_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_noEco_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_noEco_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_noEco$err.rate

# VIP
plot(full_noEco, plots.one.page=FALSE)
full_noEco_vimp <- sort(full_noEco$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_noEco_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR full_noEco model VIMP")

#print(vimp(full_noEco)$importance)

# Make a df for plots
full_noEco_pred.oob <- data.frame(obs = CHLOR_noEco$logCHLOR, pred = full_noEco$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_noEco_pred.ib <- data.frame(obs =CHLOR_noEco$logCHLOR, pred = full_noEco$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_noEco_plotdf <- rbind(full_noEco_pred.oob, full_noEco_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_noEco_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR full_noEco Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_noEco_oob.rmse,2), ' | scaled RMSE = ', round(full_noEco_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_noEco_rmse,2), ' | scaled RMSE = ', round(full_noEco_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_noEco_per.var,2), '%'), size = 3) 
```

```{r}
elev_temp <- ggplot(CHLOR, aes(x = TMAX9120WS, y = ELEVWS, color = L3Eco)) + geom_point()
elev_temp
```

```{r}
# Select response data of interest
CHLOR_WC <- chlor_vars_selected %>%
  select(-c(MLocID:COMID, Chloride, SampleStartDate)) %>%  #drop Chloride in favor of logCHLOR. Drop AWQMS columns too. 
  filter(L2Eco=="Western Cordillera") %>% 
  select(-c(L3Eco,L2Eco)) %>%  
  as.data.frame() # Make sure CHLOR is a data frame, not a tibble

# Convert character columns to factors
CHLOR_WC[sapply(CHLOR_WC, is.character)] <- lapply(CHLOR_WC[sapply(CHLOR_WC, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning 
tuning <- tune(logCHLOR~., data = CHLOR_WC, seed = 123)
tuning$optimal

full_WC <- rfsrc(logCHLOR ~ ., data = CHLOR_WC, mtry = tuning$optimal[[2]], nodesize = tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_WC # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_WC_oob.mse <- tail(full_WC$err.rate,1)
full_WC_per.var <- 100 * (1 - full_WC_oob.mse/ var(CHLOR_WC$logCHLOR, na.rm = TRUE))
print(full_WC_per.var)

range_CHLOR <- max(exp(CHLOR_WC$logCHLOR)) - min(exp(CHLOR_WC$logCHLOR))

#In-bag RMSE
full_WC_rmse<- rmse(exp(full_WC$predicted), exp(CHLOR_WC$logCHLOR))
full_WC_rmse.scaled <- full_WC_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(full_WC_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_WC_rmse.scaled*100,2), "%"))

#OOB RMSE
full_WC_oob.rmse<- rmse(exp(full_WC$predicted.oob), exp(CHLOR_WC$logCHLOR))
full_WC_oob.rmse.scaled <- full_WC_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_WC_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_WC_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_WC$err.rate

# VIP
plot(full_WC, plots.one.page=FALSE)
full_WC_vimp <- sort(full_WC$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_WC_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR full_WC model VIMP")

#print(vimp(full_WC)$importance)

# Make a df for plots
full_WC_pred.oob <- data.frame(obs = CHLOR_WC$logCHLOR, pred = full_WC$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_WC_pred.ib <- data.frame(obs =CHLOR_WC$logCHLOR, pred = full_WC$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_WC_plotdf <- rbind(full_WC_pred.oob, full_WC_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_WC_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR full_WC Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_WC_oob.rmse,2), ' | scaled RMSE = ', round(full_WC_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_WC_rmse,2), ' | scaled RMSE = ', round(full_WC_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_WC_per.var,2), '%'), size = 3) 
```


```{r}
CHLOR_CP <- chlor_vars %>% 
  select(c(MLocID:COMID,CANALDENSWS,  DAMDENSWS,FERTWS,  NPDESDENSWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTCROP2001WS,  PCTIMP2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PESTIC1997WS, POPDEN2010WS, PRECIP_MINUS_EVTWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, SEPTICWS, WWTPALLDENSWS, WWTPMAJORDENSWS, Coast_dist)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCHLOR=log(Chloride)) %>% 
  relocate(logCHLOR, .before = Chloride) %>% 
  select(-c(MLocID:COMID, Chloride, SampleStartDate)) %>%  #drop Chloride in favor of logCHLOR. Drop AWQMS columns too. 
  as.data.frame() # Make sure CHLOR is a data frame, not a tibble

# Convert character columns to factors
CHLOR_CP[sapply(CHLOR_CP, is.character)] <- lapply(CHLOR_CP[sapply(CHLOR_CP, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning 
tuning <- tune(logCHLOR~., data = CHLOR_CP, seed = 123)
tuning$optimal

full_CP <- rfsrc(logCHLOR ~ ., data = CHLOR_CP, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_CP # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_CP_oob.mse <- tail(full_CP$err.rate,1)
full_CP_per.var <- 100 * (1 - full_CP_oob.mse/ var(CHLOR_CP$logCHLOR, na.rm = TRUE))
print(full_CP_per.var)

range_CHLOR <- max(exp(CHLOR_CP$logCHLOR)) - min(exp(CHLOR_CP$logCHLOR))

#In-bag RMSE
full_CP_rmse<- rmse(exp(full_CP$predicted), exp(CHLOR_CP$logCHLOR))
full_CP_rmse.scaled <- full_CP_rmse/range_CHLOR
print(paste0('In-bag RMSE = ' , round(full_CP_rmse,2), " (mg/L)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_CP_rmse.scaled*100,2), "%"))

#OOB RMSE
full_CP_oob.rmse<- rmse(exp(full_CP$predicted.oob), exp(CHLOR_CP$logCHLOR))
full_CP_oob.rmse.scaled <- full_CP_oob.rmse/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_CP_oob.rmse,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_CP_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_CP$err.rate

# VIP
plot(full_CP, plots.one.page=FALSE)
full_CP_vimp <- sort(full_CP$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_CP_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "CHLOR full_CP model VIMP")

#print(vimp(full_CP)$importance)

# Make a df for plots
full_CP_pred.oob <- data.frame(obs = CHLOR_CP$logCHLOR, pred = full_CP$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_CP_pred.ib <- data.frame(obs =CHLOR_CP$logCHLOR, pred = full_CP$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_CP_plotdf <- rbind(full_CP_pred.oob, full_CP_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_CP_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR full_CP Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_CP_oob.rmse,2), ' | scaled RMSE = ', round(full_CP_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_CP_rmse,2), ' | scaled RMSE = ', round(full_CP_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_CP_per.var,2), '%'), size = 3) 

# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the CHLOR table. Predictions will be in the same order as the CHLOR table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(CHLOR_CP$Coast_dist))  # Coast dist works.
# any(duplicated(CHLOR$PRECIP9120WS))

# Join predictions to StreamCat table
full_CP_obs.predict <- cbind(full_CP_plotdf, CHLOR_CP) %>% 
  filter(predictions == "OOB") %>% 
  arrange(Coast_dist) # Order by unique col

full_CP_awqms <- chlor_vars_selected %>% 
  arrange(Coast_dist) %>% # Order by unique col
  select(MLocID:logCHLOR) 

# Join obs.pred with awqms table, and makes sure the CHLOR results match (Make sure logCHLOR column is the same as logCHLOR.1)
full_CP_map <- cbind(full_CP_obs.predict,full_CP_awqms)

# Then select to remove duplicate logCHLOR column and reorder.
full_CP_map <- full_CP_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, full_CP_map$obs))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, full_CP_map$obs))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(full_CP_map$delta),max(full_CP_map$delta)))

map_function(full_CP_map)

CHLOR_CP_no.hi <- CHLOR_CP %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMDENSWS", "FERTWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS",  "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS",  "WWTPMAJORDENSWS", "WWTPALLDENSWS")), ~0))

# Make predictions with adjusted human-influence to 0 
full_CP_no.hi <- predict(full_CP, newdata= CHLOR_CP_no.hi, seed = 24)

# Mean squared error (untransformed)
full_CP_mse_no.hi <- tail(full_CP_no.hi$err.rate,1)
full_CP_per.var_no.hi <- 100 * (1 - full_CP_mse_no.hi/ var(CHLOR_CP_no.hi$logCHLOR, na.rm = TRUE))

# RMSE
full_CP_rmse_no.hi<- rmse(exp(full_CP_no.hi$predicted), exp(CHLOR_CP_no.hi$logCHLOR))
full_CP_rmse.scaled_no.hi <- full_CP_rmse_no.hi/range_CHLOR
print(paste0('OOB RMSE = ' , round(full_CP_rmse_no.hi,2), " (mg/L)"))
print(paste0('Scaled OOB RMSE = ' , round(full_CP_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_CP_plotdf_no.hi <- data.frame(obs = CHLOR_CP_no.hi$logCHLOR, pred = full_CP_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_CP_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "CHLOR_CP full_CP Model predictions without human influenced vars",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x =0.9, y = 100, label = paste0('OOB RMSE = ' , round(full_CP_rmse_no.hi,2), ' | scaled RMSE = ', round(full_CP_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full_CP model, adjusted human-influenced vars to 0
full_CP_resid_no.hi <- exp(CHLOR_CP_no.hi$logCHLOR) - exp(full_CP_no.hi$predicted) # Observed minus predicted oob
plot(full_CP_resid_no.hi ~ exp(full_CP_no.hi$predicted), xlab = "Predicted CHLOR_CP (mg/L)", ylab = "Residuals (obs - pred)", main = "full_CP model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)

```

