---
title: "TSS Reference Benchmarks"
author: "S. Berzins, A. Thompson"
date: "2025-10-21"
output: html_document
---

#### **PREFACE:** This script explores the creation of a site-specific reference benchmark for total suspended solids (TSS) for ESM 566 Fall Environmental Statistics final project.

# **1 - LOAD PACKAGES**

```{r warning=TRUE}
library(tidyverse)
library(readxl)
library(writexl)
library(ggplot2)
```

# **2 - IMPORT DATA**

## STATIONS/TSS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
tss <- read_excel("tss_2025-11-04.xlsx") #output file from ref benchmarks script

comids <- tss %>% select(COMID) #get just the COMIDs for next step
```

## WATERSHED PREDICTOR VARIABLES - *FROM EPA STREAMCAT*

#### Learn about StreamCat predictors here: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

#### Make table of StreamCat metadata and clean up in Excel

```{r}
#supposedly best source for query-able metric names
#partable <- StreamCatTools::sc_get_metric_names()

#name mismatch error message suggests consulting this list
#scpar <- StreamCatTools::sc_get_params(param = 'metric_names') 

#save intermediate file initially; switch to Teams to clean up, then no need to run these first steps
#write_xlsx(partable, path = "C:/Users/athomps/OneDrive - Oregon/Desktop/partable.xlsx") 

kitty <- read_excel("C:/Users/athomps/Oregon/DEQ - Biomonitoring is Fun! - General/Env't Stats Class Project 2025/StreamCat/StreamCatMetrics_new.xlsx") #final metrics table

#keepers <- subset(kitty, kitty$`Use?`=='y') #keep only parameters of interest
#mets <- keepers[,5] #make data frame of keeper metric names capable of querying data

# FOR TESTING INDIVIDUAL METRICS/COMIDS (e.g., to ensure names match)
#test <- StreamCatTools::sc_get_data(comid = '23910659', metric = 'precipminusevt', showAreaSqKm = FALSE, aoi='watershed')
```

#### Run function to query StreamCat data by subcategory

```{r}
# full data pulls use aoi='catchment,watershed,other')) 
# generating Spearman ranks use aoi='watershed')). Run separate line for aoi='other' since they don't pull unless on their own


# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture","Climate","Dams","Flow","IWI","Land Cover","Lithology","Mining_Toxics","Roads",
                                           "Urban","Wildfire","Others")){
type <- match.arg(type)
#comids_split <- split(comids, ceiling(seq_along(comids)/750)) #removed b/c doesn't seem to be doing anything - wrong data class??

if(type == "Agriculture"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslphigh2001,pctagslpmid2001,pestic1997,sw_flux,waterinput,wdrw_LD,pctcrop2001,pcthay2001,cbnf,fert,manure,nani,nsurp,agkffact',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip2008,precip8110,precip9120,tmax8110,tmax9120,tmean2008,tmean8110,tmean9120,tmin8110,tmin9120,inorgnwetdep2008,nh42008,no32008,sn2008,precipminusevt,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens,damnidstor,damnrmstor,NABD_Dens,NABD_NIDStor,NABD_NrmStor', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev,runoff', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,habt,hyd,sed,temp', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctdecid2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturblo2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'al2o3,cao,compstrgth,fe2o3,hydrlcond,k2o,mgo,na2o,p2o5,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,sio2,clay,kffact,om,perm,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'huden2010,npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,wwtpminordens,pctimp2001,pctimpslphigh2001,pctimpslpmid2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctburnarea2002,pcthighsev2002,pctincvegresp2002,pctlowsev2002,pctmodsev2002,pctnonprocmask2002,pctundsev2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfulldepth,bankfullwidth,ici,iwi,mast2008,msst2008,mwst2008,thalwagdepth,wettedwidth', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

#### Pull StreamCat data by subcategory and merge

```{r}
# PULL DATA BY SUBCATEGORY/TYPE
ag <- get_streamcat(comids = comids, type = "Agriculture")
clim <- get_streamcat(comids = comids, type = "Climate")
dams <- get_streamcat(comids = comids, type = "Dams")
flow <- get_streamcat(comids = comids, type = "Flow")
iwi <- get_streamcat(comids = comids, type = "IWI") #except prg_bmmi0809 b/c it crashes API
cover <- get_streamcat(comids = comids, type = "Land Cover")
litho <- get_streamcat(comids = comids, type = "Lithology") 
mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
roads <- get_streamcat(comids = comids, type = "Roads")
urb <- get_streamcat(comids = comids, type = "Urban")
fire <- get_streamcat(comids = comids, type = "Wildfire")
other <- get_streamcat(comids = comids, type = "Others")

# MERGE INTO ONE TABLE
meow <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
meow <- meow %>% reduce(full_join, by='COMID')

# IDENTIFY WHICH COMID(S) DIDN'T YIELD STREAMCAT RESULTS AND INVESTIGATE (FIX IN STATIONS DB BEFORE PROCEEDING)
# kcomids <- meow %>% select(COMID)
# nomatch <- comids[!comids$COMID %in% kcomids$COMID, ] #make data frame showing which COMIDs aren't pulling StreamCat data
# #nomatch.c <- subset(tss, tss$COMID == "23764745" | tss$COMID == "24516234" | tss$COMID == "23815386") #change COMIDs as needed
# rm(list = c('nomatch', 'nomatch.c', 'kcomids')) #clean up environment

# GROUP PREDICTORS BY HUMAN VS NATURAL
human <- subset(kitty, kitty$Category == "Anthropogenic" & kitty$`Use?` == 'y')
natural <- subset(kitty, kitty$Category == "Natural" & kitty$`Use?` == 'y')
human.mets <- human[,6] #just the human related metric names
natural.mets <- natural[,6] #just the natural metric names
#then use these to filter out streamcat for human vs natural
```

#### Make Spearman correlation matrix to identify covaried parameters

```{r, warning=FALSE}
# FULL CORRELATION MATRIX - ALL VARIABLES
spear <- cor(meow, method = "spearman", use = "pairwise.complete.obs")
spear <- as.data.frame(spear)
spear.rows <- tibble::rownames_to_column(spear)

#export once, then clean up in Excel
#write_xlsx(spear.rows, path = "C:/Users/athomps/Oregon/DEQ - Biomonitoring is Fun! - General/Env't Stats Class Project 2025/StreamCat/Spearmans.xlsx")

# CONDENSED CORRELATION MATRIX - ONLY VARIABLES >|0.8|
matrix <- spear.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

#export once, then clean up in Excel
#write_xlsx(matrix, path = "C:/Users/athomps/Oregon/DEQ - Biomonitoring is Fun! - General/Env't Stats Class Project 2025/StreamCat/Spearmans2_rerun.xlsx")
```

## EXPLORE CO-VARIED PARAMETERS AND SELECT WHICH WE WANT TO RETAIN FOR MODELLING PURPOSES

## PULL STREAMCAT AGAIN BUT ONLY FOR PARAMETERS OF INTEREST

```{r}
# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture2","Climate2","Dams2","Flow2","IWI2","Land Cover2","Lithology2","Mining_Toxics2","Roads2",
                                           "Urban2","Wildfire2","Others2")){
type <- match.arg(type)
#comids_split <- split(comids, ceiling(seq_along(comids)/750)) #removed b/c doesn't seem to be doing anything - wrong data class??

if(type == "Agriculture2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslpmid2001,pestic1997,wdrw_LD,pctcrop2001,sw_flux',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip9120,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,hyd',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'compstrgth,fe2o3,hydrlcond,na2o,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,clay,kffact,om,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,pctimp2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctincvegresp2002,pctnonprocmask2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others2"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfullwidth,iwi,msst2008',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

## PULL STREAMCAT AGAIN BUT ONLY FOR PARAMETERS OF INTEREST

```{r}
# PULL DATA BY SUBCATEGORY/TYPE
ag2 <- get_streamcat(comids = comids, type = "Agriculture2")
clim2 <- get_streamcat(comids = comids, type = "Climate2")
dams2 <- get_streamcat(comids = comids, type = "Dams2")
flow2 <- get_streamcat(comids = comids, type = "Flow2")
iwi2 <- get_streamcat(comids = comids, type = "IWI2") #except prg_bmmi0809 b/c it crashes API
cover2 <- get_streamcat(comids = comids, type = "Land Cover2")
litho2 <- get_streamcat(comids = comids, type = "Lithology2") 
mintox2 <- get_streamcat(comids = comids, type = "Mining_Toxics2")
roads2 <- get_streamcat(comids = comids, type = "Roads2")
urb2 <- get_streamcat(comids = comids, type = "Urban2")
fire2 <- get_streamcat(comids = comids, type = "Wildfire2")
other2 <- get_streamcat(comids = comids, type = "Others2")

# MERGE INTO ONE TABLE
meow2 <- list(ag2, clim2, dams2, flow2, iwi2, cover2, litho2, mintox2, roads2, urb2, fire2, other2)
meow2 <- meow2 %>% reduce(full_join, by='COMID')

# FINAL CHECK FOR CORRELATED VARIABLES
spear2 <- cor(meow2, method = "spearman", use = "pairwise.complete.obs")
spear2 <- as.data.frame(spear2)
spear.rows2 <- tibble::rownames_to_column(spear2)

matrix2 <- spear.rows2 %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

# REMOVE METRICS FOR UNWANTED REPRESENTATIVE COMIDS FOR SITES ON HIGH RES NHD LAYER
scmets <- subset(meow2, meow2$COMID != '23774647' & meow2$COMID != '23877125' & meow2$COMID != '23810652')

# CLEAN UP ENVIRONMENT
rm(list=c('ag2','clim2','cover2','dams2','fire2','flow2','iwi2','litho2','mintox2','other2','roads2','urb2'))
```

# **3 - BUILD MODEL**

## Random forests...

```{r}
# EXPLORE IF REMOVED COMIDS WERE OUTLIERS 
removed <- subset(meow2, meow2$COMID == '23774647' | meow2$COMID == '23877125' | meow2$COMID == '23810652')


# MESSING AROUND WITH KERNEL DENSITY PLOTS TO SUMMARIZE METRICS
colnames(scmets)
attach(scmets)

hist(ELEVWS, breaks=100)
hist(BFIWS, breaks=100)
hist(CANALDENSWS, breaks=100)
hist(CLAYWS, breaks=100)
hist(KFFACTWS, breaks=100)

density(CLAYWS)
#-----------
library(KernSmooth)

bkde(CLAYWS)
summary(CLAYWS)

clay.ws<-CLAYWS[!is.na(CLAYWS)]
summary(clay.ws)

plot(bkde(clay.ws))

plot(density(clay.ws))
rug(jitter(clay.ws))
#-----------
kern <- scmets %>%
  select(CLAYWS, BFIWS, ELEVWS, KFFACTWS, IWI, RDDENSWS)

kern %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  geom_rug() 

tss.sc <- left_join(tss, scmets, by = "COMID")
#tss.r <- tss.sc %>% select(Result_Numeric_mod)
r <- randomForest(Result_Numeric_mod ~ CLAYWS + BFIWS + ELEVWS + KFFACTWS + IWI + RDDENSWS, data = tss.sc, na.action = na.exclude)
print(r)
```


