---
title: "Conductivity Reference Benchmarks"
author: "S. Berzins, A. Thompson"
date: "2026-03-12"
output: html_document
---

# *Conductivity*

# **1 - LOAD PACKAGES**

```{r warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(RColorBrewer)
library(htmlwidgets)
library(maps)
```

# **2 - IMPORT DATA**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
cond <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-02-10.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Conductivity') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>% 
  mutate(Conductivity = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

comids <- cond %>% 
  select(COMID) # Get just the COMIDs for next step
```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

##### Learn about StreamCat predictors: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

##### View StreamCat updates: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-updates>

### OPTION 1: Pull data direction from StreamCat.

```{r streamcat pull}
# # Run function to query ALL StreamCat data for selected COMIDs by subcategory
# source('~/BioMonORDEQ/Benchmarks/Water Chemistry/SC_pull.R')
# 
# #Pull ALL StreamCat data by subcategory and merge
# ag <- get_streamcat(comids = comids, type = "Agriculture")
# clim <- get_streamcat(comids = comids, type = "Climate")
# dams <- get_streamcat(comids = comids, type = "Dams")
# flow <- get_streamcat(comids = comids, type = "Flow")
# iwi <- get_streamcat(comids = comids, type = "IWI") 
# cover <- get_streamcat(comids = comids, type = "Land Cover")
# litho <- get_streamcat(comids = comids, type = "Lithology") 
# mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
# roads <- get_streamcat(comids = comids, type = "Roads")
# urb <- get_streamcat(comids = comids, type = "Urban")
# fire <- get_streamcat(comids = comids, type = "Wildfire")
# other <- get_streamcat(comids = comids, type = "Others")
# 
# # MERGE INTO ONE TABLE
# sc_all <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
# sc_all <- sc_all %>% reduce(full_join, by = 'COMID')
# 
# rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'roads', 'urb', 'fire', 'other'))
```

### OPTION 2: Read in SC data from pre-downloaded file.

```{r streamcat flat file}
sc_all <- read_excel("~/BioMonORDEQ/Benchmarks/Water Chemistry/SC_all_comids_2026-05-14.xlsx")
sc_all <- inner_join(sc_all, comids)  # only keep Conductivity COMIDs.
```

## C. OTHER PREDICTORS - *OUTSIDE OF STREAMCAT*

#### Erodible geology, NHD slope, stream power

```{r}
# ERODIBLE VS RESISTENT GEOLOGY (KEPT ERODIBLE, SINCE RESISTENT IS INVERSE OF ERODIBLE)
erod_resist_strmpow <- sc_all %>%
  mutate(PCT_EROD = PCTALLUVCOASTWS+PCTCARBRESIDWS+PCTCOASTCRSWS+PCTCOLLUVSEDWS+PCTEOLCRSWS+PCTEOLFINEWS+PCTGLACLAKECRSWS+PCTGLACLAKEFINEWS+PCTGLACTILCLAYWS+PCTGLACTILCRSWS+PCTGLACTILLOAMWS+PCTSALLAKEWS+PCTNONCARBRESIDWS)  %>% 
  mutate(PCT_RESIST = PCTALKINTRUVOLWS+PCTEXTRUVOLWS+PCTHYDRICWS+PCTSILICICWS) 

# SLOPE FROM NHD (TAKES 1-2 MIN)
# Note: do NOT push the downloaded file to GitHub as it's too large and will result in issues. Commit the ignore instead.
erod_resist_strmpow <- ORDEQBioassessment::get_NHD_info(erod_resist_strmpow) 

# WATERSHED AREA AND STREAM POWER
erod_resist_strmpow <- erod_resist_strmpow %>% 
  mutate(STRMPOW_CAT = totdasqkm + PRECIP9120WS * slope) %>%  # WSAREASQKM from SC, instead pulling totdasqkm from get_NHD (they are the same)
  select(COMID, PCT_EROD, STRMPOW_CAT, NHD_pSLOPE, SITE_TYPE) # Drop PCT_RESIST since it's the inverse of PCT_ERODE

# JOIN WITH OTHER PARAMETERS NOT FROM STREAMCAT
cond_vars <- sc_all %>% 
  inner_join(erod_resist_strmpow, by='COMID') %>% # Join StreamCat data with erod_resist_strmpow data calulated above
  inner_join(cond, by = 'COMID') # Join those with columns from AWQMS.

# RE-ORDER COLUMNS 
cond_vars <- cond_vars %>% 
  relocate(MLocID:Conductivity, .before = 'COMID') %>% 
  relocate(L2Eco:L3Eco, .after = SITE_TYPE) %>% 
  select(-c(HUC8, EcoRegion4)) 
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(cond_vars, function(x) sum(is.na(x))))

# THEN WRITE CODE FOR THEM IN ROUGH FIX SECTION BELOW.

# Note: Some predictors may have a large number of missing values (e.g., bankfull width, n = 68). Consider performing rough fix only for variables with ~10 or fewer missing values. This will avoid introducing error in the correlation matrix and variable selection phase.
```

#### PERFORM NA ROUGH FIX

```{r}
# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH ONLY A FEW MISSING VALUES (Leave metrics with >10 NAs [MAST2008, BMMI0809, WETTEDWITDTH, THALWEGDEPTH, BANKFULLWIDTH, BANKFULLDEPTH] until after the correlation table is built.)
cond_vars$CLAYWS  <- na.roughfix(cond_vars$CLAYWS ) # 1 NA
cond_vars$MSST2008 <- na.roughfix(cond_vars$ MSST2008) # 1 NA
cond_vars$MWST2008 <- na.roughfix(cond_vars$ MWST2008) # 11 NAs
cond_vars$NANIWS <- na.roughfix(cond_vars$NANIWS) # 1 NA
cond_vars$NSURPWS <- na.roughfix(cond_vars$NSURPWS) # 1 NA
cond_vars$OMWS <- na.roughfix(cond_vars$OMWS) # 1 NA
cond_vars$PCTIMPSLPHIGH2001WS <- na.roughfix(cond_vars$PCTIMPSLPHIGH2001WS) # 4 NAs
cond_vars$PERMWS <- na.roughfix(cond_vars$PERMWS) # 1 NA
cond_vars$RCKDEPWS <- na.roughfix(cond_vars$RCKDEPWS) # 1 NA
cond_vars$SANDWS <- na.roughfix(cond_vars$SANDWS) # 1 NA
cond_vars$SW_FLUXWS <- na.roughfix(cond_vars$SW_FLUXWS) # 2 NA
cond_vars$WTDEPWS <- na.roughfix(cond_vars$WTDEPWS) # 1 NA
```

## B. PREDICTOR VARIABLE SELECTION

#### Make Spearman correlation matrix to identify co-varied predictors (>0.8)

```{r, warning=FALSE}
# Subset numeric variables only
cond_vars_num <- cond_vars %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Conductivity)) %>%  
  relocate(sort(names(.))) # sort columns alphabetically

# FULL CORRELATION MATRIX
spear <- cor(cond_vars_num, method = "spearman", use = "pairwise.complete.obs")
spear <- as.data.frame(spear)
spear.rows <- tibble::rownames_to_column(spear)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix <- spear.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear', 'spear.rows'))
```

#### Export matrix and clean up in Excel

```{r}
# Export once, then clean up in Excel.
#write_xlsx(matrix, path = paste0("~/BioMonORDEQ/Benchmarks/Water Chemistry/Conductivity/Cond_Spearmans_Matrix", Sys.Date(), ".xlsx"))
```

#### Calculate coefficient of variance (CV), proportion of 0s, and kernel density plots to guide variable selection decisions

```{r}
# Calculate coefficient of variance (CV) for each predictor.
cv <- cond_vars_num %>% 
  summarise(across(everything(), ~ cv(.x, na.rm = TRUE))) %>% 
  pivot_longer(1:140, names_to="Var", values_to="CV") # Change first argument to 1:# columns

# Calculate % of zeros for each predictor.
cv_pctzero <- cond_vars_num %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE) / n() * 100)) %>% 
  pivot_longer(1:140, names_to="Var", values_to="Percent Zeros") %>%  # Change  first argument to 1:# columns
  inner_join(cv, by = "Var")

# Write output to Excel for variable selection meeting.
#write_xlsx(cv_pctzero, path = "~/BioMonORDEQ/Benchmarks/Water Chemistry/Conductivity/Cond_CofV.xlsx")
rm(cv)

# Make kernel density plots for predictors of interest to compare distributions.
cond_vars %>%
    select("AGKFFACTWS","AL2O3WS","BANKFULLDEPTH","BANKFULLWIDTH","BFIWS","CAOWS","CBNFWS","CLAYWS","CONNWS","DAMDENSWS","DAMNIDSTORWS","DAMNRMSTORWS") %>%  # Edit this line to include vars that you want to look at.
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  geom_rug()
```

#### Hold variable selection meeting

```{r}
# 1. Make a variable selection decisions sheet in Excel based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection decisions.xlsx) and save to model folder on GitHub.

# 2. Clean up correlation matrix Excel table manually:
  # a) Filter data and delete rows where cor = 1.
  # b) Sort row names alphabetically.
  # c) Remove composite indexes (CHEM, CONN, HABT, HYD, ICI, IWI, SED, and TEMP).
  # d) Make a new matrix predictor decisions sheet based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection correlations) and save to model folder on GitHub. 
  #     Copy row names into template (col A) and update formulas, column names into col G, and correlations into col L. 
  #     Add horizontal borders between groupings. 
  #     You may want to hide columns B and H (they are just for the formulas)

# 3. For each co-varied predictor(s), discuss ecological considerations and data distribution (CV, how well covers gradient, %0s).

# 4. Record which predictors to keep/remove and document rationale in the decisions sheet.

# 5. Remove predictors that have 100% zeros (no modeling power) and paste list of keepers below.
```

#### Update predictor list

```{r}
# Use Excel formula (=TEXTJOIN(", ", TRUE, range:range)) to compile list from decisions sheet and paste below.
cond_vars_selected <- cond_vars %>% 
  select(c(MLocID:COMID, BFIWS, CANALDENSWS, CAOWS, CLAYWS, COMPSTRGTHWS, DAMDENSWS, ELEVWS, FE2O3WS, HYDRLCONDWS, K2OWS, KFFACTWS, MINEDENSWS, MSST2008, NA2OWS, NHD_pSLOPE, NPDESDENSWS, NWS, OMWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTALKINTRUVOLWS, PCTALLUVCOASTWS, PCTBL2001WS, PCTBURNAREA2002WS, PCTCOLLUVSEDWS, PCTCONIF2001WS, PCTCROP2001WS, PCTDECID2001WS, PCTEOLCRSWS, PCTEOLFINEWS, PCTEXTRUVOLWS, PCTFRSTLOSS2002WS, PCTGLACLAKECRSWS, PCTGLACLAKEFINEWS, PCTGLACTILCRSWS, PCTGRS2001WS, PCTHBWET2001WS, PCTICE2001WS, PCTIMP2001WS, PCTINCVEGRESP2002WS, PCTNONAGINTRODMANAGVEGWS, PCTOW2001WS, PCTSALLAKEWS, PCTSHRB2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PCTWATERWS, PESTIC1997WS, POPDEN2010WS, PRECIP9120WS, RCKDEPWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, ROCKNWS, SANDWS, SEPTICWS, STRMPOW_CAT, SUPERFUNDDENSWS, SW_FLUXWS, SWS, TMAX9120WS, TRIDENSWS, WATERINPUTWS, WETINDEXWS, WETTEDWIDTH, WTDEPWS, WWTPALLDENSWS, SITE_TYPE:L3Eco)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCOND = log(Conductivity)) %>% 
  relocate(logCOND, .before= Conductivity)
```

#### Perform final correlation check to ensure no co-varied predictors remain

```{r, warning=FALSE}
# Subset numeric variables only
cond_vars_num_selected <- cond_vars_selected %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Conductivity)) 

# FULL CORRELATION MATRIX
spear2 <- cor(cond_vars_num_selected, method = "spearman", use = "pairwise.complete.obs")
spear2 <- as.data.frame(spear2)
spear2.rows <- tibble::rownames_to_column(spear2)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix2 <- spear2.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear2', 'spear2.rows', 'tp_vars_num_selected'))

# ***OPEN NEWLY CREATED MATRIX AND CONTINUE ONLY IF NO CORRELATED VARIABLES REMAIN (ALL SHOULD BE 1)***
```

#### Clear out the environment

```{r}
rm(list=c('cv_pctzero', 'erod_resist_strmpow', 'matrix', 'matrix2', 'comids',  'cond_vars_num','cond_vars_num_selected'))
```

#### Perform final NA check and perform rough fix as desired

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(cond_vars_selected, function(x) sum(is.na(x))))

# DO ANY LAST NA ROUGH FIXES AS DESIRED (write code for specific predictors) 
cond_vars_selected$WETTEDWIDTH <- na.roughfix(cond_vars$WETTEDWIDTH) # 68 NA
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r}
# HISTOGRAM
hist(cond_vars_selected$Conductivity, main = "", xlab = "Conductivity") #raw data
hist(cond_vars_selected$logCOND, main = "", xlab = "log(Conductivity)" )

# BOX PLOT & STRIP CHART
boxplot(cond_vars_selected$Conductivity, ylab = "Conductivity")
boxplot(cond_vars_selected$logCOND, ylab = "logCond")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r}
# Select response data of interest
COND <- cond_vars_selected %>%
  select(-c(MLocID:COMID, Conductivity, SampleStartDate)) #drop Conductivity in favor of logCOND. Drop AWQMS columns too.

# Convert character columns to factors
COND[sapply(COND, is.character)] <- lapply(COND[sapply(COND, is.character)], as.factor)

# Order factors
COND$L2Eco <- factor(COND$L2Eco, levels = c("Marine West Coast Forest", "Western Cordillera", "Cold Deserts"))
COND$L3Eco <- factor(COND$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
#Adjust the parameters below based on tuning (nodesize = 2; mtry = 73)
tuning <- tune(logCOND~., COND, seed = 1234)
tuning$optimal
```

#### Build full model

```{r}
full <- rfsrc(logCOND ~ ., data = COND, mtry = 73, nodesize = 2, importance = "TRUE", seed = 24)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(full_per.var)

range_COND <- max(exp(COND$logCOND)) - min(exp(COND$logCOND))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(COND$logCOND))
full_rmse.scaled <- full_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(COND$logCOND))
full_oob.rmse.scaled <- full_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = COND$logCOND, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =COND$logCOND, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Conductivity Full Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 50, y = 320, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 50, y = 390, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 50, y = 460, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```

#### Leaflet map of full model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, COND) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

full_awqms <- cond_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logCOND) 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
full_map <- cbind(full_obs.predict,full_awqms)

# Then select to remove duplicate logCOND column and reorder.
full_map <- full_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Observed Conductivity: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$pred), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Predicted Conductivity: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Obs - Pred Conductivity: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$obs, position = "bottomright" , title = "Observed Conductivity" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = data$obs, position = "bottomright" , title = "Predicted Conductivity" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred Conductivity" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map$obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map$delta), max(full_map$delta)))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//full_model_map.html", selfcontained = TRUE, title = "Full model")
```

#### Partial dependence plots for full model

```{r}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c("PRECIP9120WS", "KFFACTWS", "TMAX9120WS", "OMWS", "PCTBL2001WS", "SWS", "ELEVWS", "TRIDENSWS"), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c("PCT_EROD", "WTDEPWS", "PCTDECID2001WS", "PCTCROP2001WS", "SANDWS", "MSST2008", "PCTOW2001WS", "CLAYWS"), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("TRIDENSWS", "PCTCROP2001WS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Another way to do partial plots (slower but more customizable)
```{r}
partial.obj <- partial(full, partial.xvar = "PRECIP9120WS", partial.values = full$xvar$PRECIP9120WS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "PRECIP9120WS", ylab = "ŷ")

partial.obj <- partial(full, partial.xvar = "KFFACTWS", partial.values = full$xvar$KFFACTWS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "KFFACTWS", ylab = "ŷ")

partial.obj <- partial(full, partial.xvar = "TMAX9120WS", partial.values = full$xvar$TMAX9120WS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "TMAX9120WS", ylab = "ŷ")
```

#### Leaflet map of predictors

```{r}
map_predictors <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "PRECIP9120WS", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$PRECIP9120WS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "PRECIP9120WS: ", "</strong>", data$PRECIP9120WS))  %>% 
      addCircleMarkers(data = data, group = "KFFACTWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$KFFACTWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "KFFACTWS: ", "</strong>", data$KFFACTWS))  %>% 
     addCircleMarkers(data = data, group = "TMAX9120WS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$TMAX9120WS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "TMAX9120WS: ", "</strong>", data$TMAX9120WS))  %>% 
     addCircleMarkers(data = data, group = "OMWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$OMWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "OMWS: ", "</strong>", data$OMWS))  %>% 
    hideGroup(c("KFFACTWS","TMAX9120WS", "OMWS")) %>%  # Don't check all layers by default
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("PRECIP9120WS", "KFFACTWS","TMAX9120WS", "OMWS"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "PRECIP9120WS", values = data$PRECIP9120WS, position = "bottomright" , title = "PRECIP9120WS" ) %>% 
    addLegend(pal = pal2,group = "KFFACTWS",values = data$KFFACTWS, position = "bottomright" , title = "KFFACTWS" ) %>%  
    addLegend(pal = pal3, group = "TMAX9120WS", values = data$TMAX9120WS, position = "bottomleft" , title = "TMAX9120WS" ) %>% 
    addLegend(pal = pal4, group = "OMWS", values = data$OMWS, position = "bottomleft" , title = "OMWS" ) 
}

pal <- colorNumeric(palette = "Blues", 
  domain = c(0,max(full_map$PRECIP9120WS)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map$KFFACTWS)))
pal3 <- colorNumeric(palette = "RdBu", 
                     domain = c(min(full_map$TMAX9120WS),max(full_map$TMAX9120WS)), reverse = TRUE)
pal4 <- colorNumeric(palette = "Oranges", 
  domain = c(0, max(full_map$OMWS)))

map_predictors(full_map)

# Save map as html
# predictors_map <- map_predictors(full_map)
# saveWidget(predictors_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//predictors_map.html", selfcontained = TRUE, title = "Predictors map")
```

#### Residual plots

```{r}
# Plot residuals 
full_resid <- exp(COND$logCOND) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = "Predicted Conductivity (μS/cm)", ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(COND$logCOND) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r}
# Reduce all human-influenced vars to all 0s.
COND_no.hi <- COND %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMDENSWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WATERINPUTWS", "WWTPALLDENSWS")), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(COND_no.hi$logCOND))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model predictions without human influenced vars",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 50, y = 460, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = "Predicted COND (μS/cm)", ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map adjusting human vbls to 0

```{r}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to cal_val table
full_obs.predict_no.hi <- cbind(full_plotdf_no.hi, COND_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
full_map_no.hi <- cbind(full_awqms,full_obs.predict_no.hi)

# Then select to remove duplicate logCOND column and reorder.
full_map_no.hi <- full_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(full_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map_no.hi$delta), max(full_map_no.hi$delta)))

map_function(full_map_no.hi)

# Save map as html
# full_model_map_no_hi <- map_function(full_map_no.hi)
# saveWidget(full_model_map_no_hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//full_model_map_no_hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

## C. Model Reduction

```{r}

# Since adjusting human-influenced variables down to zero doesn't change the predictions appreciably, reducing the model doesn't seem like a good use of time.

```


## D. Reference-Only Model

#### Subset the data

```{r}
COND.ref <- cond_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, Conductivity, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMDENSWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTFRSTLOSS2002WS", "PCTIMP2001WS",  "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WWTPALLDENSWS")) # Remove anthropogenic StreamCat vars

# Convert character columns to factors
COND.ref[sapply(COND.ref, is.character)] <- lapply(COND.ref[sapply(COND.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
#Adjust the parameters below based on tuning (nodesize = 2; mtry = 32)
tuning <- tune(logCOND~., COND.ref, seed = 2)
tuning$optimal

```

#### Build full ref-only model

```{r}
ref <- rfsrc(logCOND ~ ., data = COND.ref, mtry = 32, nodesize = 2, importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(COND.ref$logCOND, na.rm = TRUE))
print(ref_per.var)

ref_range_COND <- max(exp(COND.ref$logCOND)) - min(exp(COND.ref$logCOND))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(COND.ref$logCOND))
ref_rmse.scaled <- ref.rmse/ref_range_COND
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(COND.ref$logCOND))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_COND
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = COND.ref$logCOND, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =COND.ref$logCOND, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Reference-Only Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 35, y = 110, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 35, y = 120, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 35, y = 130, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(COND.ref$logCOND) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = "Predicted COND (μS/cm)", ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Partial dependence plots for full ref-only model

```{r}
# Top importance vars
partialplots <- plot.variable(ref, xvar.names = c("CLAYWS", "PCTBL2001WS", "TMAX9120WS", "MSST2008", "ELEVWS", "PCTICE2001WS", "SANDWS", "ROCKNWS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of Ref-Only model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, COND.ref) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

ref_awqms <- cond_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  filter(ReferenceSite=='REFERENCE') %>% 
  select(MLocID:logCOND) 

# Join obs.predict and awqms tables, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
ref_map <- cbind(ref_obs.predict, ref_awqms)

# Then select to remove duplicate logCOND column and reorder.
ref_map <- ref_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "BuGn", 
  domain = c(0,max(ref_map$obs)))
pal2 <- colorNumeric(palette = "BuGn", 
  domain = c(0,max(ref_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(ref_map$delta), max(ref_map$delta)))

map_function(ref_map)

# Save map as html
# ref_model_map <- map_function(ref_map)
# saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## E. Make comparisons of predictions

#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(MLocID = full_map$MLocID, StationDes = full_map$StationDes, Obs = full_map$obs, Pred_unadj = full_map$pred, L3Eco = as.character(full_map$L3Eco),Pred_adj = full_map_no.hi$pred, Ref = full_map$ReferenceSite)

boxplot_ref<- ref_map %>%
  select(MLocID, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="MLocID")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_unadj, Pred_adj, Pred_ref, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_unadj", "Obs"))
boxplot_df$L3Eco <- factor(boxplot_df$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  theme_bw(base_size=14) + ylab("COND (μS/cm)") +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5
  )

# Ignore warning for 698 rows removed - these are NAs in the Pred_ref column.
```

#### Combined Leaflet map of predictions

```{r}
leaflet_df <- left_join(boxplot_df, cond %>%  select(MLocID, Project1, Lat_DD, Long_DD), by = "MLocID") %>% 
  pivot_wider(names_from = Model, values_from = Prediction) %>% 
  relocate(Obs, .before= Pred_unadj)

map_all <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$Obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Observed Conductivity: ", "</strong>", data$Obs))  %>% 
      addCircleMarkers(data = data, group = "Pred_unadj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$Pred_unadj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted Conductivity UNADJUSTED (full model): ", "</strong>", data$Pred_unadj))  %>% 
     addCircleMarkers(data = data, group = "Pred_adj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$Pred_adj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted Conductivity ADJUSTED: ", "</strong>", data$Pred_adj))  %>% 
     addCircleMarkers(data = data, group = "Pred_ref",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$Pred_ref), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted Conductivity REF ONLY: ", "</strong>", data$Pred_ref))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Pred_unadj", "Pred_adj", "Pred_ref"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$Obs, position = "bottomright" , title = "Observed Conductivity" ) %>% 
    addLegend(pal = pal2,group = "Pred_unadj",values = data$Obs, position = "bottomright" , title = "Predicted Conductivity (UNADJUSTED)" ) %>%  # Keep values=data$Obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Pred_adj", values = data$Pred_adj, position = "bottomleft" , title = "Predicted Conductivity (ADJUSTED)" ) %>% 
        addLegend(pal = pal4, group = "Pred_ref", values = data$Pred_adj, position = "bottomleft" , title = "Predicted Conductivity (REF ONLY)" )    # Keep values=data$Pred_adj here (instead of data$Ref) to keep the legend colors scales the same between ref expectations.
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal3 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_unadj)))
pal4 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_unadj)), na.color = "transparent") 

map_all(leaflet_df)

# Save map as html
# map_all <- map_all(leaflet_df)
# saveWidget(map_all, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//map_all.html", selfcontained = TRUE, title = "Map_All")
```
 
#### Unadjusted predictions vs. adjusted human inflence 

```{r}
leaflet_df$Ref <- factor(leaflet_df$Ref, levels = c("REFERENCE", "MODERATELY DISTURBED", "MOST DISTURBED"))

ggplot(leaflet_df, aes(x = Pred_unadj, y = Pred_adj, color = Ref, shape = Ref)) +
  geom_point(size = 1.5) + theme_bw(base_size = 15) +
  labs(
    x = "Reduced model predicted COND (μS/cm)",
    y = "Predicted COND (μS/cm) adjusted\n for no human disturbance predictions",
    color = "Reference Status", shape = "Reference Status") + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```
