---
title: "Conductivity Reference Benchmarks"
author: "S. Berzins, A. Thompson"
date: "2026-03-12"
output: html_document
---

# *Conductivity*

# **1 - LOAD PACKAGES**

```{r warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(RColorBrewer)
library(htmlwidgets)
```

# **2 - IMPORT DATA**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
cond <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-02-10.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Conductivity') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>% 
  mutate(Conductivity = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

comids <- cond %>% 
  select(COMID) # Get just the COMIDs for next step
```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

##### Learn about StreamCat predictors: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

##### View StreamCat updates: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-updates>

#### Run function to query ALL StreamCat data for selected COMIDs by subcategory

```{r}
# NOTE: StreamCat doesn't pull data for 'other' areas of interest when aoi='watershed, catchment, other'. 
# Instead, use aoi='watershed' and run separate line for aoi='other' since they don't pull unless on their own.
# ALSO: Do not separate metric lists onto a new line by hitting the enter button, as any metric after the break will be excluded from the pull.

# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture","Climate","Dams","Flow","IWI","Land Cover","Lithology","Mining_Toxics","Roads",
                                           "Urban","Wildfire","Others")){
type <- match.arg(type)

if(type == "Agriculture"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslphigh2001,pctagslpmid2001,pestic1997,sw_flux,waterinput,wdrw_LD,pctcrop2001,pcthay2001,cbnf,fert,manure,nani,nsurp,agkffact',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip2008,precip8110,precip9120,tmax8110,tmax9120,tmean2008,tmean8110,tmean9120,tmin8110,tmin9120,inorgnwetdep2008,nh42008,no32008,sn2008,precip_minus_evt,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens,damnidstor,damnrmstor,NABD_Dens,NABD_NIDStor,NABD_NrmStor', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev,runoff', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,habt,hyd,sed,temp', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctdecid2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturblo2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'al2o3,cao,compstrgth,fe2o3,hydrlcond,k2o,mgo,na2o,p2o5,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,sio2,clay,kffact,om,perm,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'huden2010,npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,wwtpminordens,pctimp2001,pctimpslphigh2001,pctimpslpmid2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctburnarea2002,pcthighsev2002,pctincvegresp2002,pctlowsev2002,pctmodsev2002,pctnonprocmask2002,pctundsev2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfulldepth,bankfullwidth,ici,iwi,mast2008,msst2008,mwst2008,thalwegdepth,wettedwidth,prg_bmmi0809', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

#### Pull ALL StreamCat data by subcategory and merge

```{r}
# PULL DATA BY SUBCATEGORY/TYPE
ag <- get_streamcat(comids = comids, type = "Agriculture")
clim <- get_streamcat(comids = comids, type = "Climate")
dams <- get_streamcat(comids = comids, type = "Dams")
flow <- get_streamcat(comids = comids, type = "Flow")
iwi <- get_streamcat(comids = comids, type = "IWI") 
cover <- get_streamcat(comids = comids, type = "Land Cover")
litho <- get_streamcat(comids = comids, type = "Lithology") 
mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
roads <- get_streamcat(comids = comids, type = "Roads")
urb <- get_streamcat(comids = comids, type = "Urban")
fire <- get_streamcat(comids = comids, type = "Wildfire")
other <- get_streamcat(comids = comids, type = "Others")

# MERGE INTO ONE TABLE
sc_all <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
sc_all <- sc_all %>% reduce(full_join, by = 'COMID')

rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'roads', 'urb', 'fire', 'other'))
```

## C. OTHER PREDICTORS - *OUTSIDE OF STREAMCAT*

#### Erodible geology, NHD slope, stream power

```{r}
# ERODIBLE VS RESISTENT GEOLOGY (KEPT ERODIBLE, SINCE RESISTENT IS INVERSE OF ERODIBLE)
erod_resist_strmpow <- sc_all %>%
  mutate(PCT_EROD = PCTALLUVCOASTWS+PCTCARBRESIDWS+PCTCOASTCRSWS+PCTCOLLUVSEDWS+PCTEOLCRSWS+PCTEOLFINEWS+PCTGLACLAKECRSWS+PCTGLACLAKEFINEWS+PCTGLACTILCLAYWS+PCTGLACTILCRSWS+PCTGLACTILLOAMWS+PCTSALLAKEWS+PCTNONCARBRESIDWS)  %>% 
  mutate(PCT_RESIST = PCTALKINTRUVOLWS+PCTEXTRUVOLWS+PCTHYDRICWS+PCTSILICICWS) 

# SLOPE FROM NHD (TAKES 1-2 MIN)
# Note: do NOT push the downloaded file to GitHub as it's too large and will result in issues. Commit the ignore instead.
erod_resist_strmpow <- ORDEQBioassessment::get_NHD_info(erod_resist_strmpow) 

# WATERSHED AREA AND STREAM POWER
erod_resist_strmpow <- erod_resist_strmpow %>% 
  mutate(STRMPOW_CAT = totdasqkm + PRECIP9120WS * slope) %>%  # WSAREASQKM from SC, instead pulling totdasqkm from get_NHD (they are the same)
  select(COMID, PCT_EROD, STRMPOW_CAT, NHD_pSLOPE, SITE_TYPE) # Drop PCT_RESIST since it's the inverse of PCT_ERODE

# JOIN WITH OTHER PARAMETERS NOT FROM STREAMCAT
cond_vars <- sc_all %>% 
  inner_join(erod_resist_strmpow, by='COMID') %>% # Join StreamCat data with erod_resist_strmpow data calulated above
  inner_join(cond, by = 'COMID') # Join those with columns from AWQMS.

# RE-ORDER COLUMNS 
cond_vars <- cond_vars %>% 
  relocate(MLocID:Conductivity, .before = 'COMID') %>% 
  relocate(L2Eco:L3Eco, .after = SITE_TYPE) %>% 
  select(-c(HUC8, EcoRegion4)) 
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(cond_vars, function(x) sum(is.na(x))))

# THEN WRITE CODE FOR THEM IN ROUGH FIX SECTION BELOW.

# Note: Some predictors may have a large number of missing values (e.g., bankfull width, n = 68). Consider performing rough fix only for variables with ~10 or fewer missing values. This will avoid introducing error in the correlation matrix and variable selection phase.
```

#### PERFORM NA ROUGH FIX

```{r}
# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH ONLY A FEW MISSING VALUES (Leave metrics with >10 NAs [MAST2008, BMMI0809, WETTEDWITDTH, THALWEGDEPTH, BANKFULLWIDTH, BANKFULLDEPTH] until after the correlation table is built.)
cond_vars$CLAYWS  <- na.roughfix(cond_vars$CLAYWS ) # 1 NA
cond_vars$MSST2008 <- na.roughfix(cond_vars$ MSST2008) # 1 NA
cond_vars$MWST2008 <- na.roughfix(cond_vars$ MWST2008) # 11 NAs
cond_vars$NANIWS <- na.roughfix(cond_vars$NANIWS) # 1 NA
cond_vars$NSURPWS <- na.roughfix(cond_vars$NSURPWS) # 1 NA
cond_vars$OMWS <- na.roughfix(cond_vars$OMWS) # 1 NA
cond_vars$PCTIMPSLPHIGH2001WS <- na.roughfix(cond_vars$PCTIMPSLPHIGH2001WS) # 4 NAs
cond_vars$PERMWS <- na.roughfix(cond_vars$PERMWS) # 1 NA
cond_vars$RCKDEPWS <- na.roughfix(cond_vars$RCKDEPWS) # 1 NA
cond_vars$SANDWS <- na.roughfix(cond_vars$SANDWS) # 1 NA
cond_vars$SW_FLUXWS <- na.roughfix(cond_vars$SW_FLUXWS) # 2 NA
cond_vars$WTDEPWS <- na.roughfix(cond_vars$WTDEPWS) # 1 NA
```

## B. PREDICTOR VARIABLE SELECTION

#### Make Spearman correlation matrix to identify co-varied predictors (\>0.8)

```{r, warning=FALSE}
# Subset numeric variables only
cond_vars_num <- cond_vars %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Conductivity)) %>%  
  relocate(sort(names(.))) # sort columns alphabetically

# FULL CORRELATION MATRIX
spear <- cor(cond_vars_num, method = "spearman", use = "pairwise.complete.obs")
spear <- as.data.frame(spear)
spear.rows <- tibble::rownames_to_column(spear)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix <- spear.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear', 'spear.rows'))
```

#### Export matrix and clean up in Excel

```{r}
# Export once, then clean up in Excel.
#write_xlsx(matrix, path = paste0("~/BioMonORDEQ/Benchmarks/Water Chemistry/Conductivity/Cond_Spearmans_Matrix", Sys.Date(), ".xlsx"))
```

#### Calculate coefficient of variance (CV), proportion of 0s, and kernel density plots to guide variable selection decisions

```{r}
# Calculate coefficient of variance (CV) for each predictor.
cv <- cond_vars_num %>% 
  summarise(across(everything(), ~ cv(.x, na.rm = TRUE))) %>% 
  pivot_longer(1:140, names_to="Var", values_to="CV") # Change first argument to 1:# columns

# Calculate % of zeros for each predictor.
cv_pctzero <- cond_vars_num %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE) / n() * 100)) %>% 
  pivot_longer(1:140, names_to="Var", values_to="Percent Zeros") %>%  # Change  first argument to 1:# columns
  inner_join(cv, by = "Var")

# Write output to Excel for variable selection meeting.
#write_xlsx(cv_pctzero, path = "~/BioMonORDEQ/Benchmarks/Water Chemistry/Conductivity/Cond_CofV.xlsx")
rm(cv)

# Make kernel density plots for predictors of interest to compare distributions.
cond_vars %>%
    select("AGKFFACTWS","AL2O3WS","BANKFULLDEPTH","BANKFULLWIDTH","BFIWS","CAOWS","CBNFWS","CLAYWS","CONNWS","DAMDENSWS","DAMNIDSTORWS","DAMNRMSTORWS") %>%  # Edit this line to include vars that you want to look at.
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  geom_rug()
```

#### Hold variable selection meeting

```{r}
# 1. Make a variable selection decisions sheet in Excel based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection decisions.xlsx) and save to model folder on GitHub.

# 2. Clean up correlation matrix Excel table manually:
  # a) Filter data and delete rows where cor = 1.
  # b) Sort row names alphabetically.
  # c) Remove composite indexes (CHEM, CONN, HABT, HYD, ICI, IWI, SED, and TEMP).
  # d) Make a new matrix predictor decisions sheet based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection correlations) and save to model folder on GitHub. 
  #     Copy row names into template (col A) and update formulas, column names into col G, and correlations into col L. 
  #     Add horizontal borders between groupings. 
  #     You may want to hide columns B and H (they are just for the formulas)

# 3. For each co-varied predictor(s), discuss ecological considerations and data distribution (CV, how well covers gradient, %0s).

# 4. Record which predictors to keep/remove and document rationale in the decisions sheet.

# 5. Remove predictors that have 100% zeros (no modeling power) and paste list of keepers below.
```

#### Update predictor list

```{r}
# Use Excel formula (=TEXTJOIN(", ", TRUE, range:range)) to compile list from decisions sheet and paste below.
cond_vars_selected <- cond_vars %>% 
  select(c(MLocID:COMID, BFIWS, CANALDENSWS, CAOWS, CLAYWS, COMPSTRGTHWS, DAMDENSWS, ELEVWS, FE2O3WS, HYDRLCONDWS, K2OWS, KFFACTWS, MINEDENSWS, MSST2008, NA2OWS, NHD_pSLOPE, NPDESDENSWS, NWS, OMWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTALKINTRUVOLWS, PCTALLUVCOASTWS, PCTBL2001WS, PCTBURNAREA2002WS, PCTCOLLUVSEDWS, PCTCONIF2001WS, PCTCROP2001WS, PCTDECID2001WS, PCTEOLCRSWS, PCTEOLFINEWS, PCTEXTRUVOLWS, PCTFRSTLOSS2002WS, PCTGLACLAKECRSWS, PCTGLACLAKEFINEWS, PCTGLACTILCRSWS, PCTGRS2001WS, PCTHBWET2001WS, PCTICE2001WS, PCTIMP2001WS, PCTINCVEGRESP2002WS, PCTNONAGINTRODMANAGVEGWS, PCTOW2001WS, PCTSALLAKEWS, PCTSHRB2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PCTWATERWS, PESTIC1997WS, POPDEN2010WS, PRECIP9120WS, RCKDEPWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, ROCKNWS, SANDWS, SEPTICWS, STRMPOW_CAT, SUPERFUNDDENSWS, SW_FLUXWS, SWS, TMAX9120WS, TRIDENSWS, WATERINPUTWS, WETINDEXWS, WETTEDWIDTH, WTDEPWS, WWTPALLDENSWS, SITE_TYPE:L3Eco)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCOND = log(Conductivity)) %>% 
  relocate(logCOND, .before= Conductivity)
```

#### Perform final correlation check to ensure no co-varied predictors remain

```{r, warning=FALSE}
# Subset numeric variables only
cond_vars_num_selected <- cond_vars_selected %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Conductivity)) 

# FULL CORRELATION MATRIX
spear2 <- cor(cond_vars_num_selected, method = "spearman", use = "pairwise.complete.obs")
spear2 <- as.data.frame(spear2)
spear2.rows <- tibble::rownames_to_column(spear2)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix2 <- spear2.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear2', 'spear2.rows', 'tp_vars_num_selected'))

# ***OPEN NEWLY CREATED MATRIX AND CONTINUE ONLY IF NO CORRELATED VARIABLES REMAIN (ALL SHOULD BE 1)***
```

#### Clear out the environment

```{r}
rm(list=c('cv_pctzero', 'erod_resist_strmpow', 'matrix', 'matrix2', 'comids',  'cond_vars_num','cond_vars_num_selected'))
```

#### Perform final NA check and perform rough fix as desired

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(cond_vars_selected, function(x) sum(is.na(x))))

# DO ANY LAST NA ROUGH FIXES AS DESIRED (write code for specific predictors) 
cond_vars_selected$WETTEDWIDTH <- na.roughfix(cond_vars$WETTEDWIDTH) # 68 NA
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r}
# HISTOGRAM
hist(cond_vars_selected$Conductivity, main = "", xlab = "Conductivity") #raw data
hist(cond_vars_selected$logCOND, main = "", xlab = "log(Conductivity)" )

# BOX PLOT & STRIP CHART
boxplot(cond_vars_selected$Conductivity, ylab = "Conductivity")
boxplot(cond_vars_selected$logCOND, ylab = "logCond")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r}
# Select response data of interest
COND <- cond_vars_selected %>%
  select(-c(MLocID:COMID, Conductivity, SampleStartDate)) #drop Conductivity in favor of logCOND. Drop AWQMS columns too.

# Convert character columns to factors
COND[sapply(COND, is.character)] <- lapply(COND[sapply(COND, is.character)], as.factor)
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
#Adjust the parameters below based on tuning (nodesize = 2; mtry = 73)
tuning <- tune(logCOND~., COND, seed = 1234)
tuning$optimal
```

#### Build full model

```{r}
full <- rfsrc(logCOND ~ ., data = COND, mtry = 73, nodesize = 2, importance = "TRUE", seed = 24)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(full_per.var)

range_COND <- max(exp(COND$logCOND)) - min(exp(COND$logCOND))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(COND$logCOND))
full_rmse.scaled <- full_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(COND$logCOND))
full_oob.rmse.scaled <- full_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = COND$logCOND, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =COND$logCOND, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Conductivity Full Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 50, y = 320, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 50, y = 390, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 50, y = 460, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```

#### Leaflet map of full model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, COND) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

full_awqms <- cond_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logCOND) 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
full_map <- cbind(full_obs.predict,full_awqms)

# Then select to remove duplicate logCOND column and reorder.
full_map <- full_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Observed Conductivity: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$pred), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Predicted Conductivity: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Obs - Pred Conductivity: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$obs, position = "bottomright" , title = "Observed Conductivity" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = data$obs, position = "bottomright" , title = "Predicted Conductivity" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred Conductivity" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, 800))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, 800))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-630,630))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//full_model_map.html", selfcontained = TRUE, title = "Full model")
```

#### Partial dependence plots for full model

```{r}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c("PRECIP9120WS", "KFFACTWS", "TMAX9120WS", "OMWS", "PCTBL2001WS", "SWS", "ELEVWS", "TRIDENSWS"), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c("PCT_EROD", "WTDEPWS", "PCTDECID2001WS", "PCTCROP2001WS", "SANDWS", "MSST2008", "PCTOW2001WS", "CLAYWS"), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("TRIDENSWS", "PCTCROP2001WS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Another way to do partial plots (slower but more customizable)
```{r}
partial.obj <- partial(full, partial.xvar = "PRECIP9120WS", partial.values = full$xvar$PRECIP9120WS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "PRECIP9120WS", ylab = "ŷ")

partial.obj <- partial(full, partial.xvar = "KFFACTWS", partial.values = full$xvar$KFFACTWS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "KFFACTWS", ylab = "ŷ")

partial.obj <- partial(full, partial.xvar = "TMAX9120WS", partial.values = full$xvar$TMAX9120WS)
pdta <- get.partial.plot.data(partial.obj)
plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "TMAX9120WS", ylab = "ŷ")
```

#### Leaflet map of predictors  - not currently working

```{r}
map_predictors <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "PRECIP9120WS", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$PRECIP9120WSWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "PRECIP9120WSWS: ", "</strong>", data$PRECIP9120WSWS))  %>% 
      addCircleMarkers(data = data, group = "KFFACTWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$KFFACTWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "KFFACTWS: ", "</strong>", data$KFFACTWS))  %>% 
     addCircleMarkers(data = data, group = "TMAX9120WS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$TMAX9120WS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "TMAX9120WS: ", "</strong>", data$TMAX9120WS))  %>% 
     addCircleMarkers(data = data, group = "WTDEPWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$WTDEPWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "WTDEPWS: ", "</strong>", data$WTDEPWS))  %>% 
     addCircleMarkers(data = data, group = "OMWS",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal5(data$OMWS), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "OMWS: ", "</strong>", data$OMWS))  %>% 
    hideGroup(c("KFFACTWS","TMAX9120WS", "WTDEPWS", "OMWS")) %>%  # Don't check all layers by default
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("PRECIP9120WSWS", "KFFACTWS","TMAX9120WS", "WTDEPWS", "OMWS"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "PRECIP9120WSWS", values = data$PRECIP9120WSWS, position = "bottomright" , title = "PRECIP9120WSWS" ) %>% 
    addLegend(pal = pal2,group = "KFFACTWS",values = data$KFFACTWS, position = "bottomright" , title = "KFFACTWS" ) %>%  
    addLegend(pal = pal3, group = "TMAX9120WS", values = data$TMAX9120WS, position = "bottomleft" , title = "TMAX9120WS" ) %>% 
    addLegend(pal = pal4, group = "WTDEPWS", values = data$WTDEPWS, position = "bottomleft" , title = "WTDEPWS" ) %>% 
    addLegend(pal = pal5, group = "OMWS", values = data$OMWS, position = "bottomleft" , title = "OMWS")
}

pal <- colorNumeric(palette = "Blues", 
  domain = c(270,4700))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, 0.45))
pal3 <- colorNumeric(palette = "RdYlBu", 
  domain = c(6,20))
pal4 <- colorNumeric(palette = "Greys", 
  domain = c(35, 185))
pal5 <- colorNumeric(palette = "Oranges", 
  domain = c(0.8,6.8))

map_predictors(full_map)

# Save map as html
# predictors_map <- map_predictors(full_map)
# saveWidget(predictors_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//predictors_map.html", selfcontained = TRUE, title = "Predictors map")
```

#### Residual plots

```{r}
# Plot residuals 
full_resid <- exp(COND$logCOND) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = "Predicted Conductivity (μS/cm)", ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(COND$logCOND) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r}
# Reduce all human-influenced vars to all 0s.
COND_no.hi <- COND %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMDENSWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WATERINPUTWS", "WWTPALLDENSWS")), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(COND_no.hi$logCOND))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model predictions without human influenced vars",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 50, y = 460, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = "Predicted COND (μS/cm)", ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map adjusting human vbls to 0

```{r}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to cal_val table
full_obs.predict_no.hi <- cbind(full_plotdf_no.hi, COND_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
full_map_no.hi <- cbind(full_awqms,full_obs.predict_no.hi)

# Then select to remove duplicate logCOND column and reorder.
full_map_no.hi <- full_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, 800))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, 800))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-630,630))

map_function(full_map_no.hi)

# Save map as html
# full_model_map_no_hi <- map_function(full_map_no.hi)
# saveWidget(full_model_map_no_hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//COND//full_model_map_no_hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

#### How do residuals compare between full model predictions and adjusted model?

```{r}
par(mar=c(5,6,4,1)+.1)
plot(full_map_no.hi$pred, (full_map$obs - full_map$pred) - (full_map_no.hi$obs - full_map_no.hi$pred) , xlab = "Full model COND predictions with no HI", ylab = "Full model resid minus \n full model with no human influenced vars resid")
abline(h=0)

```

## C. Model Reduction

#### Build model with Top 12 predictors

```{r}
red12 <- rfsrc(logTP ~ PRECIP_MINUS_EVTWS + PCTCONIF2001WS + PESTIC1997WS + WTDEPWS + ROCKNWS + RCKDEPWS + RDDENSWS + PCTBL2001WS + KFFACTWS + SEPTICWS + WETINDEXWS +OMWS, data = TP, mtry = 74, nodesize = 1, importance = "TRUE", seed = 24)
red12 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red12_oob.mse <- tail(red12$err.rate,1)
red12_per.var <- 100 * (1 - red12_oob.mse/ var(TP$logTP, na.rm = TRUE))
print(red12_per.var)

#In-bag RMSE
red12_rmse<- rmse(exp(red12$predicted), exp(TP$logTP))
red12_rmse.scaled <- red12_rmse/range_TP
print(paste0('In-bag RMSE = ' , round(red12_rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(red12_rmse.scaled*100,2), "%"))

#OOB RMSE
red12_oob.rmse<- rmse(exp(red12$predicted.oob), exp(TP$logTP))
red12_oob.rmse.scaled <- red12_oob.rmse/range_TP
print(paste0('OOB RMSE = ' , round(red12_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(red12_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red12$err.rate

# VIP
plot(red12, plots.one.page=FALSE)
red12_vimp <- sort(red12$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red12_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TP Reduced (12 variables) model VIMP")

#print(vimp(red12)$importance)

# Make a df for plots
red12_pred.oob <- data.frame(obs = TP$logTP, pred = red12$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red12_pred.ib <- data.frame(obs =TP$logTP, pred = red12$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red12_plotdf <- rbind(red12_pred.oob, red12_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red12_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Conductivity Red10 Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.015, y = 0.16, label = paste0('OOB RMSE = ' , round(red12_oob.rmse,2), ' | scaled RMSE = ', round(red12_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.015, y = 0.2, label = paste0('IB RMSE = ' , round(red12_rmse,2), ' | scaled RMSE = ', round(red12_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.015, y = 0.25, label = paste0('% variance explained = ' , round(red12_per.var,2), '%'), size = 3) 
```

#### Leaflet map of Red12 model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TP table. Predictions will be in the same order as the TP table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(TP$ELEVWS))  # Elevation works.
# any(duplicated(TP$PRECIP9120WS))

# Join predictions to StreamCat table
red12_obs.predict <- cbind(red12_plotdf, TP) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

red12_awqms <- tp_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logTP) 

# Join obs.pred with awqms table, and makes sure the TP results match (Make sure logTP column is the same as logTP.1)
red12_map <- cbind(red12_obs.predict,red12_awqms)

# Then select to remove duplicate logTP column and reorder.
red12_map <- red12_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, 0.7))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, 0.7))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-0.5,0.5))

map_function(red12_map)

# Save map as html
# red12_model_map <- map_function(red12_map)
# saveWidget(red12_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TP//red12_model_map.html", selfcontained = TRUE, title = "Red10 model")
```

#### Distributions for Human-influenced vars at all reference sites

```{r}
# Select only the human influenced predictors at reference sites from the reduced model.
TP_ref_hi <- full_map %>% 
  filter(ReferenceSite == "REFERENCE") %>% 
  select(c(MLocID:logTP, RDDENSWS, PESTIC1997WS, SEPTICWS))

# RDDENSWS: Mean Road Density (%)
par(mfrow=c(1,2))
plot.variable(red12, xvar.names = c("RDDENSWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP_ref_hi$RDDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP_ref_hi$RDDENSWS, main = '')
quantile(TP_ref_hi$RDDENSWS, probs = seq(.1, .9, by = .1))

# PESTIC1997WS: Mean pesticide use (kg/km2) in yr. 1997
par(mfrow=c(1,2))
plot.variable(red12, xvar.names = c("PESTIC1997WS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP_ref_hi$PESTIC1997WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP_ref_hi$PESTIC1997WS, main = '')
quantile(TP_ref_hi$PESTIC1997WS, probs = seq(.1, .9, by = .1))


# SEPTICWS: Density of septic systems(Kilogram/Hectare/Year)
par(mfrow=c(1,2))
plot.variable(red12, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP_ref_hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP_ref_hi$SEPTICWS, main = '')
quantile(TP_ref_hi$SEPTICWS, probs = seq(.1, .9, by = .1))
```

#### Reduce all human-influenced variables [to 0]* and make predictions using reduced model.
##### * We may decide to adjust to a non-zero percentile based on reference quantiles in the chunk above.

```{r}
# Make predictions with adjusted human-influence to 0 
red12_no.hi <- predict(red12, newdata= TP_no.hi, seed = 24)

# Mean squared error (untransformed)
red12_mse_no.hi <- tail(red12_no.hi$err.rate,1)
red12_per.var_no.hi <- 100 * (1 - red12_mse_no.hi/ var(TP_no.hi$logTP, na.rm = TRUE))

# RMSE
red12_rmse_no.hi<- rmse(exp(red12_no.hi$predicted), exp(TP_no.hi$logTP))
red12_rmse.scaled_no.hi <- red12_rmse_no.hi/range_TP
print(paste0('OOB RMSE = ' , round(red12_rmse_no.hi,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(red12_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red12_plotdf_no.hi <- data.frame(obs = TP_no.hi$logTP, pred = red12_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red12_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TP Red10 Model predictions without human influenced vars",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.01, y = 0.1, label = paste0('OOB RMSE = ' , round(red12_rmse_no.hi,2), ' | scaled RMSE = ', round(red12_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red12 model, adjusted human-influenced vars to 0
red12_resid_no.hi <- exp(TP_no.hi$logTP) - exp(red12_no.hi$predicted) # Observed minus predicted oob
plot(red12_resid_no.hi ~ exp(red12_no.hi$predicted), xlab = "Predicted TP (μS/cm)", ylab = "Residuals (obs - pred)", main = "Red10 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map after adjusting human vbls for reduced10 model

```{r}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TP table. Predictions will be in the same order as the TP table initially.

# Join predictions to cal_val table
red12_obs.predict_no.hi <- cbind(red12_plotdf_no.hi, TP_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join TP.obs_predict_no.hi with TP.awqms, and makes sure the results match.
red12_map_no.hi <- cbind(full_awqms,red12_obs.predict_no.hi)

# Then select to remove duplicate logTP column and reorder.
red12_map_no.hi <- red12_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-0.7,0.7))

map_function(red12_map_no.hi)

# Save map as html
# red12_model_map_no.hi <- map_function(red12_map_no.hi)
# saveWidget(red12_model_map_no.hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TP//red12_model_map_no.hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

#### How do residuals compare between reduced predictions and reduced adjusted model?

```{r}
par(mar=c(5,6,4,1)+.1)
plot(red12_map_no.hi$pred, (red12_map$obs - red12_map$pred) - (red12_map_no.hi$obs - red12_map_no.hi$pred) , xlab = "TP Red10 model predictions with no HI", ylab = "Red10 model resid minus \n red12 model with no human influenced vars resid")
abline(h=0)
```

#### Top 3 predictors (not using this one)

```{r}
red3 <- rfsrc(logTP ~ PRECIP_MINUS_EVTWS + PCTCONIF2001WS + PESTIC1997WS, data = TP, mtry = 57, nodesize = 7, importance = "TRUE", seed = 24)
red3 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red3_oob.mse <- tail(red3$err.rate,1)
red3_per.var <- 100 * (1 - red3_oob.mse/ var(TP$logTP, na.rm = TRUE))
print(red3_per.var)

#In-bag RMSE
red3_rmse<- rmse(exp(red3$predicted), exp(TP$logTP))
red3_rmse.scaled <- red3_rmse/range_TP
print(paste0('In-bag RMSE = ' , round(red3_rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(red3_rmse.scaled*100,2), "%"))

#OOB RMSE
red3_oob.rmse<- rmse(exp(red3$predicted.oob), exp(TP$logTP))
red3_oob.rmse.scaled <- red3_oob.rmse/range_TP
print(paste0('OOB RMSE = ' , round(red3_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(red3_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red3$err.rate

# VIP
plot(red3, plots.one.page=FALSE)
red3_vimp <- sort(red3$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red3_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TP Reduced (3 variables) model VIMP")

#print(vimp(red3)$importance)

# Make a df for plots
red3_pred.oob <- data.frame(obs = TP$logTP, pred = red3$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red3_pred.ib <- data.frame(obs =TP$logTP, pred = red3$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red3_plotdf <- rbind(red3_pred.oob, red3_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red3_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Conductivity Red3 Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.015, y = 0.16, label = paste0('OOB RMSE = ' , round(red3_oob.rmse,2), ' | scaled RMSE = ', round(red3_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.015, y = 0.2, label = paste0('IB RMSE = ' , round(red3_rmse,2), ' | scaled RMSE = ', round(red3_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.015, y = 0.25, label = paste0('% variance explained = ' , round(red3_per.var,2), '%'), size = 3) 
```

## E. Reference-Only Model

#### Subset the data

```{r}
TP.ref <- tp_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, Total_phosphorus, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMNRMSTORWS", "FERTWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTCROP2001WS", "PCTHAY2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")) # Remove anthropogenic StreamCat vars

# Convert character columns to factors
TP.ref[sapply(TP.ref, is.character)] <- lapply(TP.ref[sapply(TP.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
#Adjust the parameters below based on tuning (nodesize = 3; mtry = 20)
tuning <- tune(logTP~., TP.ref, seed = 2)
tuning$optimal

```

#### Build full ref-only model

```{r}
ref <- rfsrc(logTP ~ ., data = TP.ref, mtry = 20, nodesize = 3, importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(TP.ref$logTP, na.rm = TRUE))
print(ref_per.var)

ref_range_TP <- max(exp(TP.ref$logTP)) - min(exp(TP.ref$logTP))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(TP.ref$logTP))
ref_rmse.scaled <- ref.rmse/ref_range_TP
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(TP.ref$logTP))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_TP
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TP Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = TP.ref$logTP, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =TP.ref$logTP, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TP Full Reference-Only Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.015, y = 0.16, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.015, y = 0.2, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.015, y = 0.25, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(TP.ref$logTP) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = "Predicted TP (μS/cm)", ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map of Ref-Only model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TP table. Predictions will be in the same order as the TP table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(TP$ELEVWS))  # Elevation works.
# any(duplicated(TP$PRECIP9120WS))

# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, TP.ref) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

ref_awqms <- tp_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  filter(ReferenceSite=='REFERENCE') %>% 
  select(MLocID:logTP) 

# Join obs.predict and awqms tables, and makes sure the TP results match (Make sure logTP column is the same as logTP.1)
ref_map <- cbind(ref_obs.predict, ref_awqms)

# Then select to remove duplicate logTP column and reorder.
ref_map <- ref_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "BuGn", 
  domain = c(0, 0.11))
pal2 <- colorNumeric(palette = "BuGn", 
  domain = c(0, 0.11))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-0.085, 0.085))

map_function(ref_map)

# Save map as html
# ref_model_map <- map_function(ref_map)
# saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TP//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## F. Make comparisons of predictions
#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(MLocID = red12_map$MLocID, StationDes = red12_map$StationDes, Obs = red12_map$obs, Pred_unadj = red12_map$pred, L3Eco = as.character(red12_map$L3Eco),Pred_adj = red12_map_no.hi$pred, Ref = red12_map$ReferenceSite)

boxplot_ref<- ref_map %>%
  select(MLocID, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="MLocID")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_unadj, Pred_adj, Pred_ref, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_unadj", "Obs"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  theme_bw() + ylab("TP (μS/cm)")

# Ignore warning for 624 rows removed - these are NAs in the Pred_ref column. (112 reference sites of 729 total observations.)
```
