---
title: "StreamCat Data Puller"
author: "Biomon"
date: "2026-03-06"
output: html_document
---

# 1 - *LOAD PACKAGES*

```{r}
library(tidyverse)
library(StreamCatTools) 
library(AWQMSdata)
library(readxl)
library(writexl)
```

# 2 - *READ INPUT STATIONS* (choose Option A or B)

#### Option A: Excel

```{r}
# ENTER PATH OF EXCEL INPUT FILE
import <- read_excel("C:/Users/athomps/OneDrive - Oregon/Desktop/New folder/mergedCOMIDs_fixed.xlsx")
```

#### Option B: Stations Database

```{r}
# QUERY STATIONS OF INTEREST FROM STATIONS DATABASE
import <- query_stations(mlocs = c("ORBES0016"))
```

# 3 - *ISOLATE COMIDS FOR DATA PULL*

```{r}
comids <- import %>% 
  select(COMID)
```

# 4 - *PULL EPA STREAMCAT WATERSHED DATA*

##### Learn about StreamCat predictors: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

##### View StreamCat updates: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-updates>

#### Run function to query ALL StreamCat data for selected COMIDs by subcategory

```{r}
# NOTE: StreamCat doesn't pull data for 'other' areas of interest when aoi='watershed, catchment, other'. 
# Instead, use aoi='watershed' and run separate line for aoi='other' since they don't pull unless on their own.
# ALSO: Do not separate metric lists onto a new line by hitting the enter button, as any metric after the break will be excluded from the pull.

# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture","Climate","Dams","Flow","IWI","Land Cover","Lithology","Mining_Toxics","Roads",
                                           "Urban","Wildfire","Others")){
type <- match.arg(type)

if(type == "Agriculture"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslphigh2001,pctagslpmid2001,pestic1997,sw_flux,waterinput,wdrw_LD,pctcrop2001,pcthay2001,cbnf,fert,manure,nani,nsurp,agkffact',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip2008,precip8110,precip9120,tmax8110,tmax9120,tmean2008,tmean8110,tmean9120,tmin8110,tmin9120,inorgnwetdep2008,nh42008,no32008,sn2008,precip_minus_evt,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens,damnidstor,damnrmstor,NABD_Dens,NABD_NIDStor,NABD_NrmStor', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev,runoff', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,habt,hyd,sed,temp', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctdecid2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturblo2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'al2o3,cao,compstrgth,fe2o3,hydrlcond,k2o,mgo,na2o,p2o5,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,sio2,clay,kffact,om,perm,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'huden2010,npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,wwtpminordens,pctimp2001,pctimpslphigh2001,pctimpslpmid2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctburnarea2002,pcthighsev2002,pctincvegresp2002,pctlowsev2002,pctmodsev2002,pctnonprocmask2002,pctundsev2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfulldepth,bankfullwidth,ici,iwi,mast2008,msst2008,mwst2008,thalwegdepth,wettedwidth,prg_bmmi0809', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

#### Pull ALL StreamCat data by subcategory and merge

```{r}
# PULL DATA BY SUBCATEGORY/TYPE
ag <- get_streamcat(comids = comids, type = "Agriculture")
clim <- get_streamcat(comids = comids, type = "Climate")
dams <- get_streamcat(comids = comids, type = "Dams")
flow <- get_streamcat(comids = comids, type = "Flow")
iwi <- get_streamcat(comids = comids, type = "IWI") 
cover <- get_streamcat(comids = comids, type = "Land Cover")
litho <- get_streamcat(comids = comids, type = "Lithology") 
mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
roads <- get_streamcat(comids = comids, type = "Roads")
urb <- get_streamcat(comids = comids, type = "Urban")
fire <- get_streamcat(comids = comids, type = "Wildfire")
other <- get_streamcat(comids = comids, type = "Others")

# MERGE INTO ONE TABLE
sc_data <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
sc_data <- sc_data %>% reduce(full_join, by = 'COMID')

rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'roads', 'urb', 'fire', 'other'))
```

# 5 - *EXPORT DATA* (optional)

```{r}
# ENTER PATH OF EXPORT FILE
write_xlsx(sc_data, "C:/Users/athomps/OneDrive - Oregon/Desktop/New folder/StreamCatDataPull.xlsx")
```
