---
title: "Total Phosphorus Reference Benchmarks"
author: "S. Berzins, A. Thompson"
date: "2026-01-21"
output: html_document
---

# *Total Phosphorus*

# **1 - LOAD PACKAGES**

```{r warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(RColorBrewer)
```

# **2 - IMPORT DATA**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
# Get both TP and TN as we're using the same variable selection for both chars.

tp <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-02-10.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Total Phosphorus, mixed forms') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>%
  mutate(Total_phosphorus = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

comids <- tp %>% 
  select(COMID) # Get just the COMIDs for next step

tn <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-02-10.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Nitrogen') 

comids <- bind_rows(tp, tn) %>% select(COMID) %>% distinct() # Get just the COMIDs for the next step (TN and TP combined)

```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

##### Learn about StreamCat predictors: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

##### View StreamCat updates: <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-updates>

#### Run function to query ALL StreamCat data for selected COMIDs by subcategory

```{r}
# NOTE: StreamCat doesn't pull data for 'other' areas of interest when aoi='watershed, catchment, other'. 
# Instead, use aoi='watershed' and run separate line for aoi='other' since they don't pull unless on their own.
# ALSO: Do not separate metric lists onto a new line by hitting the enter button, as any metric after the break will be excluded from the pull.

# FUNCTION - START +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
get_streamcat <- function(comids, type = c("Agriculture","Climate","Dams","Flow","IWI","Land Cover","Lithology","Mining_Toxics","Roads",
                                           "Urban","Wildfire","Others")){
type <- match.arg(type)

if(type == "Agriculture"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'canaldens,pctagdrainage,pctagslphigh2001,pctagslpmid2001,pestic1997,sw_flux,waterinput,wdrw_LD,pctcrop2001,pcthay2001,cbnf,fert,manure,nani,nsurp,agkffact',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed')) 
  }

if(type == "Climate"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'precip2008,precip8110,precip9120,tmax8110,tmax9120,tmean2008,tmean8110,tmean9120,tmin8110,tmin9120,inorgnwetdep2008,nh42008,no32008,sn2008,precip_minus_evt,wetindex', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Dams"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'damdens,damnidstor,damnrmstor,NABD_Dens,NABD_NIDStor,NABD_NrmStor', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Flow"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bfi,elev,runoff', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "IWI"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'chem,conn,habt,hyd,sed,temp', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Land Cover"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctbl2001,pctconif2001,pctdecid2001,pctgrs2001,pcthbwet2001,pctice2001,pctmxfst2001,pctow2001,pctshrb2001,pcturbhi2001,pcturblo2001,pcturbmd2001,pcturbop2001,pctwdwet2001,pctfrstloss2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Lithology"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'al2o3,cao,compstrgth,fe2o3,hydrlcond,k2o,mgo,na2o,p2o5,pctalkintruvol,pctalluvcoast,pctcarbresid,pctcoastcrs,pctcolluvsed,pcteolcrs,pcteolfine,pctextruvol,pctglaclakecrs,pctglaclakefine,pctglactilclay,pctglactilcrs,pctglactilloam,pcthydric,pctnoncarbresid,pctsallake,pctsilicic,pctwater,rockn,s,sio2,clay,kffact,om,perm,Rckdep,rckdep,sand,wtdep,n',
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Mining_Toxics"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'coalminedens,minedens,superfunddens,tridens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Roads"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'rdcrs,rdcrsslpwtd,rddens', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Urban"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'huden2010,npdesdens,popden2010,septic,wwtpalldens,wwtpmajordens,wwtpminordens,pctimp2001,pctimpslphigh2001,pctimpslpmid2001,pctnonagintrodmanagveg', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Wildfire"){ 

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'pctfire2002,pctburnarea2002,pcthighsev2002,pctincvegresp2002,pctlowsev2002,pctmodsev2002,pctnonprocmask2002,pctundsev2002', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='watershed'))
  }

if(type == "Others"){

  streamcat <- purrr::map_dfr(comids, ~StreamCatTools::sc_get_data(.,
                                                                           metric = 'bankfulldepth,bankfullwidth,ici,iwi,mast2008,msst2008,mwst2008,thalwegdepth,wettedwidth,prg_bmmi0809', 
                                                                           showAreaSqKm = FALSE,
                                                                           aoi='other')) 
  }

names(streamcat) <- base::toupper(names(streamcat))
return(streamcat)
}
# FUNCTION - END  +++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
```

#### Pull ALL StreamCat data by subcategory and merge

```{r}
# PULL DATA BY SUBCATEGORY/TYPE
ag <- get_streamcat(comids = comids, type = "Agriculture")
clim <- get_streamcat(comids = comids, type = "Climate")
dams <- get_streamcat(comids = comids, type = "Dams")
flow <- get_streamcat(comids = comids, type = "Flow")
iwi <- get_streamcat(comids = comids, type = "IWI") 
cover <- get_streamcat(comids = comids, type = "Land Cover")
litho <- get_streamcat(comids = comids, type = "Lithology") 
mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
roads <- get_streamcat(comids = comids, type = "Roads")
urb <- get_streamcat(comids = comids, type = "Urban")
fire <- get_streamcat(comids = comids, type = "Wildfire")
other <- get_streamcat(comids = comids, type = "Others")

# MERGE INTO ONE TABLE
sc_all <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
sc_all <- sc_all %>% reduce(full_join, by = 'COMID')

rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'roads', 'urb', 'fire', 'other'))
```

## C. OTHER PREDICTORS - *OUTSIDE OF STREAMCAT*

#### Erodible geology, NHD slope, stream power

```{r}
# ERODIBLE VS RESISTENT GEOLOGY (KEPT ERODIBLE, SINCE RESISTENT IS INVERSE OF ERODIBLE)
erod_resist_strmpow <- sc_all %>%
  mutate(PCT_EROD = PCTALLUVCOASTWS+PCTCARBRESIDWS+PCTCOASTCRSWS+PCTCOLLUVSEDWS+PCTEOLCRSWS+PCTEOLFINEWS+PCTGLACLAKECRSWS+PCTGLACLAKEFINEWS+PCTGLACTILCLAYWS+PCTGLACTILCRSWS+PCTGLACTILLOAMWS+PCTSALLAKEWS+PCTNONCARBRESIDWS)  %>% 
  mutate(PCT_RESIST = PCTALKINTRUVOLWS+PCTEXTRUVOLWS+PCTHYDRICWS+PCTSILICICWS) 

# SLOPE FROM NHD (TAKES 1-2 MIN)
# Note: do NOT push the downloaded file to GitHub as it's too large and will result in issues. Commit the ignore instead.
erod_resist_strmpow <- ORDEQBioassessment::get_NHD_info(erod_resist_strmpow) 

# WATERSHED AREA AND STREAM POWER
erod_resist_strmpow <- erod_resist_strmpow %>% 
  mutate(STRMPOW_CAT = totdasqkm + PRECIP9120WS * slope) %>%  # WSAREASQKM from SC, instead pulling totdasqkm from get_NHD (they are the same)
  select(COMID, PCT_EROD, STRMPOW_CAT, NHD_pSLOPE, SITE_TYPE) # Drop PCT_RESIST since it's the inverse of PCT_ERODE

# JOIN WITH OTHER PARAMETERS NOT FROM STREAMCAT
tp_tn_vars <- sc_all %>% 
  inner_join(erod_resist_strmpow, by='COMID') %>% 
  inner_join(tp, by = 'COMID')

# RE-ORDER COLUMNS 
tp_tn_vars <- tp_tn_vars %>% 
  relocate(MLocID:Total_phosphorus, .before = 'COMID') %>% 
  relocate(L2Eco:L3Eco, .after = SITE_TYPE) %>% 
  select(-c(HUC8, EcoRegion4)) 
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(tp_tn_vars, function(x) sum(is.na(x))))

# THEN WRITE CODE FOR THEM IN ROUGH FIX SECTION BELOW.

# Note: Some predictors may have a large number of missing values (e.g., bankfull width, n = 64). Consider performing rough fix only for variables with ~10 or fewer missing values. This will avoid introducing error in the correlation matrix and variable selection phase.
```

#### PERFORM NA ROUGH FIX

```{r}
# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH ONLY A FEW MISSING VALUES (Leave metrics with >10 NAs [MAST2008, BMMI0809, WETTEDWITDTH, THALWEGDEPTH, BANKFULLWIDTH, BANKFULLDEPTH] until after the correlation table is built.)
tp_tn_vars$NSURPWS  <- na.roughfix(tp_tn_vars$NSURPWS) # 1 NA
tp_tn_vars$NANIWS  <- na.roughfix(tp_tn_vars$NANIWS) # 1 NA
tp_tn_vars$SW_FLUXWS  <- na.roughfix(tp_tn_vars$SW_FLUXWS) # 2 NA
tp_tn_vars$PCTIMPSLPHIGH2001WS  <- na.roughfix(tp_tn_vars$PCTIMPSLPHIGH2001WS) # 4 NA
tp_tn_vars$MWST2008  <- na.roughfix(tp_tn_vars$MWST2008) # 9 NA

```

## B. PREDICTOR VARIABLE SELECTION

#### Make Spearman correlation matrix to identify co-varied predictors (\>0.8)

```{r, warning=FALSE}
# Subset numeric variables only
tp_tn_vars_num <- tp_tn_vars %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Total_phosphorus)) %>%  
  relocate(sort(names(.))) # sort columns alphabetically

# FULL CORRELATION MATRIX
spear <- cor(tp_tn_vars_num, method = "spearman", use = "pairwise.complete.obs")
spear <- as.data.frame(spear)
spear.rows <- tibble::rownames_to_column(spear)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix <- spear.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear', 'spear.rows'))
```

#### Export matrix and clean up in Excel

```{r}
# Export once, then clean up in Excel.
#write_xlsx(matrix, path = paste0("~/BioMonORDEQ/Benchmarks/Water Chemistry/Total Phosphorus/TP_TN_Spearmans_Matrix", Sys.Date(), ".xlsx"))
```

#### Calculate coefficient of variance (CV), proportion of 0s, and kernel density plots to guide variable selection decisions

```{r}
# Calculate coefficient of variance (CV) for each predictor.
cv <- tp_tn_vars_num %>% 
  summarise(across(everything(), ~ cv(.x, na.rm = TRUE))) %>% 
  pivot_longer(1:140, names_to="Var", values_to="CV") # Change first argument to 1:# columns

# Calculate % of zeros for each predictor.
cv_pctzero <- tp_tn_vars_num %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE) / n() * 100)) %>% 
  pivot_longer(1:140, names_to="Var", values_to="Percent Zeros") %>%  # Change  first argument to 1:# columns
  inner_join(cv, by = "Var")

# Write output to Excel for variable selection meeting.
#write_xlsx(cv_pctzero, path = "~/BioMonORDEQ/Benchmarks/Water Chemistry/Total Phosphorus/TP_TN_CofV.xlsx")
rm(cv)

# Make kernel density plots for predictors of interest to compare distributions.
tp_tn_vars %>%
    select("AGKFFACTWS","AL2O3WS","BANKFULLDEPTH","BANKFULLWIDTH","BFIWS","CAOWS","CBNFWS","CLAYWS","CONNWS","DAMDENSWS","DAMNIDSTORWS","DAMNRMSTORWS") %>%  # Edit this line to include vars that you want to look at.
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  geom_rug()
```

#### Hold variable selection meeting

```{r}
# 1. Make a variable selection decisions sheet in Excel based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection decisions.xlsx) and save to model folder on GitHub.

# 2. Clean up correlation matrix Excel table manually:
  # a) Filter data and delete rows where cor = 1.
  # b) Sort row names alphabetically.
  # c) Remove composite indexes (CHEM, CONN, HABT, HYD, ICI, IWI, SED, and TEMP).
  # d) Make a new matrix predictor decisions sheet based on the TEMPLATE (Benchmarks >> Water Chemistry >> TEMPLATE_Var selection correlations) and save to model folder on GitHub. 
  #     Copy row names into template (col A) and update formulas, column names into col G, and correlations into col L. 
  #     Add horizontal borders between groupings. 
  #     You may want to hide columns B and H (they are just for the formulas)

# 3. For each co-varied predictor(s), discuss ecological considerations and data distribution (CV, how well covers gradient, %0s).

# 4. Record which predictors to keep/remove and document rationale in the decisions sheet.

# 5. Remove predictors that have 100% zeros (no modeling power) and paste list of keepers below.
```

#### Update predictor list

```{r}
# UNIQUE FOR THIS SCRIPT - Select only COMIDs with TP samples (drop TN only rows).
tp_vars <- semi_join(tp_tn_vars, tp, by = "COMID") 

# Use Excel formula (=TEXTJOIN(", ", TRUE, range:range)) to compile list from decisions sheet and paste below.
tp_vars_selected <- tp_vars %>% 
  select(c(MLocID:COMID, BFIWS, CANALDENSWS, CLAYWS, COMPSTRGTHWS, DAMNRMSTORWS, ELEVWS, FE2O3WS, FERTWS, HYDRLCONDWS, K2OWS, KFFACTWS, MINEDENSWS, MSST2008, NANIWS, NHD_pSLOPE, NPDESDENSWS, NWS, OMWS, P2O5WS, PCT_EROD, PCTAGDRAINAGEWS, PCTALKINTRUVOLWS, PCTALLUVCOASTWS, PCTBL2001WS, PCTBURNAREA2002WS, PCTCARBRESIDWS, PCTCOLLUVSEDWS, PCTCONIF2001WS, PCTCROP2001WS, PCTDECID2001WS, PCTEOLFINEWS, PCTEXTRUVOLWS, PCTFRSTLOSS2002WS, PCTGLACLAKECRSWS, PCTGLACLAKEFINEWS, PCTGLACTILCRSWS, PCTGRS2001WS, PCTHAY2001WS, PCTHBWET2001WS, PCTICE2001WS, PCTIMP2001WS, PCTINCVEGRESP2002WS, PCTNONAGINTRODMANAGVEGWS, PCTOW2001WS, PCTSALLAKEWS, PCTSHRB2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PCTWATERWS, PCTWDWET2001WS, PESTIC1997WS, POPDEN2010WS, PRECIP_MINUS_EVTWS, RCKDEPWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, ROCKNWS, SANDWS, SEPTICWS, STRMPOW_CAT, SUPERFUNDDENSWS, SW_FLUXWS, SWS, TMAX9120WS, TRIDENSWS, WETINDEXWS, WTDEPWS, WWTPALLDENSWS, WWTPMAJORDENSWS, SITE_TYPE:L3Eco)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logTP = log(Total_phosphorus)) %>% 
  relocate(logTP, .before= Total_phosphorus)
```

#### Perform final correlation check to ensure no co-varied predictors remain

```{r, warning=FALSE}
# Subset numeric variables only
tp_tn_vars_num_selected <- tp_vars_selected %>% 
  select(!c(SITE_TYPE:L3Eco)) %>% 
  select(!c(MLocID:Total_phosphorus)) 

# FULL CORRELATION MATRIX
spear2 <- cor(tp_tn_vars_num_selected, method = "spearman", use = "pairwise.complete.obs")
spear2 <- as.data.frame(spear2)
spear2.rows <- tibble::rownames_to_column(spear2)

# CONDENSED CORRELATION MATRIX (>0.8 ONLY)
matrix2 <- spear2.rows %>%  
  gather(-rowname, key = "colname", value = "cor") %>% 
  filter(abs(cor) > 0.8)

rm(list=c('spear2', 'spear2.rows', 'tp_vars_num_selected'))

# ***OPEN NEWLY CREATED MATRIX AND CONTINUE ONLY IF NO CORRELATED VARIABLES REMAIN (ALL SHOULD BE 1)***
```

#### Clear out the environment

```{r}
rm(list=c('cv_pctzero', 'erod_resist_strmpow', 'matrix', 'matrix2', 'comids', 'tp_tn_vars_num'))
```

#### Perform final NA check and perform rough fix as desired

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sapply(tp_vars_selected, function(x) sum(is.na(x)))

# DO ANY LAST NA ROUGH FIXES AS DESIRED (write code for specific predictors) (none for TP)
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r}
# HISTOGRAM
hist(tp_vars_selected$Total_phosphorus, main = "", xlab = "Total Phosphorus (mg/L)") #raw data
hist(tp_vars_selected$logTP, main = "", xlab = "log(TP)", breaks = 15)

# BOX PLOTS
boxplot(tp_vars_selected$Total_phosphorus, ylab = "TP")
boxplot(tp_vars_selected$logTP, ylab = "log(TP)")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r}
# Select response data of interst
TP <- tp_vars_selected %>%
  select(-c(MLocID:COMID, Total_phosphorus, SampleStartDate)) #drop tp in favor of logTP. Drop AWQMS columns too.

# Convert character columns to factors
TP[sapply(TP, is.character)] <- lapply(TP[sapply(TP, is.character)], as.factor)
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
set.seed(2) # need to set seed to make tuning reproducible.

#Adjust the parameters below based on tuning (nodesize = 2; mtry = 74)
tuning <- tune(logTP~., TP)
tuning$optimal

```

#### Build full model

```{r}
set.seed(24)

rf.full <- rfsrc(logTP ~ ., data = TP, mtry = 74, nodesize = 2, importance = "TRUE")
rf.full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
oob_mse <- tail(rf.full$err.rate,1)
per_var <- 100 * (1 - oob_mse/ var(TP$logTP, na.rm = TRUE))
print(per_var)

range_tp <- max(exp(TP$logTP)) - min(exp(TP$logTP))

#In-bag RMSE
rmse<- rmse(exp(rf.full$predicted), exp(TP$logTP))
rmse_scaled <- rmse/range_tp
print(paste0('In-bag RMSE = ' , round(rmse,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse_scaled*100,2), "%"))

#OOB RMSE
oob.rmse<- rmse(exp(rf.full$predicted.oob), exp(TP$logTP))
oob.rmse_scaled <- oob.rmse/range_tp
print(paste0('OOB RMSE = ' , round(oob.rmse,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse_scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
rf.full$err.rate

# VIP
plot(rf.full, plots.one.page=FALSE)
top_vimp <- sort(rf.full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "Full model VIMP")

#print(vimp(rf.full)$importance)

# Make a df for plots
predicted.oob <- data.frame(obs = TP$logTP, pred = rf.full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib <- data.frame(obs =TP$logTP, pred = rf.full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf <- rbind(predicted.oob, predicted.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Total Phosphorus Full Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.02, y = 0.3, label = paste0('OOB RMSE = ' , round(oob.rmse,2), ' | scaled RMSE = ', round(oob.rmse_scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.02, y = 0.39, label = paste0('IB RMSE = ' , round(rmse,2), ' | scaled RMSE = ', round(rmse_scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.02, y = 0.5, label = paste0('% variance explained = ' , round(per_var,2), '%'), size = 3) 
```

#### Leaflet map of full model

```{r}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TP table. Predictions will be in the same order as the TP table initially.

# Join predictions to cal_val table
TP.obs_predict <- cbind(plotdf, TP)

# Order the rows based on a StreamCat variables that has ALL unique values.
# any(duplicated(TP$ELEVWS))
# any(duplicated(TP$PRECIP9120WS))
TP.obs_predict <- TP.obs_predict %>% 
  arrange(predictions, ELEVWS)

TP.awqms <- tp_vars_selected %>% 
  arrange(ELEVWS) %>% 
  select(MLocID:logTP)

# Join cond_vars_selected with cond.obs_predict, and makes sure the conductivity results match.
TP.map <- cbind(TP.awqms,TP.obs_predict)

# Then filter to include only OOB predictions.
TP.map <- TP.map %>% 
  select(MLocID:delta, BFIWS:L3Eco) %>% 
  filter(predictions=='OOB')

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Observed Total Phosphorus: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$pred), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Predicted Total Phosphorus: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
                                  "<strong>", "Obs - Pred Total Phosphorus: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$obs, position = "bottomright" , title = "Observed Total Phos" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = data$obs, position = "bottomright" , title = "Predicted Total Phos" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred Total Phos" )
}

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-0.5,0.5))

map_function(TP.map)
```

#### Partial dependence plots for full model

```{r}
# Top importance vars
partialplots_imp <- plot.variable(rf.full, xvar.names = c("PRECIP_MINUS_EVTWS", "PCTCONIF2001WS", "PESTIC1997WS", "PCTBL2001", "ROCKNWS", "RDDENSWS", "WTDEPWS", "RCKDEPWS", "SEPTICWS"), partial = TRUE, plot.sone.page = TRUE)

partialplots_imp2 <- plot.variable(rf.full, xvar.names = c("KFFACTWS", "TMAX9120WS", "ELEVWS", "WETINDEXWS", "SWS", "OMWS", "HYDRLCONDWS", "PCTCROP", "RDCRSWS"), partial = TRUE, plot.sone.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(rf.full, xvar.names = c("PESTIC1997WS", "RDDENSWS", "SEPTICWS", "PCTCROP2001WS", "RDCRSWS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Residual plots

```{r}
# Plot residuals 
residuals.rf.full <- exp(TP$logTP) - exp(predicted.oob$pred) # Observed minus predicted oob
plot(residuals.rf.full ~ exp(predicted.oob$pred), xlab = "Predicted Total Phosphorus (mg/L)", ylab = "Residuals", main = "full model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=TP.map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=TP.map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# Which StreamCat column is most correlated with residuals?
TP.map_num <- TP.map %>% 
  select(!c(MLocID:StationDes, SampleStartDate, ReferenceSite:COMID, predictions,SITE_TYPE:L3Eco))
correlations <- apply(TP.map_num, 2, cor, x = TP.map_num$delta)

resid_cor <- as.data.frame(correlations)

```

#### Reduce all human-influenced variables to 0 to learn impact on the residuals

```{r}
# Reduce all human-influenced vars to all 0s.
TP_no.hi <- TP %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMNRMSTORWS", "FERTWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTCROP2001WS", "PCTHAY2001WS", 
"PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")), ~0))

# Rerun model with adjusted human-influence 
set.seed(24)

rf.full_no.hi <- rfsrc(logTP ~ ., data = TP_no.hi, mtry = 57, nodesize = 2, importance = TRUE)
rf.full_no.hi # Model output. (OOB) R squared = Percent Variance Explained.

# Make a df for no_hi plots
predicted.oob_no.hi <- data.frame(obs = TP_no.hi$logTP, pred = rf.full_no.hi$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib_no.hi <- data.frame(obs =TP_no.hi$logTP, pred = rf.full_no.hi$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.

# Mean squared error (untransformed)
oob.mse_no.hi <- tail(rf.full_no.hi$err.rate,1)
per.var_no.hi <- 100 * (1 - oob.mse_no.hi/ var(TP_no.hi$logTP, na.rm = TRUE))

#In-bag RMSE
rmse_no.hi<- rmse(exp(rf.full_no.hi$predicted), exp(TP_no.hi$logTP))
rmse.scaled_no.hi <- rmse_no.hi/range_tp
print(paste0('In-bag RMSE = ' , round(rmse_no.hi,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse.scaled_no.hi*100,2), "%"))

#OOB RMSE
oob.rmse_no.hi<- rmse(exp(rf.full_no.hi$predicted.oob), exp(TP_no.hi$logTP))
oob.rmse.scaled_no.hi <- oob.rmse_no.hi/range_tp
print(paste0('OOB RMSE = ' , round(oob.rmse_no.hi,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse.scaled_no.hi*100,2), "%"))

# VIP
#plot(rf.full_no.hi, plots.one.page=FALSE)
top.vip_no.hi <- sort(rf.full_no.hi$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top.vip_no.hi, horiz = TRUE, cex.names = 0.5, las = 1, main = "Full model VIMP")

# Make a df for plots
predicted.oob_no.hi <- data.frame(obs = TP_no.hi$logTP, pred = rf.full_no.hi$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib_no.hi <- data.frame(obs =TP_no.hi$logTP, pred = rf.full_no.hi$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf_no.hi <- rbind(predicted.oob_no.hi, predicted.ib_no.hi) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf_no.hi, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TP Full Model without human influenced vars",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.02, y = 0.3, label = paste0('OOB RMSE = ' , round(oob.rmse_no.hi,2), ' | scaled RMSE = ', round(oob.rmse.scaled_no.hi*100,2), "%"), size = 3) +
    annotate("text", x = 0.02, y = 0.39, label = paste0('IB RMSE = ' , round(rmse_no.hi,2), ' | scaled RMSE = ', round(rmse.scaled_no.hi*100,2), "%"), size = 3) +
      annotate("text", x = 0.02, y = 0.5, label = paste0('% variance explained = ' , round(per.var_no.hi,2), '%'), size = 3) 

# Plot residuals - full model, adjusted human-influenced vars to 0
residuals.rf.full_no.hi <- exp(TP_no.hi$logTP) - exp(predicted.oob_no.hi$pred) # Observed minus predicted oob
plot(residuals.rf.full_no.hi ~ exp(predicted.oob_no.hi$pred), xlab = "Predicted TP (mg/L)", ylab = "Residuals", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)

# Partial Dependence for top vars
plot.variable(rf.full_no.hi, xvar.names = c("PCTCONIF2001WS", "PRECIP_MINUS_EVTWS", "WTDEPWS", "PCTBL2001WS", "ROCKNWS", "TMAX9120WS", "WETINDEXWS", "ELEVWS", "RCKDEPWS"), partial = TRUE, plots.one.page = TRUE)
```

#### Look at map after adjusting human vbls to 0

```{r}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the cond.src table. Predictions will be in the same order as the cond.src table initially.

# Join predictions to cal_val table
TP.obs_predict_no.hi <- cbind(plotdf_no.hi, TP_no.hi)

# Order the rows based on a StreamCat variables that has ALL unique values.
# any(duplicated(cond.src$ELEVWS))
# any(duplicated(cond.src$PRECIP9120WS))
TP.obs_predict_no.hi <- TP.obs_predict_no.hi %>% 
  arrange(predictions, ELEVWS)

# Join TP.obs_predict_no.hi with TP.awqms, and makes sure the conductivity results match.
TP.map_no.hi <- cbind(TP.awqms,TP.obs_predict_no.hi)

# Then get rid of StreamCat columns.
TP.map_no.hi <- TP.map_no.hi %>% 
  select(MLocID:delta, L2Eco, L3Eco) %>% 
  filter(predictions=='OOB')

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, 0.7))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-0.5,0.5))

map_function(TP.map_no.hi)

plot(TP.map$pred, TP.map_no.hi$pred, xlab = "full model predictions with HI", ylab = "full model no HI")
abline(0,1)
```

## C. Model Reduction

#### Top 10 predictors

```{r}
set.seed(24)

rf.red10 <- rfsrc(logTP ~ PRECIP_MINUS_EVTWS + PCTCONIF2001WS + PESTIC1997WS + PCTBL2001WS + ROCKNWS + WTDEPWS + RDDENSWS + RCKDEPWS + KFFACTWS + SEPTICWS, data = TP, mtry = 74, nodesize = 2, importance = "TRUE")
rf.red10 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
oob_mse.red10 <- tail(rf.red10$err.rate,1)
per_var.red10 <- 100 * (1 - oob_mse.red10/ var(TP$logTP, na.rm = TRUE))
print(per_var.red10)

#In-bag RMSE
rmse.red10<- rmse(exp(rf.red10$predicted), exp(TP$logTP))
rmse_scaled.red10 <- rmse.red10/range_tp
print(paste0('In-bag RMSE = ' , round(rmse.red10,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse_scaled.red10*100,2), "%"))

#OOB RMSE
oob.rmse.red10<- rmse(exp(rf.red10$predicted.oob), exp(TP$logTP))
oob.rmse_scaled.red10 <- oob.rmse.red10/range_tp
print(paste0('OOB RMSE = ' , round(oob.rmse.red10,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse_scaled.red10*100,2), "%"))

# Model tuning - How many trees? (default: 500)
rf.red10$err.rate

# VIP
plot(rf.red10, plots.one.page=FALSE)
top_vimp.red10 <- sort(rf.red10$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top_vimp.red10, horiz = TRUE, cex.names = 0.5, las = 1, main = "Reduced (10 variables) model VIMP")

#print(vimp(rf.red10)$importance)

# Make a df for plots
predicted.oob.red10 <- data.frame(obs = TP$logTP, pred = rf.red10$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib.red10 <- data.frame(obs =TP$logTP, pred = rf.red10$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf.red10 <- rbind(predicted.oob.red10, predicted.ib.red10) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf.red10, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Total Phosphorus Red10 Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.02, y = 0.3, label = paste0('OOB RMSE = ' , round(oob.rmse.red10,2), ' | scaled RMSE = ', round(oob.rmse_scaled.red10*100,2), "%"), size = 3) +
    annotate("text", x = 0.02, y = 0.39, label = paste0('IB RMSE = ' , round(rmse.red10,2), ' | scaled RMSE = ', round(rmse_scaled.red10*100,2), "%"), size = 3) +
      annotate("text", x = 0.02, y = 0.5, label = paste0('% variance explained = ' , round(per_var.red10,2), '%'), size = 3) 
```

#### Top 3 predictors

```{r}
set.seed(24)

rf.red3 <- rfsrc(logTP ~ PRECIP_MINUS_EVTWS + PCTCONIF2001WS + RDDENSWS, data = TP, mtry = 74, nodesize = 2, importance = "TRUE")
rf.red3 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
oob_mse.red3 <- tail(rf.red3$err.rate,1)
per_var.red3 <- 100 * (1 - oob_mse.red3/ var(TP$logTP, na.rm = TRUE))
print(per_var.red3)

#In-bag RMSE
rmse.red3<- rmse(exp(rf.red3$predicted), exp(TP$logTP))
rmse_scaled.red3 <- rmse.red3/range_tp
print(paste0('In-bag RMSE = ' , round(rmse.red3,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse_scaled.red3*100,2), "%"))

#OOB RMSE
oob.rmse.red3<- rmse(exp(rf.red3$predicted.oob), exp(TP$logTP))
oob.rmse_scaled.red3 <- oob.rmse.red3/range_tp
print(paste0('OOB RMSE = ' , round(oob.rmse.red3,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse_scaled.red3*100,2), "%"))

# Model tuning - How many trees? (default: 500)
rf.red3$err.rate

# VIP
plot(rf.red3, plots.one.page=FALSE)
top_vimp.red3 <- sort(rf.red3$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top_vimp.red3, horiz = TRUE, cex.names = 0.5, las = 1, main = "Reduced (3 variables) model VIMP")

#print(vimp(rf.red3)$importance)

# Make a df for plots
predicted.oob.red3 <- data.frame(obs = TP$logTP, pred = rf.red3$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib.red3 <- data.frame(obs =TP$logTP, pred = rf.red3$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf.red3 <- rbind(predicted.oob.red3, predicted.ib.red3) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf.red3, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Total Phosphorus Red3 Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.02, y = 0.3, label = paste0('OOB RMSE = ' , round(oob.rmse.red3,2), ' | scaled RMSE = ', round(oob.rmse_scaled.red3*100,2), "%"), size = 3) +
    annotate("text", x = 0.02, y = 0.39, label = paste0('IB RMSE = ' , round(rmse.red3,2), ' | scaled RMSE = ', round(rmse_scaled.red3*100,2), "%"), size = 3) +
      annotate("text", x = 0.02, y = 0.5, label = paste0('% variance explained = ' , round(per_var.red3,2), '%'), size = 3) 
```

## D. Adjust human-influenced vars down in Red10 model

#### Subset the data

```{r}
TP.ref <- tp_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, Total_phosphorus, SampleStartDate)) # Remove AWQMS columns

# Convert character columns to factors
TP.ref[sapply(TP.ref, is.character)] <- lapply(TP.ref[sapply(TP.ref, is.character)], as.factor)
```

#### Distributions for Human-influenced vars at all reference sites

```{r}
# Select only the 4 human influenced predictors from the reduced model.
TP.ref.hi <- TP.ref %>% 
  #filter(cal_val == 'CAL') %>% 
  select(c(RDDENSWS, PESTIC1997WS, SEPTICWS))

# Mean pesticide use kg/sq km
# Want to scale back to 0 (50th percentile = 0)
par(mfrow=c(1,2))
plot.variable(rf.red10, xvar.names = c("PESTIC1997WS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP.ref.hi$PESTIC1997WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP.ref.hi$PESTIC1997WS, main = '')
quantile(TP.ref.hi$PESTIC1997WS, probs = seq(.1, .9, by = .1))

# Developed, Low Intensity Land Use Percentage
# Scale back to 0 (50th percentile = 0)
par(mfrow=c(1,2))
plot.variable(rf.red10, xvar.names = c("RDDENSWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP.ref.hi$RDDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP.ref.hi$RDDENSWS, main = '')
quantile(TP.ref.hi$RDDENSWS, probs = seq(.1, .9, by = .1))

# Water Input (km2/cm): Ratio of the total area of irrigated land to precipitation within AOI
# Scale back to 0 (50th percentile = 0)
par(mfrow=c(1,2))
plot.variable(rf.red10, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(TP.ref.hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(TP.ref.hi$SEPTICWS, main = '')
quantile(TP.ref.hi$SEPTICWS, probs = seq(.1, .9, by = .1))

```

#### Adjust human-influenced variables to ref levels

```{r}
TP_ref.adj <- TP %>% 
  select(c(logTP, PRECIP_MINUS_EVTWS, PCTCONIF2001WS, PESTIC1997WS, PCTBL2001WS, ROCKNWS, WTDEPWS, RDDENSWS, RCKDEPWS, KFFACTWS, SEPTICWS)) %>% 
  
    mutate(PESTIC1997WS = case_when(
    PESTIC1997WS > 0 ~ 0, 
    TRUE ~ PESTIC1997WS)) %>% 
  
    mutate(RDDENSWS = case_when(
    RDDENSWS > 0 ~ 0, 
    TRUE ~ RDDENSWS)) %>% 
  
    mutate(SEPTICWS = case_when(
    SEPTICWS > 0 ~ 0, 
    TRUE ~ SEPTICWS)) 
```

#### Top 10 predictors *ref adjusted*

```{r}
set.seed(24)

rf.red10adj <- rfsrc(logTP ~ ., data = TP_ref.adj, mtry = 74, nodesize = 2, importance = "TRUE")
rf.red10adj # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
oob_mse.red10adj <- tail(rf.red10adj$err.rate,1)
per_var.red10adj <- 100 * (1 - oob_mse.red10/ var(TP$logTP, na.rm = TRUE))
print(per_var.red10adj)

#In-bag RMSE
rmse.red10adj<- rmse(exp(rf.red10adj$predicted), exp(TP$logTP))
rmse_scaled.red10adj <- rmse.red10adj/range_tp
print(paste0('In-bag RMSE = ' , round(rmse.red10adj,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse_scaled.red10adj*100,2), "%"))

#OOB RMSE
oob.rmse.red10adj<- rmse(exp(rf.red10adj$predicted.oob), exp(TP$logTP))
oob.rmse_scaled.red10adj <- oob.rmse.red10adj/range_tp
print(paste0('OOB RMSE = ' , round(oob.rmse.red10adj,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse_scaled.red10adj*100,2), "%"))

# Model tuning - How many trees? (default: 500)
rf.red10adj$err.rate

# VIP
plot(rf.red10adj, plots.one.page=FALSE)
top_vimp.red10adj <- sort(rf.red10adj$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top_vimp.red10adj, horiz = TRUE, cex.names = 0.5, las = 1, main = "Reduced (10 variables) model VIMP")

#print(vimp(rf.red10)$importance)

# Make a df for plots
predicted.oob.red10adj <- data.frame(obs = TP$logTP, pred = rf.red10adj$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib.red10adj <- data.frame(obs =TP$logTP, pred = rf.red10adj$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf.red10adj <- rbind(predicted.oob.red10adj, predicted.ib.red10adj) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf.red10adj, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Total Phosphorus Red10 Model with adjusted human-influenced vars",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.02, y = 0.3, label = paste0('OOB RMSE = ' , round(oob.rmse.red10adj,2), ' | scaled RMSE = ', round(oob.rmse_scaled.red10adj*100,2), "%"), size = 3) +
    annotate("text", x = 0.02, y = 0.39, label = paste0('IB RMSE = ' , round(rmse.red10adj,2), ' | scaled RMSE = ', round(rmse_scaled.red10adj*100,2), "%"), size = 3) +
      annotate("text", x = 0.02, y = 0.5, label = paste0('% variance explained = ' , round(per_var.red10adj,2), '%'), size = 3) 
```

## E. Reference-Only Model

#### Subset the data

```{r}
TP.ref <- tp_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, Total_phosphorus, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMNRMSTORWS", "FERTWS", "MINEDENSWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTCROP2001WS", "PCTHAY2001WS", 
"PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "SW_FLUXWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")) # Remove anthropogenic StreamCat vars

# Convert character columns to factors
TP.ref[sapply(TP.ref, is.character)] <- lapply(TP.ref[sapply(TP.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r}
set.seed(2) # need to set seed to make tuning reproducible.

#Adjust the parameters below based on tuning (nodesize = 2; mtry = 47)
tuning <- tune(logTP~., TP.ref)
tuning$optimal

```

#### Build full ref-only model

```{r}
set.seed(24)

rf.full.ref <- rfsrc(logTP ~ ., data = TP.ref, mtry = 47, nodesize = 2, importance = "TRUE")
rf.full.ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
oob_mse.ref <- tail(rf.full.ref$err.rate,1)
per_var.ref <- 100 * (1 - oob_mse.ref/ var(TP.ref$logTP, na.rm = TRUE))
print(per_var.ref)

range_tp_ref <- max(exp(TP.ref$logTP)) - min(exp(TP.ref$logTP))

#In-bag RMSE
rmse.ref<- rmse(exp(rf.full.ref$predicted), exp(TP.ref$logTP))
rmse_scaled.ref <- rmse.ref/range_tp_ref
print(paste0('In-bag RMSE = ' , round(rmse.ref,2), " (μS/cm)"))
print(paste0('Scaled In-bag RMSE = ' , round(rmse_scaled.ref*100,2), "%"))

#OOB RMSE
oob.rmse.ref<- rmse(exp(rf.full.ref$predicted.oob), exp(TP.ref$logTP))
oob.rmse_scaled.ref <- oob.rmse.ref/range_tp_ref
print(paste0('OOB RMSE = ' , round(oob.rmse.ref,2), " (μS/cm)"))
print(paste0('Scaled OOB RMSE = ' , round(oob.rmse_scaled.ref*100,2), "%"))

# Model tuning - How many trees? (default: 500)
rf.full.ref$err.rate

# VIP
plot(rf.full.ref, plots.one.page=FALSE)
top_vimp.ref <- sort(rf.full.ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(top_vimp.ref, horiz = TRUE, cex.names = 0.5, las = 1, main = "Full Reference-Only Model VIMP")

#print(vimp(rf.full.ref)$importance)

# Make a df for plots
predicted.oob.ref <- data.frame(obs = TP.ref$logTP, pred = rf.full.ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
predicted.ib.ref <- data.frame(obs =TP.ref$logTP, pred = rf.full.ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
plotdf.ref<- rbind(predicted.oob.ref, predicted.ib.ref) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = plotdf.ref, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "Total Phosphorus Full Reference-Only Model",
       x = "Predicted values (mg/L)",
       y = "Observed values (mg/L)") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.009, y = 0.09, label = paste0('OOB RMSE = ' , round(oob.rmse.ref,2), ' | scaled RMSE = ', round(oob.rmse_scaled.ref*100,2), "%"), size = 3) +
    annotate("text", x = 0.009, y = 0.10, label = paste0('IB RMSE = ' , round(rmse.ref,2), ' | scaled RMSE = ', round(rmse_scaled.ref*100,2), "%"), size = 3) +
      annotate("text", x = 0.009, y = 0.11, label = paste0('% variance explained = ' , round(per_var,2), '%'), size = 3) 
```
