---
title: "EPA_test_TN"
output: html_document
---
# Taking data from EPA/Lester Yuan (ss.bugs) and plugging it into our Benchmarks code structure. No need to pull from StreamCat since ss.bugs comes with all predictor data.

# **1 - LOAD PACKAGES**

```{r load packages, warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(maps)
library(RColorBrewer)
library(htmlwidgets)
```

```{r pre-define inputs}

units <- "(mg/L)" # For plot labels

# plot labels
x <- 0.18
y1 <- 5
y2 <- 7
y3 <- 10

# number of variables to include in the reduced model
nred <- 11

# Character vector of all human-influenced variables
HI_vars <- c("canaldensws", "coalminedensws", "damdensws", "damnrmstorws", "manurews", "minedensws", "minedenswsrp100", "npdesdensws","pctagslpmid20XXws", "pctagslphigh20XXws", "pctcrop20XXws", "pctfrstloss20XXws", "pcthay20XXws", "pctimp20XXws", "pctnonagintrodmanagvegws","pcturbop20XXws", "rdcrsws", "superfunddensws", "superfunddenswsrp100", "tridenswsrp100")

```


# **2 - IMPORT DATA**

```{r}
load("~/BioMonORDEQ/Benchmarks/Water Chemistry/EPA National Model/ss.bugs.rda")

epa_data <- as.data.frame(ss.bugs[2]) 
colnames(epa_data) <- gsub("sstest.", "", colnames(epa_data))

epa_data <- epa_data %>% 
  select(-c(ABISKOMYIA:ZUMATRICHIA)) %>%  # Select out the bugs
  relocate(site.id2, .after = site.id) %>% 
  relocate(date.col, .after = year) %>% 
  filter(protocol == "WADEABLE") %>%  # Drop boatable sites
  select(-c(fstfrz.pt,kffact.pt,rhmean.pt,pctcarbonatews,pctsandstone.carbonatews,pctsandstonews,pctsemiconsol.sandws,pctvolcanicws,elev.delt,hort,bfi.gam,bfi.rf,ref2)) %>%   # Per 6/22 communication w Lester Yuan, drop these vars as they are no longer present in StreamCat.
  select(-c(elevation)) # drop in favor of elevws from SC

# Also per LY and the paper, add other SC vars
comids <- epa_data$comid

ws<-sc_get_data(comids, metric = c('bfi','damdens','pctagslphigh2006','precip9120','elev'), aoi = 'ws')
# Not sure why these numbers are all negative?? Multiplying by -1 to fix it
other<--sc_get_data(comids, metric = c('bankfulldepth','bankfullwidth','mwst2014'), aoi = 'other') * -1 

epa_data <- epa_data %>% 
  left_join(ws, by = "comid") %>% 
  left_join(other, by = "comid") %>% 
  select(-c(ppt, wdsum.pt))

rm(list=c("ws", "other","ss.bugs"))
```

```{r}
# Make EPA table look more like ORDEQ table
epa_data <- epa_data %>% 
  rename(Lat_DD = lat.dd83, 
         Long_DD = lon.dd83,
         COMID = comid,
         SampleStartDate = date.col) %>% 
  select(uid:xfc.alg, order(colnames(.))) %>% 
  # Remove years from EPA table so colnames can match DEQ table later
  rename(
         inorgnwetdep20XXws = inorgnwetdep.2008ws,
         msst.20XX = msst.2019,
         mwst.20XX = mwst2014,
         pctagslpmid20XXws = pctag2006slp10ws,
         pctagslphigh20XXws = pctagslphigh2006ws,
         pctbl20XXws = pctbl2019ws,
         pctconif20XXws = pctconif2019ws,
         pctcrop20XXws = pctcrop2019ws,
         pctdecid20XXws = pctdecid2019ws,
         pctfire20XXws = pctfire2010ws,
         pctfrstloss20XXws = pctfrstloss2013ws,
         pctgrs20XXws = pctgrs2019ws,
         pcthay20XXws = pcthay2019ws,
         pcthbwet20XXws = pcthbwet2019ws,
         pctice20XXws = pctice2019ws,
         pctimp20XXws = pctimp2019ws,
         pctmxtfst20XXws = pctmxfst2019ws,
         pctow20XXws = pctow2019ws,
         pctshrb20XXws = pctshrb2019ws,
         pcturbop20XXws = pcturbop2019ws,
         pctwdwet20XXws = pctwdwet2019ws) %>% 
  mutate(pctcoastcrsws = as.numeric(pctcoastcrsws)) 
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r handle NAs}
# Drop rows where TN result has NA
epa_data <- epa_data %>% 
  drop_na(ntl)
```

#### PERFORM NA ROUGH FIX

```{r NA roughfix}
sort(sapply(epa_data %>% select(c(bankfulldepth:rt.master)), function(x) sum(is.na(x))))

# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH MISSING VALUES
epa_data$bfiws  <- na.roughfix(epa_data$bfiws)
epa_data$damdensws <- na.roughfix(epa_data$damdensws) 
epa_data$pctagslphigh20XXws <- na.roughfix(epa_data$pctagslphigh20XXws) 
epa_data$precip9120ws <- na.roughfix(epa_data$precip9120ws) 
epa_data$bankfulldepth <- na.roughfix(epa_data$bankfulldepth) 
epa_data$bankfullwidth <- na.roughfix(epa_data$bankfullwidth) 
epa_data$mwst.20XX <- na.roughfix(epa_data$mwst.20XX) 
epa_data$elevws <- na.roughfix(epa_data$elevws)
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r plot response variable}
# HISTOGRAM
hist(epa_data$ntl, main = "", xlab = "TN") #raw data
hist(log(epa_data$ntl), main = "", xlab = "log(TN)", breaks = 15)

# BOX PLOTS
boxplot(epa_data$ntl, ylab = "TN")
boxplot(log(epa_data$ntl), ylab = "log(TN)")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r prep response data}
# Select response data of interest
TN <- epa_data %>%
  select(-c(uid:ag.eco9, COMID:ptl, chloride.result:xfc.alg,rt.master)) %>% # select ntl
  mutate(ntl = ntl/1000) %>%  # convert micrograms/L to mg/L
  mutate(logTN = log(ntl)) %>% 
  relocate(logTN, .before = everything()) %>% 
  select(-c(ntl)) %>% 
  as.data.frame() %>% # Make sure TN is a data frame, not a tibble
  select(1, sort(names(.)))

# for sorting purposes later
TN_uid <- epa_data %>%
  select(c(uid, COMID,ntl,Lat_DD,Long_DD,us.l3code, state.nm,SampleStartDate)) %>% 
  mutate(ntl = ntl/1000) %>%  # convert micrograms/L to mg/L
  mutate(logTN = log(ntl))

# Ref only data (rt.master = B or S, immutable vars only)
TN_ref <- epa_data %>% 
  select(-c(uid:ag.eco9, COMID:ptl, chloride.result:xfc.alg)) %>% # select ntl
  mutate(ntl = ntl/1000) %>%  # convert micrograms/L to mg/L
  mutate(logTN = log(ntl)) %>% 
  relocate(logTN, .before = everything()) %>% 
  select(-c(ntl)) %>% 
  as.data.frame() %>% # Make sure TN is a data frame, not a tibble
  select(1, sort(names(.))) %>% 
  filter(rt.master=="B" | rt.master == "S") %>% 
  select(-c(rt.master)) %>% 
  select(-c(HI_vars))

# for sorting purposes later
TN_uid.ref <- epa_data %>% 
  filter(rt.master=="B" | rt.master == "S") %>% 
  select(c(uid, COMID,ntl,Lat_DD,Long_DD,us.l3code, state.nm,SampleStartDate)) %>% 
  mutate(ntl = ntl/1000) %>%  # convert micrograms/L to mg/L
  mutate(logTN = log(ntl))

# Convert character columns to factors
TN[sapply(TN, is.character)] <- lapply(TN[sapply(TN, is.character)], as.factor)
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r full model tuning}
#Adjust the parameters below based on tuning
full_tuning <- tune(logTN~., data = TN, seed = 1234)
full_tuning$optimal
```

#### Build full model

```{r full model}
full <- rfsrc(logTN ~ ., data = TN, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 2)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(TN$logTN, na.rm = TRUE))
print(full_per.var)

range_TN <- max(exp(TN$logTN)) - min(exp(TN$logTN))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(TN$logTN))
full_rmse.scaled <- full_rmse/range_TN
print(paste("In-bag RMSE = ", round(full_rmse,2), units))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(TN$logTN))
full_oob.rmse.scaled <- full_oob.rmse/range_TN
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), units))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TN Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = TN$logTN, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =TN$logTN, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN Full Model",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y1, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = x, y = y2, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = x, y = y3, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```

#### Partial dependence plots for full model

```{r partial dependence plots for full model}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c(names(full_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c(names(full_vimp[9:16])), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("pctcrop20XXws","npdesdensws", "pctimp20XXws", "pcthay20XXws","canaldensws","pctfrstloss20XXws","tridenswsrp100","manurews"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```
#### Another way to do partial plots (slower but more customizable)
```{r more partial plots for full model}
# partial.obj <- partial(full, partial.xvar = "pctimp2019ws", partial.values = full$xvar$pctimp2019ws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctimp2019ws", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "runoffws", partial.values = full$xvar$runoffws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "runoffws", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "pctforest", partial.values = full$xvar$pctforest)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctforest", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "ppt", partial.values = full$xvar$ppt)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "ppt", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "pctcrop2019ws", partial.values = full$xvar$pctcrop2019ws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctcrop2019ws", ylab = "ŷ")
```

#### Leaflet map of full model

```{r leaflet map of full model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TN table. Predictions will be in the same order as the TN table initially.

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, TN) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the TN results match (Make sure logTN column is the same as logTN.1)
full_map <- cbind(full_obs.predict,TN_uid) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:wtdepws)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -98.583, lat =39.833, zoom =4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(log(data$obs)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Observed TN: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(log(data$pred)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Predicted TN: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Obs - Pred TN: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = log(data$obs), position = "bottomright" , title = "Observed logTN" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = log(data$obs), position = "bottomright" , title = "Predicted logTN" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred COND" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(log(min(full_map$obs)), log(max(full_map$obs))))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(log(min(full_map$obs)), log(max(full_map$obs))))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map$delta),max(full_map$delta)))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TN//full_model_map.html", selfcontained = TRUE, title = "Full model")
```
#### Map of predictors  

```{r Map predictors}

# define predictor
predictor <- "pctcrop20XXws"
predictor_map <- map_data("usa", region = ".")

ggplot() + 
  geom_polygon(data = predictor_map, aes(x = long, y = lat, group = group),
               fill = "white", color = "black") +
  geom_point(data = full_map, aes(x = Long_DD, y = Lat_DD, fill = get(predictor)), shape = 21, size = 2.5) + 
  scale_fill_gradient(low = "white", high = "green", name = predictor) +
  #scale_size_area(max_size = 8) +
  scale_size_continuous(breaks = c(1, 3, 5 ,10), name = predictor) +
  theme_void() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  labs(title = predictor)
```

#### Residual plots
```{r full model residual plots}
# Plot residuals 
full_resid <- exp(TN$logTN) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = paste("Predicted TN", units), ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(TN$logTN) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r adj full model}
# Reduce all human-influenced vars to all 0s.
TN_no.hi <- TN %>% 
  mutate(across(all_of(HI_vars), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= TN_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(TN_no.hi$logTN, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(TN_no.hi$logTN))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_TN
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), units))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = TN_no.hi$logTN, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN Full Model predictions without human influenced vars",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y3, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(TN_no.hi$logTN) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

## C. Model Reduction

#### Build model with top predictors

```{r reduced model tuning}
# build formula
red_vars <- as.formula(paste("logTN~", paste(names(full_vimp)[1:nred], collapse = "+")))

#Adjust the parameters below based on tuning 
red11_tuning <- tune(red_vars, data = TN, seed = 1234)

paste0(nred_text,"_tuning") 
red11_tuning$optimal
```

```{r reduced model}
red11 <- rfsrc(red_vars, data = TN, mtry = red11_tuning$optimal[[2]], nodesize = red11_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red11 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red11_oob.mse <- tail(red11$err.rate,1)
red11_per.var <- 100 * (1 - red11_oob.mse/ var(TN$logTN, na.rm = TRUE))
print(red11_per.var)

#In-bag RMSE
red11_rmse<- rmse(exp(red11$predicted), exp(TN$logTN))
red11_rmse.scaled <- red11_rmse/range_TN
print(paste0('In-bag RMSE = ' , round(red11_rmse,2), units))
print(paste0('Scaled In-bag RMSE = ' , round(red11_rmse.scaled*100,2), "%"))

#OOB RMSE
red11_oob.rmse<- rmse(exp(red11$predicted.oob), exp(TN$logTN))
red11_oob.rmse.scaled <- red11_oob.rmse/range_TN
print(paste0('OOB RMSE = ' , round(red11_oob.rmse,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red11_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red11$err.rate

# VIP
plot(red11, plots.one.page=FALSE)
red11_vimp <- sort(red11$importance, decreasing = TRUE)[1:nred]
par(mar=c(5,7,4,1)+.1)
barplot(red11_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TN Reduced (red11) model VIMP")

#print(vimp(red11)$importance)

# Make a df for plots
red11_pred.oob <- data.frame(obs = TN$logTN, pred = red11$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red11_pred.ib <- data.frame(obs =TN$logTN, pred = red11$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red11_plotdf <- rbind(red11_pred.oob, red11_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red11_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red11 Model",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y1, label = paste0('OOB RMSE = ' , round(red11_oob.rmse,2), ' | scaled RMSE = ', round(red11_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = x, y = y2, label = paste0('IB RMSE = ' , round(red11_rmse,2), ' | scaled RMSE = ', round(red11_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = x, y = y3, label = paste0('% variance explained = ' , round(red11_per.var,2), '%'), size = 3) 
```

#### Partial dependence plots for red11 model

```{r partial dependence plots for reduced model}
# Top importance vars
red11_partialplots <- plot.variable(red11, xvar.names = c(names(red11_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

red11_partialplots2 <- plot.variable(red11, xvar.names = c(names(red11_vimp[9:11])), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
red11_partialplots_hi <- plot.variable(red11, xvar.names = c("pctcrop20XXws","npdesdensws", "pctimp20XXws", "pcthay20XXws","canaldensws","pctfrstloss20XXws"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of red11 model

```{r leaflet map of reduced model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the TN table. Predictions will be in the same order as the TN table initially.


# Join predictions to StreamCat table
red11_obs.predict <- cbind(red11_plotdf, TN) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the TN results match (Make sure logTN column is the same as logTN.1)
red11_map <- cbind(red11_obs.predict,TN_uid) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:wtdepws)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, log(red11_map$obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, log(red11_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(red11_map$delta),max(red11_map$delta)))

map_function(red11_map)

# Save map as html
# red11_model_map <- map_function(red11_map)
# saveWidget(red11_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TN//red11_model_map.html", selfcontained = TRUE, title = "red11 model")
```

#### Overlay reduced and full models

```{r reduced and full model prediction plot}
full_red_map <- left_join(full_map, red11_map %>% select(uid, pred), by = "uid") %>% 
  mutate(full_pred = pred.x) %>% 
  mutate(red11_pred = pred.y) %>% 
  select(-c(pred.x,pred.y)) %>% 
  relocate(red11_pred, .after = obs) %>% 
  relocate(full_pred, .after = obs) %>% 
  select(uid:predictions) %>% 
  gather(key = "Model", value = "Prediction", full_pred, red11_pred)

ggplot(full_red_map, aes(x = Prediction, y = obs, color = Model)) +
  geom_point(size = 1.5, alpha = 0.5) + theme_bw(base_size = 15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = paste("Predicted TN", units),
    y = paste("Observed TN", units)) + scale_x_log10() + scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```

#### Distributions for Human-influenced vars at all reference sites (N/A since we do not know reference status for EPA sites)

```{r distributions of human influenced variables at all ref sites}
# # Select only the human influenced predictors at reference sites from the reduced model.
# TN_ref_hi <- full_map %>% 
#   filter(ReferenceSite == "REFERENCE") %>% 
#   select(c(MLocID:logTN, SEPTICWS, PCTFRSTLOSS2002WS, PCTURBOP2001WS, CANALDENSWS, PCTNONAGINTRODMANAGVEGWS))
# 
# # PCTFRSTLOSS2002WS
# par(mfrow=c(1,2))
# plot.variable(red11, xvar.names = c("PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(TN_ref_hi$PCTFRSTLOSS2002WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(TN_ref_hi$PCTFRSTLOSS2002WS, main = '')
# quantile(TN_ref_hi$PCTFRSTLOSS2002WS, probs = seq(.1, .9, by = .1))
# 
# # SEPTICWS
# par(mfrow=c(1,2))
# plot.variable(red11, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(TN_ref_hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(TN_ref_hi$SEPTICWS, main = '')
# quantile(TN_ref_hi$SEPTICWS, probs = seq(.1, .9, by = .1))

# BELOW not in reduced model

# # PCTURBOP2001WS
# par(mfrow=c(1,2))
# plot.variable(red11, xvar.names = c("PCTURBOP2001WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(TN_ref_hi$PCTURBOP2001WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(TN_ref_hi$PCTURBOP2001WS, main = '')
# quantile(TN_ref_hi$PCTURBOP2001WS, probs = seq(.1, .9, by = .1))
# 
# # CANALDENSWS
# par(mfrow=c(1,2))
# plot.variable(red11, xvar.names = c("CANALDENSWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(TN_ref_hi$CANALDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(TN_ref_hi$CANALDENSWS, main = '')
# quantile(TN_ref_hi$CANALDENSWS, probs = seq(.1, .9, by = .1))
# 
# # PCTNONAGINTRODMANAGVEGWS
# par(mfrow=c(1,2))
# plot.variable(red11, xvar.names = c("PCTNONAGINTRODMANAGVEGWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(TN_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(TN_ref_hi$PCTNONAGINTRODMANAGVEGWS, main = '')
# quantile(TN_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = seq(.1, .9, by = .1))
```

#### Reduce all human-influenced variables [to 0]* and make predictions using reduced model.
##### * We may decide to adjust to a non-zero percentile based on reference quantiles in the chunk above.

```{r adj reduced model}
# Make predictions with adjusted human-influence to 0 
red11_no.hi <- predict(red11, newdata= TN_no.hi, seed = 24)

# Mean squared error (untransformed)
red11_mse_no.hi <- tail(red11_no.hi$err.rate,1)
red11_per.var_no.hi <- 100 * (1 - red11_mse_no.hi/ var(TN_no.hi$logTN, na.rm = TRUE))

# RMSE
red11_rmse_no.hi<- rmse(exp(red11_no.hi$predicted), exp(TN_no.hi$logTN))
red11_rmse.scaled_no.hi <- red11_rmse_no.hi/range_TN
print(paste0('OOB RMSE = ' , round(red11_rmse_no.hi,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red11_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red11_plotdf_no.hi <- data.frame(obs = TN_no.hi$logTN, pred = red11_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red11_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red11 Model predictions without human influenced vars",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y3, label = paste0('OOB RMSE = ' , round(red11_rmse_no.hi,2), ' | scaled RMSE = ', round(red11_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red11 model, adjusted human-influenced vars to 0
red11_resid_no.hi <- exp(TN_no.hi$logTN) - exp(red11_no.hi$predicted) # Observed minus predicted oob
plot(red11_resid_no.hi ~ exp(red11_no.hi$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "red11 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map after adjusting human vbls for red11 model

```{r leaflet map after adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to StreamCat table
red11_obs.predict_no.hi <- cbind(red11_plotdf_no.hi, TN_no.hi) #%>% 
#  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
red11_map_no.hi <- cbind(red11_obs.predict_no.hi,TN_uid) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:wtdepws)

pal <- colorNumeric(palette = "Purples", 
  domain = c(min(log(red11_map_no.hi$obs)), max(log(red11_map_no.hi$obs))))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(min(log(red11_map_no.hi$obs)), max(log(red11_map_no.hi$obs))))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1* max(red11_map_no.hi$delta), max(red11_map_no.hi$delta)))

map_function(red11_map_no.hi)

# Save map as html
# red11_model_map_no.hi <- map_function(red11_map_no.hi)
# saveWidget(red11_model_map_no.hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//red11_model_map_no.hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```


#### Top 5 predictors (not using this one)

```{r reducing the model further tuning}
#Adjust the parameters below based on tuning
red5_tuning <- tune(logTN ~ pctcrop20XXws+pctforest+pctconif20XXws+npdesdensws+pctimp20XXws, data = TN, seed = 1234)
red5_tuning$optimal
```

```{r reducing the model further}
red5 <- rfsrc(logTN ~ pctcrop20XXws+pctforest+pctconif20XXws+npdesdensws+pctimp20XXws, data = TN, mtry = red5_tuning$optimal[[2]], nodesize = red5_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red5 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red5_oob.mse <- tail(red5$err.rate,1)
red5_per.var <- 100 * (1 - red5_oob.mse/ var(TN$logTN, na.rm = TRUE))
print(red5_per.var)

#In-bag RMSE
red5_rmse<- rmse(exp(red5$predicted), exp(TN$logTN))
red5_rmse.scaled <- red5_rmse/range_TN
print(paste0('In-bag RMSE = ' , round(red5_rmse,2), units))
print(paste0('Scaled In-bag RMSE = ' , round(red5_rmse.scaled*100,2), "%"))

#OOB RMSE
red5_oob.rmse<- rmse(exp(red5$predicted.oob), exp(TN$logTN))
red5_oob.rmse.scaled <- red5_oob.rmse/range_TN
print(paste0('OOB RMSE = ' , round(red5_oob.rmse,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red5_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red5$err.rate

# VIP
plot(red5, plots.one.page=FALSE)
red5_vimp <- sort(red5$importance, decreasing = TRUE)[1:5]
par(mar=c(5,7,4,1)+.1)
barplot(red5_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TN Reduced (4 variables) model VIMP")

#print(vimp(red5)$importance)

# Make a df for plots
red5_pred.oob <- data.frame(obs = TN$logTN, pred = red5$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red5_pred.ib <- data.frame(obs =TN$logTN, pred = red5$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red5_plotdf <- rbind(red5_pred.oob, red5_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red5_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red5 Model",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y1, label = paste0('OOB RMSE = ' , round(red5_oob.rmse,2), ' | scaled RMSE = ', round(red5_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = x, y = y2, label = paste0('IB RMSE = ' , round(red5_rmse,2), ' | scaled RMSE = ', round(red5_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = x, y = y3, label = paste0('% variance explained = ' , round(red5_per.var,2), '%'), size = 3) 

```

#### Reduce all human-influenced variables [to 0]* and make predictions using red5 model.

```{r adj reduced model}
# Make predictions with adjusted human-influence to 0 
red5_no.hi <- predict(red5, newdata= TN_no.hi, seed = 24)

# Mean squared error (untransformed)
red5_mse_no.hi <- tail(red5_no.hi$err.rate,1)
red5_per.var_no.hi <- 100 * (1 - red5_mse_no.hi/ var(TN_no.hi$logTN, na.rm = TRUE))

# RMSE
red5_rmse_no.hi<- rmse(exp(red5_no.hi$predicted), exp(TN_no.hi$logTN))
red5_rmse.scaled_no.hi <- red5_rmse_no.hi/range_TN
print(paste0('OOB RMSE = ' , round(red5_rmse_no.hi,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red5_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red5_plotdf_no.hi <- data.frame(obs = TN_no.hi$logTN, pred = red5_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red5_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red5 Model predictions without human influenced vars",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y2, label = paste0('OOB RMSE = ' , round(red5_rmse_no.hi,2), ' | scaled RMSE = ', round(red5_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red5 model, adjusted human-influenced vars to 0
red5_resid_no.hi <- exp(TN_no.hi$logTN) - exp(red5_no.hi$predicted) # Observed minus predicted oob
plot(red5_resid_no.hi ~ exp(red5_no.hi$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "red5 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

## E. Reference-Only Model

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r tuning ref only model}
#Adjust the parameters below based on tuning 
ref_tuning <- tune(logTN~., TN_ref, seed = 2)
ref_tuning$optimal
```

#### Build full ref-only model

```{r ref only model}
ref <- rfsrc(logTN ~ ., data = TN_ref, mtry = ref_tuning$optimal[[2]], nodesize = ref_tuning$optimal[[1]], importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(TN_ref$logTN, na.rm = TRUE))
print(ref_per.var)

ref_range_TN <- max(exp(TN_ref$logTN)) - min(exp(TN_ref$logTN))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(TN_ref$logTN))
ref_rmse.scaled <- ref.rmse/ref_range_TN
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), units))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(TN_ref$logTN))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_TN
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), units))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "TN Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = TN_ref$logTN, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =TN_ref$logTN, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN Full Reference-Only Model",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y1, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = x, y = y2, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = x, y = y3, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(TN_ref$logTN) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Partial dependence plots for full ref-only model

```{r partial dependence plots for full ref only model}
# Top importance vars
partialplots <- plot.variable(ref, xvar.names = c(names(ref_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of Ref-Only model

```{r leaflet map of ref only model}
# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, TN_ref) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the TN results match (Make sure logTN column is the same as logTN.1)
ref_map <- cbind(ref_obs.predict,TN_uid.ref) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:wtdepws)

pal <- colorNumeric(palette = "BuGn",
  domain = c(log(min(ref_map$obs)), log(max(ref_map$obs))))
pal2 <- colorNumeric(palette = "BuGn",
  domain = c(log(min(ref_map$obs)), log(max(ref_map$obs))))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(ref_map$delta), max(ref_map$delta)))
# 
map_function(ref_map)
# 
# # Save map as html
# # ref_model_map <- map_function(ref_map)
# # saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TNide//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## F. Make comparisons of predictions

#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r comopare reduced model (adjusted and unadjusted) by ecoregion}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(uid = red11_map$uid, Obs = red11_map$obs, Pred_reduced = red11_map$pred, L3Eco = as.character(red11_map$us.l3code), State = red11_map$state.nm, Pred_adj = red11_map_no.hi$pred, Pred_full = full_map$pred)

boxplot_preds <- boxplot_preds %>%
  mutate(L1Eco = case_when(
    L3Eco %in% c("49", "50", "58", "62") ~ "Northern Forests",
    L3Eco %in% c("78","77","41","21","19","17","16","15","11","9","5","4") ~ "Northwestern Forested Mountains",
    L3Eco %in% c("1", "2", "3") ~ "Marine West Coast Forest",
    L3Eco %in% c("33","35","36","37","38","39","45","51","52","53","54","55","56","57","59","60","61","63","64","65","66","67","68","69","70","71","72","73","74","75","82","83","84"
) ~ "Eastern Temperate Forests",
    L3Eco %in% c("25","26","27","28","29","30","31","32","34","40","42","43","44","46","47","48"
) ~ "Great Plains",
    L3Eco %in% c("10","12","13","14","18","20","22","24","80","81"
) ~ "North American Deserts",
    L3Eco %in% c("6","7","8","85"
) ~ "Mediterranean California",
    L3Eco %in% c("79") ~ "Southern Semi-Arid Highlands",
    L3Eco %in% c("23") ~ "Temperate Sierras",
    L3Eco %in% c("76") ~ "Tropical Wet Forests"
  )) %>% 
  mutate(L3Eco = case_when(
    L3Eco == "3" ~ "Willamette Valley",
    L3Eco == "9" ~ "Eastern Cascades Slopes and Foothills",
    L3Eco == "11" ~ "Blue Mountains",
    L3Eco == "78" ~ "Klamath Mountains",
    L3Eco == "1" ~ "Coast Range",
    L3Eco == "4" ~ "Cascades",
    L3Eco == "80" ~ "Northern Basin and Range",
    L3Eco == "10" ~ "Columbia Plateau",
    L3Eco == "12" ~ "Snake River Plain",
    TRUE ~ L3Eco))

boxplot_preds$L3Eco <- factor(boxplot_preds$L3Eco)

boxplot_ref<- ref_map %>%
  select(uid, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="uid")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_full, Pred_reduced, Pred_adj, Pred_ref, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_reduced", "Pred_full", "Obs"))
# boxplot_df$L3Eco <- factor(boxplot_df$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L1Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  labs(title = "EPA model predictions by L1Ecoregion") +
  theme_bw(base_size=14) + ylab(paste("TN", units)) +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5, size = 3
  )

ggplot(boxplot_df %>% filter(L3Eco %in% c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain")), aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  labs(title = "EPA model predictions Oregon ecoregions") +
  theme_bw(base_size=14) + ylab(paste("TN", units)) +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5, size = 3
  ) 
  
# Ignore warning for rows removed - these are NAs in the Pred_ref column.
```

#### Combined Leaflet map of predictions

```{r combinded leaflet map of predictions}
leaflet_df <- left_join(boxplot_df, TN_uid %>%  select(uid, Lat_DD, Long_DD), by = "uid") %>% 
  pivot_wider(names_from = Model, values_from = Prediction) %>% 
  relocate(Obs, .before= Pred_full)

map_all <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$Obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Observed TN: ", "</strong>", data$Obs))  %>% 
      addCircleMarkers(data = data, group = "Pred_reduced",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$Pred_reduced), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted TN UNADJUSTED (reduced model): ", "</strong>", data$Pred_reduced))  %>% 
     addCircleMarkers(data = data, group = "Pred_adj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$Pred_adj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted TN ADJUSTED: ", "</strong>", data$Pred_adj))  %>% 
     addCircleMarkers(data = data, group = "Pred_ref",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$Pred_ref), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted TN REF ONLY: ", "</strong>", data$Pred_ref))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Pred_reduced", "Pred_adj", "Pred_ref"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$Obs, position = "bottomright" , title = "Observed TN" ) %>% 
    addLegend(pal = pal2,group = "Pred_reduced",values = data$Obs, position = "bottomright" , title = "Predicted TN (UNADJUSTED)" ) %>%  # Keep values=data$Obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Pred_adj", values = data$Pred_adj, position = "bottomleft" , title = "Predicted TN (ADJUSTED)" ) %>% 
        addLegend(pal = pal4, group = "Pred_ref", values = data$Pred_adj, position = "bottomleft" , title = "Predicted TN (REF ONLY)" )    # Keep values=data$Pred_adj here (instead of data$Ref) to keep the legend colors scales the same between ref expectations.
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal3 <- colorNumeric(palette = "Oranges", c(0, max(leaflet_df$Pred_full)))
pal4 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_full)), na.color = "transparent") 

map_all(leaflet_df)

# Save map as html
# map_all <- map_all(leaflet_df)
# saveWidget(map_all, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//TNide//map_all.html", selfcontained = TRUE, title = "Map_All")
```
 
#### Unadjusted predictions vs. adjusted human inflence 

```{r unadjusted predictions vs adj human influence}
# leaflet_df$Ref <- factor(leaflet_df$Ref, levels = c("B", "S", "T"))
# 
# ggplot(leaflet_df, aes(x = Pred_reduced, y = Pred_adj, color = Ref, shape = Ref)) +
#   geom_point(size = 1.5) + theme_bw(base_size = 15) +
#   labs(
#     x = paste("Reduced model predicted TN", units),
#     y = paste("Predicted TN", units, "adjusted\n for no human disturbance predictions"),
#     color = "Reference Status", shape = "Reference Status") + 
#   geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```

# **6 - Make predictions on Oregon data**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
deq_TN <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-05-14.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Nitrogen') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>% 
  mutate(TN = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

deq_comids <- deq_TN %>% 
  select(COMID) # Get just the COMIDs for next step
```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

```{r streamcat data}
deq_sc_ws <- read_csv(r"(\\deqlab1\\BioMon\\Projects\\Biomon Redux\\StreamCat data June 2026\\ws.csv)", col_select = 
                         c(comid, elevws,bfiws,canaldensws,caows,clayws,coalminedensws,compstrgthws,damdensws,
                           damnrmstorws, fe2o3ws, hydrlcondws,inorgnwetdep2008ws, k2ows, kffactws,manurews, mgows,
                           na2ows, minedensws,npdesdensws, nws, omws, p2o5ws,pctagslpmid2001ws,pctagslphigh2001ws,
                           pctalkintruvolws,pctalluvcoastws,pctbl2001ws,pctcarbresidws,pctcoastcrsws,pctcolluvsedws,
                           pctconif2001ws,pctcrop2001ws,pctdecid2001ws,pcteolcrsws,pcteolfinews,pctextruvolws,
                           pctfire2002ws, pctfrstloss2002ws,pctglaclakecrsws,pctglaclakefinews,pctglactilclayws,
                           pctglactilcrsws,pctglactilloamws,pctgrs2001ws,pcthay2001ws,pcthbwet2001ws,pcthydricws,
                           pctice2001ws,pctimp2001ws,pctmxfst2001ws,pctnonagintrodmanagvegws,pctnoncarbresidws,
                           pctow2001ws,pctsallakews,pctshrb2001ws,pctsilicicws,pcturbop2001ws,pctwaterws,
                           pctwdwet2001ws,permws,precip9120ws,rckdepws,rdcrsws,runoffws,sio2ws,superfunddensws,
                           wetindexws,wsareasqkm,wtdepws)) %>% 
  filter(comid %in% deq_comids$COMID)

deq_sc_other <- read_csv(r"(\\deqlab1\\BioMon\\Projects\\Biomon Redux\\StreamCat data June 2026\\other.csv)", 
                         col_select = c(comid,bankfullwidth,bankfulldepth,msst2008,mwst2008)) %>% 
  filter(comid %in% deq_comids$COMID)

deq_sc_rp100 <- read_csv(r"(\\deqlab1\\BioMon\\Projects\\Biomon Redux\\StreamCat data June 2026\\wsrp100.csv)", 
                         col_select = c(comid,minedenswsrp100,tridenswsrp100,superfunddenswsrp100)) %>% 
  filter(comid %in% deq_comids$COMID)

deq_sc_all <- deq_sc_ws %>% 
  full_join(deq_sc_other, by = "comid") %>% 
  full_join(deq_sc_rp100, by = "comid") %>% 
  mutate(pctforest = rowSums(across(c(pctconif2001ws, pctdecid2001ws, pctmxfst2001ws)))) # Create pctforest var.

rm(list=c("deq_sc_ws", "deq_sc_other","deq_sc_rp100"))
```

#### Join with AWQMS columns and rename/reorder columns to match EPA table

```{r}
TN_awqms <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-05-14.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Nitrogen') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, COMID)) %>% 
  mutate(TN = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod) %>%  
  mutate(logTN = log(TN)) %>% 
  relocate(logTN, .before= TN) %>% 
  mutate(comid = COMID)

deq_TN <- deq_sc_all %>% 
  left_join(select(TN_aqwms, comid, logTN, ReferenceSite), by = "comid") %>% 
  # Remove years from table so colnames can match EPA table
  rename(msst.20XX = msst2008,
         mwst.20XX = mwst2008,
         inorgnwetdep20XXws = inorgnwetdep2008ws,
         pctagslpmid20XXws = pctagslpmid2001ws,
         pctagslphigh20XXws = pctagslphigh2001ws,
         pcthbwet20XXws = pcthbwet2001ws,
         pctbl20XXws = pctbl2001ws,
         pctconif20XXws = pctconif2001ws,
         pctcrop20XXws = pctcrop2001ws,
         pctdecid20XXws = pctdecid2001ws,
         pctmxtfst20XXws = pctmxfst2001ws,
         pctfire20XXws = pctfire2002ws,
         pctfrstloss20XXws = pctfrstloss2002ws,
         pctgrs20XXws = pctgrs2001ws,
         pcthay20XXws = pcthay2001ws,
         pctice20XXws = pctice2001ws,
         pctimp20XXws = pctimp2001ws,
         pctow20XXws = pctow2001ws,
         pctshrb20XXws = pctshrb2001ws,
         pcturbop20XXws = pcturbop2001ws,
         pctwdwet20XXws = pctwdwet2001ws,
         ) %>% 
  select(1, sort(names(.))) %>% 
  relocate(logTN, .before = everything()) %>% 
  mutate(rt.master = ReferenceSite) %>% 
  select(-c(comid, ReferenceSite)) %>% 
  mutate(rt.master = case_when(
    rt.master == "MOST DISTURBED" ~ "T",
    rt.master == "MODERATELY DISTURBED" ~ "S",
    rt.master == "REFERENCE" ~ "B")) %>% 
  mutate(rt.master = factor(rt.master, levels = c("B","S","T")))
  
```

## C. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(deq_TN, function(x) sum(is.na(x))))

deq_TN$mwst.20XX <- na.roughfix(deq_TN$ mwst.20XX) 
deq_TN$bankfulldepth <- na.roughfix(deq_TN$bankfulldepth) 
deq_TN$bankfullwidth <- na.roughfix(deq_TN$bankfullwidth) 

```

```{r}
# HISTOGRAM
hist(deq_TN$logTN, main = "", xlab = "log(TN)" )

# BOX PLOT & STRIP CHART
boxplot(deq_TN$logTN, ylab = "logTN")
```

## D. Make predictions

```{r}
red11_deq <- predict(red11, newdata= deq_TN, seed = 24)

# Mean squared error (untransformed)
red11_mse_deq <- tail(red11_deq$err.rate,1)
red11_per.var_deq <- 100 * (1 - red11_mse_deq/ var(deq_TN$logTN, na.rm = TRUE))

# RMSE
red11_rmse_deq<- rmse(exp(red11_deq$predicted), exp(deq_TN$logTN))
deq_range_TN <- max(exp(deq_TN$logTN)) - min(exp(deq_TN$logTN))
red11_rmse.scaled_deq <- red11_rmse_deq/deq_range_TN
print(paste0('OOB RMSE = ' , round(red11_rmse_deq,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red11_rmse.scaled_deq*100,2), "%"))

# Make a df for plots
red11_plotdf_deq <- data.frame(obs = deq_TN$logTN, pred = red11_deq$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red11_plotdf_deq, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red11 Model predictions on DEQ data",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y2, label = paste0('OOB RMSE = ' , round(red11_rmse_deq,2), ' | scaled RMSE = ', round(red11_rmse.scaled_deq*100,2), "%"), size = 3)

# Plot residuals - red11 model, adjusted human-influenced vars to 0
red11_resid_deq <- exp(deq_TN$logTN) - exp(red11_deq$predicted) # Observed minus predicted oob
plot(red11_resid_deq ~ exp(red11_deq$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "red11 DEQ predictions residuals")
abline(h=0, col = "red", lty = 2)
```
## E. Reduce human-influenced variables to 0 and make predictions on DEQ data

```{r}
deq_TN_no.hi <- deq_TN %>% 
  mutate(across(all_of(HI_vars), ~0))

red11_deq_no.hi <- predict(red11, newdata= deq_TN_no.hi, seed = 24)

# Mean squared error (untransformed)
red11_mse_deq_no.hi  <- tail(red11_deq_no.hi $err.rate,1)
red11_per.var_deq_no.hi  <- 100 * (1 - red11_mse_deq_no.hi / var(deq_TN_no.hi$logTN, na.rm = TRUE))

# RMSE
red11_rmse_deq_no.hi <- rmse(exp(red11_deq_no.hi$predicted), exp(deq_TN_no.hi$logTN))

red11_rmse.scaled_deq_no.hi  <- red11_rmse_deq_no.hi /deq_range_TN
print(paste0('OOB RMSE = ' , round(red11_rmse_deq_no.hi ,2), units))
print(paste0('Scaled OOB RMSE = ' , round(red11_rmse.scaled_deq_no.hi *100,2), "%"))

# Make a df for plots
red11_plotdf_deq_no.hi <- data.frame(obs = deq_TN_no.hi$logTN, pred = red11_deq_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red11_plotdf_deq_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "TN red11 Model predictions on DEQ data no HI",
       x = paste("Predicted values", units),
       y = paste("Observed values", units)) +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = x, y = y2, label = paste0('OOB RMSE = ' , round(red11_rmse_deq_no.hi,2), ' | scaled RMSE = ', round(red11_rmse.scaled_deq_no.hi*100,2), "%"), size = 3)

# Plot residuals - red11 model, adjusted human-influenced vars to 0
red11_resid_deq_no.hi  <- exp(deq_TN_no.hi$logTN) - exp(red11_deq_no.hi$predicted) # Observed minus predicted oob
plot(red11_resid_deq_no.hi ~ exp(red11_deq_no.hi$predicted), xlab = paste("Predicted TN", units), ylab = "Residuals (obs - pred)", main = "red11 DEQ predictions residuals no HI")
abline(h=0, col = "red", lty = 2)
```

```{r Add predictions to compare boxplot}
# DEQ obs and predictions
any(duplicated(deq_sc_all$elevws)) # false

# Join predictions to StreamCat table
deq_obs.predict <- cbind(red11_plotdf_deq, deq_sc_all) %>% 
  arrange(elevws)

deq_aqwms <- inner_join(TN_awqms,deq_sc_all, by = join_by(COMID == comid)) %>% 
  arrange(elevws) %>% 
  select(c(MLocID:TN))

# No HI deq predictions
deq_obs.predict_no.hi <- cbind(red11_plotdf_deq_no.hi, deq_TN) %>% 
  arrange(elevws)

deq_map <- cbind(deq_aqwms, deq_obs.predict, deq_obs.predict_no.hi %>% select(pred_no.hi = pred))

# Construct a dataframe of predictions to compare
deq_boxplot_preds <- data.frame(MLocID = deq_map$MLocID, Obs = deq_map$obs, Pred_deq = deq_map$pred, Pred_no.hi = deq_map$pred_no.hi, L3Eco = as.character(deq_map$L3Eco))

deq_boxplot_preds$L3Eco <- factor(deq_boxplot_preds$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

deq_boxplot_df <- gather(deq_boxplot_preds, key = "Model", value = "Prediction", Obs, Pred_deq, Pred_no.hi)
deq_boxplot_df$Model <- factor(deq_boxplot_df$Model, levels = c("Pred_no.hi", "Pred_deq", "Obs"))

# Clean out environment
rm=(c("deq_boxplot_preds"))

ggplot(deq_boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  labs(title = "National red11 model predictions on ORDEQ data") +
  theme_bw(base_size=14) + ylab(paste("TN", units)) +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5, size = 3
  )

```

