---
title: "EPA_test_COND"
output: html_document
---
# Taking data from EPA/Lester Yuan (ss.bugs) and plugging it into our Benchmarks code structure. No need to pull from StreamCat since ss.bugs comes with all predictor data.

# SYB = Next steps:
# 1. Look in the Supplementals for the publication and only keep the predictors listed in that table.
# 2. Drop predictors if they didn't come from StreamCat (unless they are important... if so maybe try to figure out where they came from)
# 3. Pull StreamCat data for Oregon COMIDs
# 4. Run predictions on OR data
# 5. Re-build the comparison boxplot (Obs, Full model predictions, Reduced Model predictions, Adjusted predictions, and Reference)
# 6. Rinse and repeat for other parameters (nutrients next)

# **1 - LOAD PACKAGES**

```{r load packages, warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(maps)
library(RColorBrewer)
library(htmlwidgets)
```

# **2 - IMPORT DATA**

```{r}
load("~/BioMonORDEQ/Benchmarks/Water Chemistry/EPA National Model/ss.bugs.rda")

epa_data <- as.data.frame(ss.bugs[2]) 
colnames(epa_data) <- gsub("sstest.", "", colnames(epa_data))

epa_data <- epa_data %>% 
  select(-c(ABISKOMYIA:ZUMATRICHIA)) %>%  # Select out the bugs
  relocate(site.id2, .after = site.id) %>% 
  relocate(date.col, .after = year) %>% 
  filter(protocol == "WADEABLE") # Drop boatable sites
```

```{r}
# Make EPA table look more like ORDEQ table
epa_data <- epa_data %>% 
  rename(Lat_DD = lat.dd83, 
         Long_DD = lon.dd83,
         COMID = comid,
         SampleStartDate = date.col)
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r handle NAs}
# Drop rows where result has NA
epa_data <- epa_data %>% 
  drop_na(cond)
```

#### PERFORM NA ROUGH FIX

```{r NA roughfix}
sort(sapply(epa_data %>% select(c(canaldensws:pctforest)), function(x) sum(is.na(x))))

# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH A FEW MISSING VALUES
epa_data$hort  <- na.roughfix(epa_data$hort) # 1 NA
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r plot response variable}
# HISTOGRAM
hist(epa_data$cond, main = "", xlab = "Conductivity") #raw data
hist(log(epa_data$cond), main = "", xlab = "log(Conductivity)", breaks = 15)

# BOX PLOTS
boxplot(epa_data$cond, ylab = "Conductivity")
boxplot(log(epa_data$cond), ylab = "log(Conductivity)")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r prep response data}
# Select response data of interest
COND <- epa_data %>%
  select(-c(uid:ag.eco9, COMID:anc.result, doc:xfc.alg)) %>% 
  mutate(logCOND = log(cond)) %>% 
  relocate(logCOND, .before = canaldensws) %>% 
  relocate(elevation, .before = fe2o3ws) %>% 
  select(-c(cond)) %>% 
  as.data.frame() # Make sure COND is a data frame, not a tibble

# for sorting purposes later
COND_uid <- epa_data %>%
  select(-c(site.id2:year, visit.no:huc8, us.l3code, ag.eco9:anc.result, doc:xfc.alg, canaldensws:pctforest)) %>% 
  mutate(logCOND = log(cond))

# Convert character columns to factors
COND[sapply(COND, is.character)] <- lapply(COND[sapply(COND, is.character)], as.factor)
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r full model tuning}
#Adjust the parameters below based on tuning
full_tuning <- tune(logCOND~., data = COND, seed = 1234)
full_tuning$optimal
```

#### Build full model

```{r full model}
full <- rfsrc(logCOND ~ ., data = COND, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 2)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(full_per.var)

range_COND <- max(exp(COND$logCOND)) - min(exp(COND$logCOND))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(COND$logCOND))
full_rmse.scaled <- full_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(COND$logCOND))
full_oob.rmse.scaled <- full_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = COND$logCOND, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =COND$logCOND, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 100, y = 10000, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 100, y = 14500, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 100, y = 20000, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```

#### Leaflet map of full model

```{r leaflet map of full model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, COND) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
full_map <- cbind(full_obs.predict,COND_uid) %>% 
  relocate(uid:cond, .before = obs) %>% 
  select(uid:pctforest)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -98.583, lat =39.833, zoom =4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(log(data$obs)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Observed COND: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(log(data$pred)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Predicted COND: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Obs - Pred COND: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$obs, position = "bottomright" , title = "Observed COND" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = data$obs, position = "bottomright" , title = "Predicted COND" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred COND" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, log(max(full_map$obs))))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, log(max(full_map$obs))))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map$delta),max(full_map$delta)))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//full_model_map.html", selfcontained = TRUE, title = "Full model")
```

#### Partial dependence plots for full model

```{r partial dependence plots for full model}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c(names(full_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c(names(full_vimp[9:16])), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("pctimp2019ws", "coalminedensws", "pctcrop2019ws"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Another way to do partial plots (slower but more customizable)
```{r more partial plots for full model}
# partial.obj <- partial(full, partial.xvar = "ELEVWS", partial.values = full$xvar$ELEVWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "ELEVWS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "TMAX9120WS", partial.values = full$xvar$TMAX9120WS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "TMAX9120WS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "CLAYWS", partial.values = full$xvar$CLAYWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "CLAYWS", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "PRECIP_MINUS_EVTWS", partial.values = full$xvar$PRECIP_MINUS_EVTWS)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "PRECIP_MINUS_EVTWS", ylab = "ŷ")

```

#### Leaflet map of predictors  
```{r leaflet map of selected predictors}
# map_predictors <- function(data) {
#   leaflet() %>%
#   addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
#   addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
#   addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
#   setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
#   addCircleMarkers(data = data, group = "ELEVWS", lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal(data$ELEVWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "ELEVWS: ", "</strong>", data$ELEVWS))  %>% 
#       addCircleMarkers(data = data, group = "TMAX9120WS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal2(data$TMAX9120WS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "TMAX9120WS: ", "</strong>", data$TMAX9120WS))  %>% 
#      addCircleMarkers(data = data, group = "CLAYWS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal3(data$CLAYWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "CLAYWS: ", "</strong>", data$CLAYWS))  %>% 
#      addCircleMarkers(data = data, group = "PRECIP_MINUS_EVTWS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal4(data$PRECIP_MINUS_EVTWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "PRECIP_MINUS_EVTWS: ", "</strong>", data$PRECIP_MINUS_EVTWS))  %>% 
#      addCircleMarkers(data = data, group = "PCTBL2001WS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal5(data$PCTBL2001WS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "PCTBL2001WS: ", "</strong>", data$PCTBL2001WS))  %>% 
#     hideGroup(c("TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS")) %>%  # Don't check all layers by default
#     addLayersControl(
#       baseGroups = c("Basic Map", "Terrain", "Satellite"),
#       overlayGroups = c("ELEVWS", "TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS"),
#       options = layersControlOptions(collapsed = FALSE)) %>% 
#     addLegend(pal = pal, group = "ELEVWS", values = data$ELEVWS, position = "bottomright" , title = "ELEVWS" ) %>% 
#     addLegend(pal = pal2,group = "TMAX9120WS",values = data$TMAX9120WS, position = "bottomright" , title = "TMAX9120WS" ) %>%  
#     addLegend(pal = pal3, group = "CLAYWS", values = data$CLAYWS, position = "bottomleft" , title = "CLAYWS" ) %>% 
#     addLegend(pal = pal4, group = "PRECIP_MINUS_EVTWS", values = data$PRECIP_MINUS_EVTWS, position = "bottomleft" , title = "PRECIP_MINUS_EVTWS" ) %>% 
#     addLegend(pal = pal5, group = "PCTBL2001WS", values = data$PCTBL2001WS, position = "bottomleft" , title = "PCTBL2001WS" )
# }
# 
# pal <- colorNumeric(palette = "Oranges", 
#   domain = c(min(full_map$ELEVWS),max(full_map$ELEVWS)))
# pal2 <- colorNumeric(palette = "RdBu", 
#   domain = c(min(full_map$TMAX9120WS), max(full_map$TMAX9120WS)), reverse = TRUE)
# pal3 <- colorNumeric(palette = "Reds", 
#                      domain = c(min(full_map$CLAYWS), max(full_map$CLAYWS)))
# pal4 <- colorNumeric(palette = "Blues", 
#   domain = c(min(full_map$PRECIP_MINUS_EVTWS), max(full_map$PRECIP_MINUS_EVTWS)))
# pal5 <- colorNumeric(palette = "Greys", 
#   domain = c(min(full_map$PCTBL2001WS), max(full_map$PCTBL2001WS)))
# 
# map_predictors(full_map)
# 
# # Save map as html
# # predictors_map <- map_predictors(full_map)
# # saveWidget(predictors_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//predictors_map.html", selfcontained = TRUE, title = "Predictors map")
```

#### Residual plots
```{r full model residual plots}
# Plot residuals 
full_resid <- exp(COND$logCOND) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(COND$logCOND) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r adj full model}
# Reduce all human-influenced vars to all 0s.
COND_no.hi <- COND %>% 
  mutate(across(all_of(c("canaldensws", "coalminedensws", "damnrmstorws", "manurews", "minedensws", "minedenswsrp100", "npdesdensws","pctag2006slp10ws", "pctcrop2019ws", "pcthay2019ws", "pctimp2019ws", "pctnonagintrodmanagvegws","pcturbop2019ws", "rdcrsws", "superfunddensws", "superfunddenswsrp100", "tridenswsrp100")), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(COND_no.hi$logCOND))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model predictions without human influenced vars",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x =100, y = 10000, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```















# Not edited beyond this point.  5/22/26 -SYB










#### Leaflet map adjusting human vbls to 0

```{r leaflet map adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to cal_val table
full_obs.predict_no.hi <- cbind(full_plotdf_no.hi, COND_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
full_map_no.hi <- cbind(full_awqms,full_obs.predict_no.hi)

# Then select to remove duplicate logCOND column and reorder.
full_map_no.hi <- full_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, max(full_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, max(full_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map_no.hi$delta), max(full_map_no.hi$delta)))
map_function(full_map_no.hi)

# Save map as html
# full_model_map_no_hi <- map_function(full_map_no.hi)
# saveWidget(full_model_map_no_hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//full_model_map_no_hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

## C. Model Reduction

#### Build model with Top 10 predictors

```{r reduced model tuning}
#Adjust the parameters below based on tuning 
red10_tuning <- tune(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS + L2Eco + PRECIP_MINUS_EVTWS + PCTBL2001WS + PCTCONIF2001WS + SEPTICWS + PCTFRSTLOSS2002WS , data = COND, seed = 1234)
red10_tuning$optimal
```

```{r reduced model}
red10 <- rfsrc(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS + L2Eco + PRECIP_MINUS_EVTWS + PCTBL2001WS + PCTCONIF2001WS + SEPTICWS + PCTFRSTLOSS2002WS , data = COND, mtry = red10_tuning$optimal[[2]], nodesize = red10_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red10 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red10_oob.mse <- tail(red10$err.rate,1)
red10_per.var <- 100 * (1 - red10_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(red10_per.var)

#In-bag RMSE
red10_rmse<- rmse(exp(red10$predicted), exp(COND$logCOND))
red10_rmse.scaled <- red10_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(red10_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(red10_rmse.scaled*100,2), "%"))

#OOB RMSE
red10_oob.rmse<- rmse(exp(red10$predicted.oob), exp(COND$logCOND))
red10_oob.rmse.scaled <- red10_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(red10_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red10_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red10$err.rate

# VIP
plot(red10, plots.one.page=FALSE)
red10_vimp <- sort(red10$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red10_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Reduced (10 variables) model VIMP")

#print(vimp(red10)$importance)

# Make a df for plots
red10_pred.oob <- data.frame(obs = COND$logCOND, pred = red10$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red10_pred.ib <- data.frame(obs =COND$logCOND, pred = red10$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red10_plotdf <- rbind(red10_pred.oob, red10_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red10_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Red10 Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.7, y = 60, label = paste0('OOB RMSE = ' , round(red10_oob.rmse,2), ' | scaled RMSE = ', round(red10_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.7, y = 77, label = paste0('IB RMSE = ' , round(red10_rmse,2), ' | scaled RMSE = ', round(red10_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.7, y = 100, label = paste0('% variance explained = ' , round(red10_per.var,2), '%'), size = 3) 
```

#### Leaflet map of red10 model

```{r leaflet map of reduced model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
red10_obs.predict <- cbind(red10_plotdf, COND) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

red10_awqms <- COND_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  select(MLocID:logCOND) 

# Join obs.pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
red10_map <- cbind(red10_obs.predict,red10_awqms)

# Then select to remove duplicate logCOND column and reorder.
red10_map <- red10_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, red10_map$obs))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, red10_map$obs))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(red10_map$delta),max(red10_map$delta)))

map_function(red10_map)

# Save map as html
# red10_model_map <- map_function(red10_map)
# saveWidget(red10_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//red10_model_map.html", selfcontained = TRUE, title = "red10 model")
```

#### Overlay reduced and full models

```{r reduced and full model prediction plot}
full_red_map <- left_join(full_map, red10_map %>% select(MLocID, pred), by = "MLocID") %>% 
  mutate(full_pred = pred.x) %>% 
  mutate(red10_pred = pred.y) %>% 
  select(-c(pred.x,pred.y)) %>% 
  relocate(red10_pred, .after = obs) %>% 
  relocate(full_pred, .after = obs) %>% 
  select(MLocID:predictions) %>% 
  gather(key = "Model", value = "Prediction", full_pred, red10_pred)

ggplot(full_red_map, aes(x = Prediction, y = obs, color = Model)) +
  geom_point(size = 1.5, alpha = 0.5) + theme_bw(base_size = 15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Predicted COND ((μS/cm))",
    y = "Observed COND ((μS/cm))") + scale_x_log10() + scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```

#### Distributions for Human-influenced vars at all reference sites

```{r distributions of human influenced variables at all ref sites}
# Select only the human influenced predictors at reference sites from the reduced model.
COND_ref_hi <- full_map %>% 
  filter(ReferenceSite == "REFERENCE") %>% 
  select(c(MLocID:logCOND, SEPTICWS, PCTFRSTLOSS2002WS, PCTURBOP2001WS, CANALDENSWS, PCTNONAGINTRODMANAGVEGWS))

# PCTFRSTLOSS2002WS
par(mfrow=c(1,2))
plot.variable(red10, xvar.names = c("PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(COND_ref_hi$PCTFRSTLOSS2002WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(COND_ref_hi$PCTFRSTLOSS2002WS, main = '')
quantile(COND_ref_hi$PCTFRSTLOSS2002WS, probs = seq(.1, .9, by = .1))

# SEPTICWS
par(mfrow=c(1,2))
plot.variable(red10, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
abline(v=quantile(COND_ref_hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
hist(COND_ref_hi$SEPTICWS, main = '')
quantile(COND_ref_hi$SEPTICWS, probs = seq(.1, .9, by = .1))

# BELOW not in reduced model

# # PCTURBOP2001WS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("PCTURBOP2001WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$PCTURBOP2001WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$PCTURBOP2001WS, main = '')
# quantile(COND_ref_hi$PCTURBOP2001WS, probs = seq(.1, .9, by = .1))
# 
# # CANALDENSWS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("CANALDENSWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$CANALDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$CANALDENSWS, main = '')
# quantile(COND_ref_hi$CANALDENSWS, probs = seq(.1, .9, by = .1))
# 
# # PCTNONAGINTRODMANAGVEGWS
# par(mfrow=c(1,2))
# plot.variable(red10, xvar.names = c("PCTNONAGINTRODMANAGVEGWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, main = '')
# quantile(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = seq(.1, .9, by = .1))
```

#### Reduce all human-influenced variables [to 0]* and make predictions using reduced model.
##### * We may decide to adjust to a non-zero percentile based on reference quantiles in the chunk above.

```{r adj reduced model}
# Make predictions with adjusted human-influence to 0 
red10_no.hi <- predict(red10, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
red10_mse_no.hi <- tail(red10_no.hi$err.rate,1)
red10_per.var_no.hi <- 100 * (1 - red10_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
red10_rmse_no.hi<- rmse(exp(red10_no.hi$predicted), exp(COND_no.hi$logCOND))
red10_rmse.scaled_no.hi <- red10_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(red10_rmse_no.hi,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red10_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red10_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = red10_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red10_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND red10 Model predictions without human influenced vars",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.6, y = 90, label = paste0('OOB RMSE = ' , round(red10_rmse_no.hi,2), ' | scaled RMSE = ', round(red10_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red10 model, adjusted human-influenced vars to 0
red10_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(red10_no.hi$predicted) # Observed minus predicted oob
plot(red10_resid_no.hi ~ exp(red10_no.hi$predicted), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals (obs - pred)", main = "Red10 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map after adjusting human vbls for reduced10 model

```{r leaflet map after adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to cal_val table
red10_obs.predict_no.hi <- cbind(red10_plotdf_no.hi, COND_no.hi) %>% 
  arrange(ELEVWS) # Use unique col

# Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
red10_map_no.hi <- cbind(full_awqms,red10_obs.predict_no.hi)

# Then select to remove duplicate logCOND column and reorder.
red10_map_no.hi <- red10_map_no.hi %>% 
  select(MLocID:delta, BFIWS:L3Eco)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red10_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red10_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1* max(red10_map_no.hi$delta), max(red10_map_no.hi$delta)))

map_function(red10_map_no.hi)

# Save map as html
# red10_model_map_no.hi <- map_function(red10_map_no.hi)
# saveWidget(red10_model_map_no.hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//red10_model_map_no.hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```
#### Partial dependence plots for red10 model

```{r partial dependence plots for reduced model}
# Top importance vars
red10_partialplots <- plot.variable(red10, xvar.names = c(names(red10_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

red10_partialplots2 <- plot.variable(red10, xvar.names = c(names(red10_vimp[9:10])), partial = TRUE, plots.one.page = TRUE)

par(mar = c(10, 5, 2, 2))
red10_partialplots2 <- plot.variable(red10, xvar.names = c("L2Eco", "L3Eco"), partial = TRUE, las = 2, cex.axis = 0.7, plots.one.page = TRUE)

# Human-influenced vars
red10_partialplots_hi <- plot.variable(red10, xvar.names = c("SEPTICWS", "PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Top 4 predictors (not using this one)

```{r reducing the model further tuning}
#Adjust the parameters below based on tuning
red4_tuning <- tune(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = COND, seed = 1234)
red4_tuning$optimal
```

```{r reducing the model further}
red4 <- rfsrc(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = COND, mtry = red4_tuning$optimal[[2]], nodesize = red4_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red4 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red4_oob.mse <- tail(red4$err.rate,1)
red4_per.var <- 100 * (1 - red4_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(red4_per.var)

#In-bag RMSE
red4_rmse<- rmse(exp(red4$predicted), exp(COND$logCOND))
red4_rmse.scaled <- red4_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(red4_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(red4_rmse.scaled*100,2), "%"))

#OOB RMSE
red4_oob.rmse<- rmse(exp(red4$predicted.oob), exp(COND$logCOND))
red4_oob.rmse.scaled <- red4_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(red4_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red4_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red4$err.rate

# VIP
plot(red4, plots.one.page=FALSE)
red4_vimp <- sort(red4$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red4_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Reduced (4 variables) model VIMP")

#print(vimp(red4)$importance)

# Make a df for plots
red4_pred.oob <- data.frame(obs = COND$logCOND, pred = red4$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red4_pred.ib <- data.frame(obs =COND$logCOND, pred = red4$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red4_plotdf <- rbind(red4_pred.oob, red4_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red4_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND red4 Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.7, y = 60, label = paste0('OOB RMSE = ' , round(red4_oob.rmse,2), ' | scaled RMSE = ', round(red4_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.7, y = 75, label = paste0('IB RMSE = ' , round(red4_rmse,2), ' | scaled RMSE = ', round(red4_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.7, y = 100, label = paste0('% variance explained = ' , round(red4_per.var,2), '%'), size = 3) 
```

## E. Reference-Only Model

#### Subset the data

```{r subset reference only data}
COND.ref <- COND_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, CONDide, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMDENSWS", "FERTWS", "MINEDENSWS", "NANIWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")) %>% # Remove anthropogenic StreamCat vars
  as.data.frame()

# Convert character columns to factors
COND.ref[sapply(COND.ref, is.character)] <- lapply(COND.ref[sapply(COND.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r tuning ref only model}
#Adjust the parameters below based on tuning 
ref_tuning <- tune(logCOND~., COND.ref, seed = 2)
ref_tuning$optimal

```

#### Build full ref-only model

```{r ref only model}
ref <- rfsrc(logCOND ~ ., data = COND.ref, mtry = ref_tuning$optimal[[2]], nodesize = ref_tuning$optimal[[1]], importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(COND.ref$logCOND, na.rm = TRUE))
print(ref_per.var)

ref_range_COND <- max(exp(COND.ref$logCOND)) - min(exp(COND.ref$logCOND))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(COND.ref$logCOND))
ref_rmse.scaled <- ref.rmse/ref_range_COND
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(COND.ref$logCOND))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_COND
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = COND.ref$logCOND, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =COND.ref$logCOND, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Reference-Only Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.75, y = 15, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.75, y = 19, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.75, y = 25, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(COND.ref$logCOND) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Partial dependence plots for full ref-only model

```{r partial dependence plots for full ref only model}
# Top importance vars
partialplots <- plot.variable(ref, xvar.names = c(names(ref_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of Ref-Only model

```{r leaflet map of ref only model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, COND.ref) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

ref_awqms <- COND_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  filter(ReferenceSite=='REFERENCE') %>% 
  select(MLocID:logCOND) 

# Join obs.predict and awqms tables, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
ref_map <- cbind(ref_obs.predict, ref_awqms)

# Then select to remove duplicate logCOND column and reorder.
ref_map <- ref_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal2 <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(ref_map$delta), max(ref_map$delta)))

map_function(ref_map)

# Save map as html
# ref_model_map <- map_function(ref_map)
# saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## F. Make comparisons of predictions

#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r comopare reduced model (adjusted and unadjusted) by ecoregion}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(MLocID = red10_map$MLocID, StationDes = red10_map$StationDes, Obs = red10_map$obs, Pred_reduced = red10_map$pred, L3Eco = as.character(red10_map$L3Eco),Pred_adj = red10_map_no.hi$pred, Ref = red10_map$ReferenceSite, Pred_full = full_map$pred)

boxplot_ref<- ref_map %>%
  select(MLocID, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="MLocID")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_full, Pred_reduced, Pred_adj, Pred_ref, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_reduced", "Pred_full", "Obs"))
boxplot_df$L3Eco <- factor(boxplot_df$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  theme_bw(base_size=14) + ylab("COND ((μS/cm))") +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5
  )
  
# Ignore warning for 540 rows removed - these are NAs in the Pred_ref column.
```

#### Combined Leaflet map of predictions

```{r combinded leaflet map of predictions}
leaflet_df <- left_join(boxplot_df, COND %>%  select(MLocID, Project1, Lat_DD, Long_DD), by = "MLocID") %>% 
  pivot_wider(names_from = Model, values_from = Prediction) %>% 
  relocate(Obs, .before= Pred_unadj)

map_all <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$Obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Observed COND: ", "</strong>", data$Obs))  %>% 
      addCircleMarkers(data = data, group = "Pred_unadj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$Pred_unadj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND UNADJUSTED (reduced model): ", "</strong>", data$Pred_unadj))  %>% 
     addCircleMarkers(data = data, group = "Pred_adj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$Pred_adj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND ADJUSTED: ", "</strong>", data$Pred_adj))  %>% 
     addCircleMarkers(data = data, group = "Pred_ref",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$Pred_ref), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND REF ONLY: ", "</strong>", data$Pred_ref))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Pred_unadj", "Pred_adj", "Pred_ref"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$Obs, position = "bottomright" , title = "Observed COND" ) %>% 
    addLegend(pal = pal2,group = "Pred_unadj",values = data$Obs, position = "bottomright" , title = "Predicted COND (UNADJUSTED)" ) %>%  # Keep values=data$Obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Pred_adj", values = data$Pred_adj, position = "bottomleft" , title = "Predicted COND (ADJUSTED)" ) %>% 
        addLegend(pal = pal4, group = "Pred_ref", values = data$Pred_adj, position = "bottomleft" , title = "Predicted COND (REF ONLY)" )    # Keep values=data$Pred_adj here (instead of data$Ref) to keep the legend colors scales the same between ref expectations.
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal3 <- colorNumeric(palette = "Oranges", c(0, max(leaflet_df$Pred_unadj)))
pal4 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_unadj)), na.color = "transparent") 

map_all(leaflet_df)

# Save map as html
# map_all <- map_all(leaflet_df)
# saveWidget(map_all, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//map_all.html", selfcontained = TRUE, title = "Map_All")
```
 
#### Unadjusted predictions vs. adjusted human inflence 

```{r unadjusted predictions vs adj human influence}
leaflet_df$Ref <- factor(leaflet_df$Ref, levels = c("REFERENCE", "MODERATELY DISTURBED", "MOST DISTURBED"))

ggplot(leaflet_df, aes(x = Pred_unadj, y = Pred_adj, color = Ref, shape = Ref)) +
  geom_point(size = 1.5) + theme_bw(base_size = 15) +
  labs(
    x = "Reduced model predicted COND ((μS/cm))",
    y = "Predicted COND ((μS/cm)) adjusted\n for no human disturbance predictions",
    color = "Reference Status", shape = "Reference Status") + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```
## G. EXPERIMENTATION with the models
##### Take all this with a grain of salt!  (No CONDide pun intended)

#### Remove L2 and L3 Ecoregion as predictors
```{r}
# Select response data of interest
COND_noEco <- COND_vars_selected %>%
  select(-c(MLocID:COMID, CONDide, SampleStartDate, L2Eco, L3Eco, ELEVWS)) %>%  #drop CONDide in favor of logCOND. Drop AWQMS columns too. Drop Ecoregions.
  as.data.frame() # Make sure COND is a data frame, not a tibble

# Convert character columns to factors
COND_noEco[sapply(COND_noEco, is.character)] <- lapply(COND_noEco[sapply(COND_noEco, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning
tuning <- tune(logCOND~., data = COND_noEco, seed = 123)
tuning$optimal

full_noEco <- rfsrc(logCOND ~ ., data = COND_noEco, mtry = tuning$optimal[[2]], nodesize = tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_noEco # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_noEco_oob.mse <- tail(full_noEco$err.rate,1)
full_noEco_per.var <- 100 * (1 - full_noEco_oob.mse/ var(COND_noEco$logCOND, na.rm = TRUE))
print(full_noEco_per.var)

range_COND <- max(exp(COND_noEco$logCOND)) - min(exp(COND_noEco$logCOND))

#In-bag RMSE
full_noEco_rmse<- rmse(exp(full_noEco$predicted), exp(COND_noEco$logCOND))
full_noEco_rmse.scaled <- full_noEco_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_noEco_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(full_noEco_rmse.scaled*100,2), "%"))

#OOB RMSE
full_noEco_oob.rmse<- rmse(exp(full_noEco$predicted.oob), exp(COND_noEco$logCOND))
full_noEco_oob.rmse.scaled <- full_noEco_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_noEco_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_noEco_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_noEco$err.rate

# VIP
plot(full_noEco, plots.one.page=FALSE)
full_noEco_vimp <- sort(full_noEco$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_noEco_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND full_noEco model VIMP")

#print(vimp(full_noEco)$importance)

# Make a df for plots
full_noEco_pred.oob <- data.frame(obs = COND_noEco$logCOND, pred = full_noEco$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_noEco_pred.ib <- data.frame(obs =COND_noEco$logCOND, pred = full_noEco$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_noEco_plotdf <- rbind(full_noEco_pred.oob, full_noEco_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_noEco_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND full_noEco Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_noEco_oob.rmse,2), ' | scaled RMSE = ', round(full_noEco_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_noEco_rmse,2), ' | scaled RMSE = ', round(full_noEco_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_noEco_per.var,2), '%'), size = 3) 
```

```{r}
elev_temp <- ggplot(COND, aes(x = TMAX9120WS, y = ELEVWS, color = L3Eco)) + geom_point()
elev_temp
```

```{r}
# Select response data of interest
COND_WC <- COND_vars_selected %>%
  select(-c(MLocID:COMID, CONDide, SampleStartDate)) %>%  #drop CONDide in favor of logCOND. Drop AWQMS columns too. 
  filter(L2Eco=="Western Cordillera") %>% 
  select(-c(L3Eco,L2Eco)) %>%  
  as.data.frame() # Make sure COND is a data frame, not a tibble

# Convert character columns to factors
COND_WC[sapply(COND_WC, is.character)] <- lapply(COND_WC[sapply(COND_WC, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning 
tuning <- tune(logCOND~., data = COND_WC, seed = 123)
tuning$optimal

full_WC <- rfsrc(logCOND ~ ., data = COND_WC, mtry = tuning$optimal[[2]], nodesize = tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_WC # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_WC_oob.mse <- tail(full_WC$err.rate,1)
full_WC_per.var <- 100 * (1 - full_WC_oob.mse/ var(COND_WC$logCOND, na.rm = TRUE))
print(full_WC_per.var)

range_COND <- max(exp(COND_WC$logCOND)) - min(exp(COND_WC$logCOND))

#In-bag RMSE
full_WC_rmse<- rmse(exp(full_WC$predicted), exp(COND_WC$logCOND))
full_WC_rmse.scaled <- full_WC_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_WC_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(full_WC_rmse.scaled*100,2), "%"))

#OOB RMSE
full_WC_oob.rmse<- rmse(exp(full_WC$predicted.oob), exp(COND_WC$logCOND))
full_WC_oob.rmse.scaled <- full_WC_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_WC_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_WC_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_WC$err.rate

# VIP
plot(full_WC, plots.one.page=FALSE)
full_WC_vimp <- sort(full_WC$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_WC_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND full_WC model VIMP")

#print(vimp(full_WC)$importance)

# Make a df for plots
full_WC_pred.oob <- data.frame(obs = COND_WC$logCOND, pred = full_WC$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_WC_pred.ib <- data.frame(obs =COND_WC$logCOND, pred = full_WC$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_WC_plotdf <- rbind(full_WC_pred.oob, full_WC_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_WC_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND full_WC Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_WC_oob.rmse,2), ' | scaled RMSE = ', round(full_WC_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_WC_rmse,2), ' | scaled RMSE = ', round(full_WC_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_WC_per.var,2), '%'), size = 3) 
```


```{r}
COND_CP <- COND_vars %>% 
  select(c(MLocID:COMID,CANALDENSWS,  DAMDENSWS,FERTWS,  NPDESDENSWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTCROP2001WS,  PCTIMP2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PESTIC1997WS, POPDEN2010WS, PRECIP_MINUS_EVTWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, SEPTICWS, WWTPALLDENSWS, WWTPMAJORDENSWS, Coast_dist)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCOND=log(CONDide)) %>% 
  relocate(logCOND, .before = CONDide) %>% 
  select(-c(MLocID:COMID, CONDide, SampleStartDate)) %>%  #drop CONDide in favor of logCOND. Drop AWQMS columns too. 
  as.data.frame() # Make sure COND is a data frame, not a tibble

# Convert character columns to factors
COND_CP[sapply(COND_CP, is.character)] <- lapply(COND_CP[sapply(COND_CP, is.character)], as.factor)

# Build model

#Adjust the parameters below based on tuning 
tuning <- tune(logCOND~., data = COND_CP, seed = 123)
tuning$optimal

full_CP <- rfsrc(logCOND ~ ., data = COND_CP, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 160)
full_CP # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_CP_oob.mse <- tail(full_CP$err.rate,1)
full_CP_per.var <- 100 * (1 - full_CP_oob.mse/ var(COND_CP$logCOND, na.rm = TRUE))
print(full_CP_per.var)

range_COND <- max(exp(COND_CP$logCOND)) - min(exp(COND_CP$logCOND))

#In-bag RMSE
full_CP_rmse<- rmse(exp(full_CP$predicted), exp(COND_CP$logCOND))
full_CP_rmse.scaled <- full_CP_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_CP_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(full_CP_rmse.scaled*100,2), "%"))

#OOB RMSE
full_CP_oob.rmse<- rmse(exp(full_CP$predicted.oob), exp(COND_CP$logCOND))
full_CP_oob.rmse.scaled <- full_CP_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_CP_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_CP_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full_CP$err.rate

# VIP
plot(full_CP, plots.one.page=FALSE)
full_CP_vimp <- sort(full_CP$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_CP_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND full_CP model VIMP")

#print(vimp(full_CP)$importance)

# Make a df for plots
full_CP_pred.oob <- data.frame(obs = COND_CP$logCOND, pred = full_CP$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_CP_pred.ib <- data.frame(obs =COND_CP$logCOND, pred = full_CP$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_CP_plotdf <- rbind(full_CP_pred.oob, full_CP_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_CP_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND full_CP Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.8, y = 80, label = paste0('OOB RMSE = ' , round(full_CP_oob.rmse,2), ' | scaled RMSE = ', round(full_CP_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.8, y = 100, label = paste0('IB RMSE = ' , round(full_CP_rmse,2), ' | scaled RMSE = ', round(full_CP_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.8, y = 125, label = paste0('% variance explained = ' , round(full_CP_per.var,2), '%'), size = 3) 

# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND_CP$Coast_dist))  # Coast dist works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
full_CP_obs.predict <- cbind(full_CP_plotdf, COND_CP) %>% 
  filter(predictions == "OOB") %>% 
  arrange(Coast_dist) # Order by unique col

full_CP_awqms <- COND_vars_selected %>% 
  arrange(Coast_dist) %>% # Order by unique col
  select(MLocID:logCOND) 

# Join obs.pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
full_CP_map <- cbind(full_CP_obs.predict,full_CP_awqms)

# Then select to remove duplicate logCOND column and reorder.
full_CP_map <- full_CP_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, full_CP_map$obs))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, full_CP_map$obs))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(full_CP_map$delta),max(full_CP_map$delta)))

map_function(full_CP_map)

COND_CP_no.hi <- COND_CP %>% 
  mutate(across(all_of(c("CANALDENSWS", "DAMDENSWS", "FERTWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS",  "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS",  "WWTPMAJORDENSWS", "WWTPALLDENSWS")), ~0))

# Make predictions with adjusted human-influence to 0 
full_CP_no.hi <- predict(full_CP, newdata= COND_CP_no.hi, seed = 24)

# Mean squared error (untransformed)
full_CP_mse_no.hi <- tail(full_CP_no.hi$err.rate,1)
full_CP_per.var_no.hi <- 100 * (1 - full_CP_mse_no.hi/ var(COND_CP_no.hi$logCOND, na.rm = TRUE))

# RMSE
full_CP_rmse_no.hi<- rmse(exp(full_CP_no.hi$predicted), exp(COND_CP_no.hi$logCOND))
full_CP_rmse.scaled_no.hi <- full_CP_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(full_CP_rmse_no.hi,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_CP_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_CP_plotdf_no.hi <- data.frame(obs = COND_CP_no.hi$logCOND, pred = full_CP_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_CP_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND_CP full_CP Model predictions without human influenced vars",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x =0.9, y = 100, label = paste0('OOB RMSE = ' , round(full_CP_rmse_no.hi,2), ' | scaled RMSE = ', round(full_CP_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full_CP model, adjusted human-influenced vars to 0
full_CP_resid_no.hi <- exp(COND_CP_no.hi$logCOND) - exp(full_CP_no.hi$predicted) # Observed minus predicted oob
plot(full_CP_resid_no.hi ~ exp(full_CP_no.hi$predicted), xlab = "Predicted COND_CP ((μS/cm))", ylab = "Residuals (obs - pred)", main = "full_CP model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)

```

