---
title: "EPA_test_COND"
output: html_document
---
# Taking data from EPA/Lester Yuan (ss.bugs) and plugging it into our Benchmarks code structure. No need to pull from StreamCat since ss.bugs comes with all predictor data.

# SYB = Next steps:
# 1. Look in the Supplementals for the publication and only keep the predictors listed in that table.
# 2. Drop predictors if they didn't come from StreamCat (unless they are important... if so maybe try to figure out where they came from)
# 3. Pull StreamCat data for Oregon COMIDs
# 4. Run predictions on OR data
# 5. Re-build the comparison boxplot (Obs, Full model predictions, Reduced Model predictions, Adjusted predictions, and Reference)
# 6. Rinse and repeat for other parameters (nutrients next)

# **1 - LOAD PACKAGES**

```{r load packages, warning=FALSE}
library(tidyverse)
library(StreamCatTools) 
library(readxl)
library(writexl)
library(randomForest)
library(ORDEQBioassessment)
library(EnvStats)
library(Metrics)
library(randomForestSRC)
library(leaflet)
library(maps)
library(RColorBrewer)
library(htmlwidgets)
```

# **2 - IMPORT DATA**

```{r}
load("~/BioMonORDEQ/Benchmarks/Water Chemistry/EPA National Model/ss.bugs.rda")

epa_data <- as.data.frame(ss.bugs[2]) 
colnames(epa_data) <- gsub("sstest.", "", colnames(epa_data))

epa_data <- epa_data %>% 
  select(-c(ABISKOMYIA:ZUMATRICHIA)) %>%  # Select out the bugs
  relocate(site.id2, .after = site.id) %>% 
  relocate(date.col, .after = year) %>% 
  filter(protocol == "WADEABLE") %>%  # Drop boatable sites
  select(-c(fstfrz.pt,kffact.pt,rhmean.pt,pctcarbonatews,pctsandstone.carbonatews,pctsandstonews,pctsemiconsol.sandws,pctvolcanicws,elev.delt,hort,bfi.gam,bfi.rf,ref2))  # Per 6/22 communication w Lester Yuan, drop these vars as they are no longer present in StreamCat.

# Also per LY and the paper, add other SC vars
comids <- epa_data$comid

ws<-sc_get_data(comids, metric = c('bfi','damdens','pctagslphigh2006','precip9120'), aoi = 'ws')
# Not sure why these numbers are all negative?? Multiplying by -1 to fix it
other<--sc_get_data(comids, metric = c('bankfulldepth','bankfullwidth','mwst2014'), aoi = 'other') * -1 

epa_data <- epa_data %>% 
  left_join(ws, by = "comid") %>% 
  left_join(other, by = "comid") %>% 
  select(-c(ppt, wdsum.pt))

rm(list=c("ws", "other","ss.bugs"))
```

```{r}
# Make EPA table look more like ORDEQ table
epa_data <- epa_data %>% 
  rename(Lat_DD = lat.dd83, 
         Long_DD = lon.dd83,
         COMID = comid,
         SampleStartDate = date.col) %>% 
  select(uid:xfc.alg, order(colnames(.))) %>% 
  relocate(rt.master, .after=everything())
```

# **3 - PREPARE DATA FOR MODELS**

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r handle NAs}
# Drop rows where result has NA
epa_data <- epa_data %>% 
  drop_na(cond)
```

#### PERFORM NA ROUGH FIX

```{r NA roughfix}
sort(sapply(epa_data %>% select(c(bankfulldepth:rt.master)), function(x) sum(is.na(x))))

# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH A FEW MISSING VALUES
epa_data$bfiws  <- na.roughfix(epa_data$bfiws) # 56 NA
epa_data$damdensws <- na.roughfix(epa_data$damdensws) # 56 NA
epa_data$pctagslphigh2006ws <- na.roughfix(epa_data$pctagslphigh2006ws) # 56 NA
epa_data$precip9120ws <- na.roughfix(epa_data$precip9120ws) # 56 NA
epa_data$bankfulldepth <- na.roughfix(epa_data$bankfulldepth) # 96 NA
epa_data$bankfullwidth <- na.roughfix(epa_data$bankfullwidth) # 96 NA
epa_data$mwst2014 <- na.roughfix(epa_data$mwst2014) # 147 NA
```

# **4 - EXPLORE DATA**

## A. RESPONSE VARIABLE

```{r plot response variable}
# HISTOGRAM
hist(epa_data$cond, main = "", xlab = "Conductivity") #raw data
hist(log(epa_data$cond), main = "", xlab = "log(Conductivity)", breaks = 15)

# BOX PLOTS
boxplot(epa_data$cond, ylab = "Conductivity")
boxplot(log(epa_data$cond), ylab = "log(Conductivity)")
```

# **5 - RANDOM FOREST MODELS**

## A. PREP RESPONSE DATA

```{r prep response data}
# Select response data of interest
COND <- epa_data %>%
  select(-c(uid:ag.eco9, COMID:anc.result, doc:xfc.alg)) %>% 
  mutate(logCOND = log(cond)) %>% 
  relocate(logCOND, .before = everything()) %>% 
  select(-c(cond)) %>% 
  as.data.frame() # Make sure COND is a data frame, not a tibble

# for sorting purposes later
COND_uid <- epa_data %>%
  select(c(uid, COMID,cond,Lat_DD,Long_DD,SampleStartDate)) %>% 
  mutate(logCOND = log(cond))

# Convert character columns to factors
COND[sapply(COND, is.character)] <- lapply(COND[sapply(COND, is.character)], as.factor)
```

## B. FULL MODEL - ALL PREDICTORS INCLUDED

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r full model tuning}
#Adjust the parameters below based on tuning
full_tuning <- tune(logCOND~., data = COND, seed = 1234)
full_tuning$optimal
```

#### Build full model

```{r full model}
full <- rfsrc(logCOND ~ ., data = COND, mtry = full_tuning$optimal[[2]], nodesize = full_tuning$optimal[[1]], importance = "TRUE", seed = 2)
full # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (transformed)
full_oob.mse <- tail(full$err.rate,1)
full_per.var <- 100 * (1 - full_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(full_per.var)

range_COND <- max(exp(COND$logCOND)) - min(exp(COND$logCOND))

#In-bag RMSE
full_rmse<- rmse(exp(full$predicted), exp(COND$logCOND))
full_rmse.scaled <- full_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(full_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(full_rmse.scaled*100,2), "%"))

#OOB RMSE
full_oob.rmse<- rmse(exp(full$predicted.oob), exp(COND$logCOND))
full_oob.rmse.scaled <- full_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(full_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
#full$err.rate

# VIP
plot(full, plots.one.page=FALSE)
full_vimp <- sort(full$importance, decreasing = TRUE)[1:20]
par(mar=c(5,8.5,4,1)+.1)
barplot(full_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full model VIMP")

#print(vimp(full)$importance)

# Make a df for plots
full_pred.oob <- data.frame(obs = COND$logCOND, pred = full$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
full_pred.ib <- data.frame(obs =COND$logCOND, pred = full$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
full_plotdf <- rbind(full_pred.oob, full_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model",
       x = "Predicted values (μS/cm)",
       y = "Observed values (μS/cm)") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 100, y = 10000, label = paste0('OOB RMSE = ' , round(full_oob.rmse,2), ' | scaled RMSE = ', round(full_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 100, y = 14500, label = paste0('IB RMSE = ' , round(full_rmse,2), ' | scaled RMSE = ', round(full_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 100, y = 20000, label = paste0('% variance explained = ' , round(full_per.var,2), '%'), size = 3) 
```
#### Partial dependence plots for full model

```{r partial dependence plots for full model}
# Top importance vars
partialplots <- plot.variable(full, xvar.names = c(names(full_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

partialplots2 <- plot.variable(full, xvar.names = c(names(full_vimp[9:16])), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
partialplots_hi <- plot.variable(full, xvar.names = c("pctimp2019ws", "coalminedensws", "pcthay2019ws"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```
#### Another way to do partial plots (slower but more customizable)
```{r more partial plots for full model}
# partial.obj <- partial(full, partial.xvar = "pctimp2019ws", partial.values = full$xvar$pctimp2019ws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctimp2019ws", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "runoffws", partial.values = full$xvar$runoffws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "runoffws", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "pctforest", partial.values = full$xvar$pctforest)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctforest", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "ppt", partial.values = full$xvar$ppt)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "ppt", ylab = "ŷ")
# 
# partial.obj <- partial(full, partial.xvar = "pctcrop2019ws", partial.values = full$xvar$pctcrop2019ws)
# pdta <- get.partial.plot.data(partial.obj)
# plot(pdta$x, exp(pdta$yhat), type = "b", pch = 16, xlab = "pctcrop2019ws", ylab = "ŷ")
```

#### Leaflet map of full model

```{r leaflet map of full model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Join predictions to StreamCat table
full_obs.predict <- cbind(full_plotdf, COND) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
full_map <- cbind(full_obs.predict,COND_uid) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:rt.master)

map_function <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -98.583, lat =39.833, zoom =4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(log(data$obs)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Observed COND: ", "</strong>", data$obs))  %>% 
      addCircleMarkers(data = data, group = "Predicted",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(log(data$pred)), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State Description: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Predicted COND: ", "</strong>", data$pred))  %>% 
     addCircleMarkers(data = data, group = "Obs-Pred",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$delta), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "site.id: ","</strong>", data$site.id, "<br>",
                                  "<strong>", "State: ", "</strong>", data$state.nm, "<br>",
                                  "<strong>", "Obs - Pred COND: ", "</strong>", data$delta))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Predicted", "Obs-Pred"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = log(data$obs), position = "bottomright" , title = "Observed logCOND" ) %>% 
    addLegend(pal = pal2,group = "Predicted",values = log(data$obs), position = "bottomright" , title = "Predicted logCOND" ) %>%  # Keep values=data$obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Obs-Pred", values = data$delta, position = "bottomleft" , title = "Obs - Pred COND" )
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, log(max(full_map$obs))))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, log(max(full_map$obs))))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map$delta),max(full_map$delta)))

map_function(full_map)

# Save map as html
# full_model_map <- map_function(full_map)
# saveWidget(full_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//full_model_map.html", selfcontained = TRUE, title = "Full model")
```
#### Leaflet map of predictors  
```{r leaflet map of selected predictors}
# map_predictors <- function(data) {
#   leaflet() %>%
#   addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
#   addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
#   addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
#   setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
#   addCircleMarkers(data = data, group = "ELEVWS", lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal(data$ELEVWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "ELEVWS: ", "</strong>", data$ELEVWS))  %>% 
#       addCircleMarkers(data = data, group = "TMAX9120WS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal2(data$TMAX9120WS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "TMAX9120WS: ", "</strong>", data$TMAX9120WS))  %>% 
#      addCircleMarkers(data = data, group = "CLAYWS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal3(data$CLAYWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "CLAYWS: ", "</strong>", data$CLAYWS))  %>% 
#      addCircleMarkers(data = data, group = "PRECIP_MINUS_EVTWS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal4(data$PRECIP_MINUS_EVTWS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "PRECIP_MINUS_EVTWS: ", "</strong>", data$PRECIP_MINUS_EVTWS))  %>% 
#      addCircleMarkers(data = data, group = "PCTBL2001WS",lng = ~data$Long_DD, 
#                      lat = ~data$Lat_DD, fillColor = pal5(data$PCTBL2001WS), stroke = TRUE,
#                    color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
#                    popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
#                                   "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
#                                   "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
#                                   "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
#                                   "<strong>", "Reference Status: ", "</strong>", data$ReferenceSite, "<br>",
#                                   "<strong>", "PCTBL2001WS: ", "</strong>", data$PCTBL2001WS))  %>% 
#     hideGroup(c("TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS")) %>%  # Don't check all layers by default
#     addLayersControl(
#       baseGroups = c("Basic Map", "Terrain", "Satellite"),
#       overlayGroups = c("ELEVWS", "TMAX9120WS","CLAYWS", "PRECIP_MINUS_EVTWS", "PCTBL2001WS"),
#       options = layersControlOptions(collapsed = FALSE)) %>% 
#     addLegend(pal = pal, group = "ELEVWS", values = data$ELEVWS, position = "bottomright" , title = "ELEVWS" ) %>% 
#     addLegend(pal = pal2,group = "TMAX9120WS",values = data$TMAX9120WS, position = "bottomright" , title = "TMAX9120WS" ) %>%  
#     addLegend(pal = pal3, group = "CLAYWS", values = data$CLAYWS, position = "bottomleft" , title = "CLAYWS" ) %>% 
#     addLegend(pal = pal4, group = "PRECIP_MINUS_EVTWS", values = data$PRECIP_MINUS_EVTWS, position = "bottomleft" , title = "PRECIP_MINUS_EVTWS" ) %>% 
#     addLegend(pal = pal5, group = "PCTBL2001WS", values = data$PCTBL2001WS, position = "bottomleft" , title = "PCTBL2001WS" )
# }
# 
# pal <- colorNumeric(palette = "Oranges", 
#   domain = c(min(full_map$ELEVWS),max(full_map$ELEVWS)))
# pal2 <- colorNumeric(palette = "RdBu", 
#   domain = c(min(full_map$TMAX9120WS), max(full_map$TMAX9120WS)), reverse = TRUE)
# pal3 <- colorNumeric(palette = "Reds", 
#                      domain = c(min(full_map$CLAYWS), max(full_map$CLAYWS)))
# pal4 <- colorNumeric(palette = "Blues", 
#   domain = c(min(full_map$PRECIP_MINUS_EVTWS), max(full_map$PRECIP_MINUS_EVTWS)))
# pal5 <- colorNumeric(palette = "Greys", 
#   domain = c(min(full_map$PCTBL2001WS), max(full_map$PCTBL2001WS)))
# 
# map_predictors(full_map)
# 
# # Save map as html
# # predictors_map <- map_predictors(full_map)
# # saveWidget(predictors_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//predictors_map.html", selfcontained = TRUE, title = "Predictors map")
```

#### Residual plots
```{r full model residual plots}
# Plot residuals 
full_resid <- exp(COND$logCOND) - exp(full_pred.oob$pred) # Observed minus predicted oob
plot(full_resid ~ exp(full_pred.oob$pred), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals", main = "Full Model residuals")
abline(h=0, col = "red", lty = 2)

# Resids by lat and long
ggplot(data=full_map, aes(x = Lat_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Latitude", y = "Obs - Pred")

ggplot(data=full_map, aes(x = Long_DD, y = delta)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Longitude", y = "Obs - Pred") 

# MSE by lat and long
mse <- (exp(COND$logCOND) - exp(full_pred.oob$pred))^2
plot(mse ~ full_map$Lat_DD, xlab = "Latitude", ylab = "MSE")
plot(mse ~ full_map$Long_DD, xlab = "Longitude", ylab = "MSE")
```

#### Reduce all human-influenced variables to 0 and make predictions using full model.

```{r adj full model}
# Reduce all human-influenced vars to all 0s.
COND_no.hi <- COND %>% 
  mutate(across(all_of(c("canaldensws", "coalminedensws", "damnrmstorws", "manurews", "minedensws", "minedenswsrp100", "npdesdensws","pctag2006slp10ws", "pctcrop2019ws", "pctfrstloss2013ws", "pcthay2019ws", "pctimp2019ws", "pctnonagintrodmanagvegws","pcturbop2019ws", "rdcrsws", "superfunddensws", "superfunddenswsrp100", "tridenswsrp100")), ~0))

# Make predictions with adjusted human-influence to 0 
full_no.hi <- predict(full, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
full_mse_no.hi <- tail(full_no.hi$err.rate,1)
full_per.var_no.hi <- 100 * (1 - full_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
full_rmse_no.hi<- rmse(exp(full_no.hi$predicted), exp(COND_no.hi$logCOND))
full_rmse.scaled_no.hi <- full_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(full_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
full_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = full_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = full_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Model predictions without human influenced vars",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x =100, y = 10000, label = paste0('OOB RMSE = ' , round(full_rmse_no.hi,2), ' | scaled RMSE = ', round(full_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - full model, adjusted human-influenced vars to 0
full_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(full_no.hi$predicted) # Observed minus predicted oob
plot(full_resid_no.hi ~ exp(full_no.hi$predicted), xlab = "Predicted COND (μS/cm)", ylab = "Residuals (obs - pred)", main = "full model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map adjusting human vbls to 0

```{r leaflet map adj human variables to zero}
# # CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# # Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.
# 
# # Join predictions to cal_val table
# full_obs.predict_no.hi <- cbind(full_plotdf_no.hi, COND_no.hi) %>% 
#   arrange(ELEVWS) # Use unique col
# 
# # Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
# full_map_no.hi <- cbind(full_awqms,full_obs.predict_no.hi)
# 
# # Then select to remove duplicate logCOND column and reorder.
# full_map_no.hi <- full_map_no.hi %>% 
#   select(MLocID:delta, BFIWS:L3Eco)
# 
# pal <- colorNumeric(palette = "Purples", 
#   domain = c(0, max(full_map_no.hi$obs)))
# pal2 <- colorNumeric(palette = "Purples", 
#   domain = c(0, max(full_map_no.hi$obs)))
# pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(full_map_no.hi$delta), max(full_map_no.hi$delta)))
# map_function(full_map_no.hi)
# 
# # Save map as html
# # full_model_map_no_hi <- map_function(full_map_no.hi)
# # saveWidget(full_model_map_no_hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//full_model_map_no_hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```

## C. Model Reduction

#### Build model with Top 10 predictors

```{r reduced model tuning}
#Adjust the parameters below based on tuning 
red15_tuning <- tune(logCOND ~ runoffws+pctforest+precip9120ws+pctimp2019ws+pctsilicicws+sio2ws+msst.2019+pctconif2019ws+pctice2019ws+pcteolcrsws+wsareasqkm+pctshrb2019ws+clayws+nws+coalminedensws, data = COND, seed = 1234)
red15_tuning$optimal
```

```{r reduced model}
red15 <- rfsrc(logCOND ~ runoffws+pctforest+precip9120ws+pctimp2019ws+pctsilicicws+sio2ws+msst.2019+pctconif2019ws+pctice2019ws+pcteolcrsws+wsareasqkm+pctshrb2019ws+clayws+nws+coalminedensws, data = COND, mtry = red15_tuning$optimal[[2]], nodesize = red15_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red15 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red15_oob.mse <- tail(red15$err.rate,1)
red15_per.var <- 100 * (1 - red15_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(red15_per.var)

#In-bag RMSE
red15_rmse<- rmse(exp(red15$predicted), exp(COND$logCOND))
red15_rmse.scaled <- red15_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(red15_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(red15_rmse.scaled*100,2), "%"))

#OOB RMSE
red15_oob.rmse<- rmse(exp(red15$predicted.oob), exp(COND$logCOND))
red15_oob.rmse.scaled <- red15_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(red15_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red15_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red15$err.rate

# VIP
plot(red15, plots.one.page=FALSE)
red15_vimp <- sort(red15$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red15_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Reduced (10 variables) model VIMP")

#print(vimp(red15)$importance)

# Make a df for plots
red15_pred.oob <- data.frame(obs = COND$logCOND, pred = red15$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red15_pred.ib <- data.frame(obs =COND$logCOND, pred = red15$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red15_plotdf <- rbind(red15_pred.oob, red15_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red15_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND red15 Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 100, y = 10000, label = paste0('OOB RMSE = ' , round(red15_oob.rmse,2), ' | scaled RMSE = ', round(red15_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 100, y = 14500, label = paste0('IB RMSE = ' , round(red15_rmse,2), ' | scaled RMSE = ', round(red15_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 100, y = 20000, label = paste0('% variance explained = ' , round(red15_per.var,2), '%'), size = 3) 
```

#### Partial dependence plots for red15 model

```{r partial dependence plots for reduced model}
# Top importance vars
red15_partialplots <- plot.variable(red15, xvar.names = c(names(red15_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

red15_partialplots2 <- plot.variable(red15, xvar.names = c(names(red15_vimp[9:10])), partial = TRUE, plots.one.page = TRUE)

# Human-influenced vars
red15_partialplots_hi <- plot.variable(red15, xvar.names = c("pctimp2019ws", "coalminedensws", "pcthay2019ws"), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of red15 model

```{r leaflet map of reduced model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.


# Join predictions to StreamCat table
red15_obs.predict <- cbind(red15_plotdf, COND) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
red15_map <- cbind(red15_obs.predict,COND_uid) %>% 
  relocate(uid:SampleStartDate, .before = obs) %>% 
  select(uid:rt.master)

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, log(red15_map$obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, log(red15_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1 * max(red15_map$delta),max(red15_map$delta)))

map_function(red15_map)

# Save map as html
# red15_model_map <- map_function(red15_map)
# saveWidget(red15_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//red15_model_map.html", selfcontained = TRUE, title = "red15 model")
```

#### Overlay reduced and full models

```{r reduced and full model prediction plot}
full_red_map <- left_join(full_map, red15_map %>% select(uid, pred), by = "uid") %>% 
  mutate(full_pred = pred.x) %>% 
  mutate(red15_pred = pred.y) %>% 
  select(-c(pred.x,pred.y)) %>% 
  relocate(red15_pred, .after = obs) %>% 
  relocate(full_pred, .after = obs) %>% 
  select(uid:predictions) %>% 
  gather(key = "Model", value = "Prediction", full_pred, red15_pred)

ggplot(full_red_map, aes(x = Prediction, y = obs, color = Model)) +
  geom_point(size = 1.5, alpha = 0.5) + theme_bw(base_size = 15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Predicted COND (μS/cm)",
    y = "Observed COND (μS/cm)") + scale_x_log10() + scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```









# Not edited beyond this point.  6/23/26 -SYB










#### Distributions for Human-influenced vars at all reference sites (N/A since we do not know reference status for EPA sites)

```{r distributions of human influenced variables at all ref sites}
# # Select only the human influenced predictors at reference sites from the reduced model.
# COND_ref_hi <- full_map %>% 
#   filter(ReferenceSite == "REFERENCE") %>% 
#   select(c(MLocID:logCOND, SEPTICWS, PCTFRSTLOSS2002WS, PCTURBOP2001WS, CANALDENSWS, PCTNONAGINTRODMANAGVEGWS))
# 
# # PCTFRSTLOSS2002WS
# par(mfrow=c(1,2))
# plot.variable(red15, xvar.names = c("PCTFRSTLOSS2002WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$PCTFRSTLOSS2002WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$PCTFRSTLOSS2002WS, main = '')
# quantile(COND_ref_hi$PCTFRSTLOSS2002WS, probs = seq(.1, .9, by = .1))
# 
# # SEPTICWS
# par(mfrow=c(1,2))
# plot.variable(red15, xvar.names = c("SEPTICWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$SEPTICWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$SEPTICWS, main = '')
# quantile(COND_ref_hi$SEPTICWS, probs = seq(.1, .9, by = .1))

# BELOW not in reduced model

# # PCTURBOP2001WS
# par(mfrow=c(1,2))
# plot.variable(red15, xvar.names = c("PCTURBOP2001WS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$PCTURBOP2001WS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$PCTURBOP2001WS, main = '')
# quantile(COND_ref_hi$PCTURBOP2001WS, probs = seq(.1, .9, by = .1))
# 
# # CANALDENSWS
# par(mfrow=c(1,2))
# plot.variable(red15, xvar.names = c("CANALDENSWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$CANALDENSWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$CANALDENSWS, main = '')
# quantile(COND_ref_hi$CANALDENSWS, probs = seq(.1, .9, by = .1))
# 
# # PCTNONAGINTRODMANAGVEGWS
# par(mfrow=c(1,2))
# plot.variable(red15, xvar.names = c("PCTNONAGINTRODMANAGVEGWS"), partial = TRUE, plots.one.page = TRUE)
# abline(v=quantile(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = c(0.05, 0.25, 0.5, 0.90)), col = c('blue', 'red', 'green', 'black'),lty = 'dashed')
# hist(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, main = '')
# quantile(COND_ref_hi$PCTNONAGINTRODMANAGVEGWS, probs = seq(.1, .9, by = .1))
```

#### Reduce all human-influenced variables [to 0]* and make predictions using reduced model.
##### * We may decide to adjust to a non-zero percentile based on reference quantiles in the chunk above.

```{r adj reduced model}
# Make predictions with adjusted human-influence to 0 
red15_no.hi <- predict(red15, newdata= COND_no.hi, seed = 24)

# Mean squared error (untransformed)
red15_mse_no.hi <- tail(red15_no.hi$err.rate,1)
red15_per.var_no.hi <- 100 * (1 - red15_mse_no.hi/ var(COND_no.hi$logCOND, na.rm = TRUE))

# RMSE
red15_rmse_no.hi<- rmse(exp(red15_no.hi$predicted), exp(COND_no.hi$logCOND))
red15_rmse.scaled_no.hi <- red15_rmse_no.hi/range_COND
print(paste0('OOB RMSE = ' , round(red15_rmse_no.hi,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red15_rmse.scaled_no.hi*100,2), "%"))

# Make a df for plots
red15_plotdf_no.hi <- data.frame(obs = COND_no.hi$logCOND, pred = red15_no.hi$predicted) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red15_plotdf_no.hi, aes(x = pred, y = obs)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND red15 Model predictions without human influenced vars",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.6, y = 90, label = paste0('OOB RMSE = ' , round(red15_rmse_no.hi,2), ' | scaled RMSE = ', round(red15_rmse.scaled_no.hi*100,2), "%"), size = 3)

# Plot residuals - red15 model, adjusted human-influenced vars to 0
red15_resid_no.hi <- exp(COND_no.hi$logCOND) - exp(red15_no.hi$predicted) # Observed minus predicted oob
plot(red15_resid_no.hi ~ exp(red15_no.hi$predicted), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals (obs - pred)", main = "red15 model residuals, no HI vars")
abline(h=0, col = "red", lty = 2)
```

#### Leaflet map after adjusting human vbls for reduced10 model

```{r leaflet map after adj human variables to zero}
# CREATE DF FOR MAPPING PURPOSES to explore how residuals changed on the map after scaling all HI to zero
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# # Join predictions to cal_val table
# red15_obs.predict_no.hi <- cbind(red15_plotdf_no.hi, COND_no.hi) %>% 
#   arrange(ELEVWS) # Use unique col
# 
# # Join COND.obs_predict_no.hi with COND.awqms, and makes sure the results match.
# red15_map_no.hi <- cbind(full_awqms,red15_obs.predict_no.hi)
# 
# # Then select to remove duplicate logCOND column and reorder.
# red15_map_no.hi <- red15_map_no.hi %>% 
#   select(MLocID:delta, BFIWS:L3Eco)
# 
# 

# Join predictions to StreamCat table
red15_obs.predict_no.hi <- cbind(red15_plotdf_no.hi, COND_no.hi) %>% 
  filter(predictions == "OOB") 

# Join pred with awqms table, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
red15_map_no.hi <- cbind(red15_obs.predict_no.hi,COND_uid) %>% 
  relocate(uid:cond, .before = obs) %>% 
  select(uid:pctforest)

pal <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red15_map_no.hi$obs)))
pal2 <- colorNumeric(palette = "Purples", 
  domain = c(0, max(red15_map_no.hi$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1* max(red15_map_no.hi$delta), max(red15_map_no.hi$delta)))

map_function(red15_map_no.hi)

# Save map as html
# red15_model_map_no.hi <- map_function(red15_map_no.hi)
# saveWidget(red15_model_map_no.hi, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//red15_model_map_no.hi.html", selfcontained = TRUE, title = "Full model Adjusted (no human-influenced vars)")
```


#### Top 4 predictors (not using this one)

```{r reducing the model further tuning}
#Adjust the parameters below based on tuning
red4_tuning <- tune(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = COND, seed = 1234)
red4_tuning$optimal
```

```{r reducing the model further}
red4 <- rfsrc(logCOND ~ ELEVWS + TMAX9120WS + L3Eco + CLAYWS, data = COND, mtry = red4_tuning$optimal[[2]], nodesize = red4_tuning$optimal[[1]], importance = "TRUE", seed = 24)
red4 # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
red4_oob.mse <- tail(red4$err.rate,1)
red4_per.var <- 100 * (1 - red4_oob.mse/ var(COND$logCOND, na.rm = TRUE))
print(red4_per.var)

#In-bag RMSE
red4_rmse<- rmse(exp(red4$predicted), exp(COND$logCOND))
red4_rmse.scaled <- red4_rmse/range_COND
print(paste0('In-bag RMSE = ' , round(red4_rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(red4_rmse.scaled*100,2), "%"))

#OOB RMSE
red4_oob.rmse<- rmse(exp(red4$predicted.oob), exp(COND$logCOND))
red4_oob.rmse.scaled <- red4_oob.rmse/range_COND
print(paste0('OOB RMSE = ' , round(red4_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(red4_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
red4$err.rate

# VIP
plot(red4, plots.one.page=FALSE)
red4_vimp <- sort(red4$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(red4_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Reduced (4 variables) model VIMP")

#print(vimp(red4)$importance)

# Make a df for plots
red4_pred.oob <- data.frame(obs = COND$logCOND, pred = red4$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
red4_pred.ib <- data.frame(obs =COND$logCOND, pred = red4$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
red4_plotdf <- rbind(red4_pred.oob, red4_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = red4_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND red4 Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.7, y = 60, label = paste0('OOB RMSE = ' , round(red4_oob.rmse,2), ' | scaled RMSE = ', round(red4_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.7, y = 75, label = paste0('IB RMSE = ' , round(red4_rmse,2), ' | scaled RMSE = ', round(red4_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.7, y = 100, label = paste0('% variance explained = ' , round(red4_per.var,2), '%'), size = 3) 
```

## E. Reference-Only Model

#### Subset the data

```{r subset reference only data}
COND.ref <- COND_vars_selected %>% 
  filter(ReferenceSite=="REFERENCE") %>% 
  select(-c(MLocID:COMID, CONDide, SampleStartDate))  %>%  # Remove AWQMS columns
  select(-c("CANALDENSWS", "DAMDENSWS", "FERTWS", "MINEDENSWS", "NANIWS", "NPDESDENSWS", "PCTAGDRAINAGEWS", "PCTAGSLPMID2001WS", "PCTCROP2001WS", "PCTIMP2001WS", "PCTFRSTLOSS2002WS", "PCTNONAGINTRODMANAGVEGWS", "PCTURBHI2001WS", "PCTURBLO2001WS","PCTURBOP2001WS", "PESTIC1997WS", "POPDEN2010WS", "RDCRSSLPWTDWS", "RDCRSWS", "RDDENSWS", "SEPTICWS", "SUPERFUNDDENSWS", "TRIDENSWS", "WWTPMAJORDENSWS", "WWTPALLDENSWS")) %>% # Remove anthropogenic StreamCat vars
  as.data.frame()

# Convert character columns to factors
COND.ref[sapply(COND.ref, is.character)] <- lapply(COND.ref[sapply(COND.ref, is.character)], as.factor)
```

#### Model tuning - mtry and node size (number of trees happens in next step)

```{r tuning ref only model}
#Adjust the parameters below based on tuning 
ref_tuning <- tune(logCOND~., COND.ref, seed = 2)
ref_tuning$optimal

```

#### Build full ref-only model

```{r ref only model}
ref <- rfsrc(logCOND ~ ., data = COND.ref, mtry = ref_tuning$optimal[[2]], nodesize = ref_tuning$optimal[[1]], importance = "TRUE", seed = 24)
ref # Model output. (OOB) R squared = Percent Variance Explained.

# Mean squared error (untransformed)
ref_oob.mse <- tail(ref$err.rate,1)
ref_per.var <- 100 * (1 - ref_oob.mse/ var(COND.ref$logCOND, na.rm = TRUE))
print(ref_per.var)

ref_range_COND <- max(exp(COND.ref$logCOND)) - min(exp(COND.ref$logCOND))

#In-bag RMSE
ref.rmse<- rmse(exp(ref$predicted), exp(COND.ref$logCOND))
ref_rmse.scaled <- ref.rmse/ref_range_COND
print(paste0('In-bag RMSE = ' , round(ref.rmse,2), " ((μS/cm))"))
print(paste0('Scaled In-bag RMSE = ' , round(ref_rmse.scaled*100,2), "%"))

#OOB RMSE
ref_oob.rmse<- rmse(exp(ref$predicted.oob), exp(COND.ref$logCOND))
ref_oob.rmse.scaled <- ref_oob.rmse/ref_range_COND
print(paste0('OOB RMSE = ' , round(ref_oob.rmse,2), " ((μS/cm))"))
print(paste0('Scaled OOB RMSE = ' , round(ref_oob.rmse.scaled*100,2), "%"))

# Model tuning - How many trees? (default: 500)
ref$err.rate

# VIP
plot(ref, plots.one.page=FALSE)
ref_vimp <- sort(ref$importance, decreasing = TRUE)[1:20]
par(mar=c(5,7,4,1)+.1)
barplot(ref_vimp, horiz = TRUE, cex.names = 0.5, las = 1, main = "COND Full Reference-Only Model VIMP")

#print(vimp(ref)$importance)

# Make a df for plots
ref_pred.oob <- data.frame(obs = COND.ref$logCOND, pred = ref$predicted.oob, predictions = "OOB") # The prediction is computed using only the trees in which that specific case was not used for training.
ref_pred.ib <- data.frame(obs =COND.ref$logCOND, pred = ref$predicted, predictions = "In-Bag") # This contains the predicted values for each case in the training data based on the entire ensemble (all trees). It uses the data that was "in-bag" (used to grow the trees) to determine the terminal node predictions.
ref_plotdf<- rbind(ref_pred.oob, ref_pred.ib) %>% 
  mutate(obs = exp(obs)) %>% 
  mutate(pred = exp(pred)) %>% 
  mutate(delta = obs - pred)

# Plot
ggplot(data = ref_plotdf, aes(x = pred, y = obs, color = predictions, shape = predictions)) +
  geom_point(alpha = 0.75) +
  labs(title = "COND Full Reference-Only Model",
       x = "Predicted values ((μS/cm))",
       y = "Observed values ((μS/cm))") +
  theme_bw() + 
  geom_abline() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  annotate("text", x = 0.75, y = 15, label = paste0('OOB RMSE = ' , round(ref_oob.rmse,2), ' | scaled RMSE = ', round(ref_oob.rmse.scaled*100,2), "%"), size = 3) +
    annotate("text", x = 0.75, y = 19, label = paste0('IB RMSE = ' , round(ref.rmse,2), ' | scaled RMSE = ', round(ref_rmse.scaled*100,2), "%"), size = 3) +
      annotate("text", x = 0.75, y = 25, label = paste0('% variance explained = ' , round(ref_per.var,2), '%'), size = 3) 

# Plot residuals - 
ref_resid <- exp(COND.ref$logCOND) - exp(ref$predicted) # Observed minus predicted oob
plot(ref_resid ~ exp(ref$predicted), xlab = "Predicted COND ((μS/cm))", ylab = "Residuals (obs - pred)", main = "Ref only model residuals")
abline(h=0, col = "red", lty = 2)
```

#### Partial dependence plots for full ref-only model

```{r partial dependence plots for full ref only model}
# Top importance vars
partialplots <- plot.variable(ref, xvar.names = c(names(ref_vimp[1:8])), partial = TRUE, plots.one.page = TRUE)

#******************consider adding code to look at individual metrics
```

#### Leaflet map of Ref-Only model

```{r leaflet map of ref only model}
# CREATE DF FOR MAPPING PURPOSES
# Need to join predictions to AWQMS columns, but there are no unique identifiers in the COND table. Predictions will be in the same order as the COND table initially.

# Find a StreamCat row that has ALL unique values
# any(duplicated(COND$ELEVWS))  # Elevation works.
# any(duplicated(COND$PRECIP9120WS))

# Join predictions to StreamCat table
ref_obs.predict <- cbind(ref_plotdf, COND.ref) %>% 
  filter(predictions == "OOB") %>% 
  arrange(ELEVWS) # Order by unique col

ref_awqms <- COND_vars_selected %>% 
  arrange(ELEVWS) %>% # Order by unique col
  filter(ReferenceSite=='REFERENCE') %>% 
  select(MLocID:logCOND) 

# Join obs.predict and awqms tables, and makes sure the COND results match (Make sure logCOND column is the same as logCOND.1)
ref_map <- cbind(ref_obs.predict, ref_awqms)

# Then select to remove duplicate logCOND column and reorder.
ref_map <- ref_map %>% 
  select(obs:COMID) %>% 
  relocate(MLocID:COMID, .before = obs)

pal <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal2 <- colorNumeric(palette = "BuGn", 
  domain = c(0, max(ref_map$obs)))
pal3 <- colorNumeric(palette = "RdBu", domain = c(-1*max(ref_map$delta), max(ref_map$delta)))

map_function(ref_map)

# Save map as html
# ref_model_map <- map_function(ref_map)
# saveWidget(ref_model_map, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//ref_model_map.html", selfcontained = TRUE, title = "Ref-only model")
```

## F. Make comparisons of predictions

#### Compare reduced model (adjusted and unadjusted) by ecoregion

```{r comopare reduced model (adjusted and unadjusted) by ecoregion}
# Construct a dataframe of predictions to compare
boxplot_preds <- data.frame(MLocID = red15_map$MLocID, StationDes = red15_map$StationDes, Obs = red15_map$obs, Pred_reduced = red15_map$pred, L3Eco = as.character(red15_map$L3Eco),Pred_adj = red15_map_no.hi$pred, Ref = red15_map$ReferenceSite, Pred_full = full_map$pred)

boxplot_ref<- ref_map %>%
  select(MLocID, Pred_ref = pred)

boxplot_combo <- left_join(boxplot_preds, boxplot_ref, by="MLocID")

boxplot_df <- gather(boxplot_combo, key = "Model", value = "Prediction", Pred_full, Pred_reduced, Pred_adj, Pred_ref, Obs)
boxplot_df$Model <- factor(boxplot_df$Model, levels = c("Pred_ref", "Pred_adj","Pred_reduced", "Pred_full", "Obs"))
boxplot_df$L3Eco <- factor(boxplot_df$L3Eco, levels = c("Coast Range", "Willamette Valley", "Klamath Mountains", "Cascades", "Eastern Cascades Slopes and Foothills", "Columbia Plateau","Blue Mountains", "Northern Basin and Range", "Snake River Plain"))

# Clean out environment
rm=(c("boxplot_preds", "boxplot_ref", "boxplot_combo"))

ggplot(boxplot_df, aes(x = L3Eco, y = Prediction, fill = Model)) + 
  geom_boxplot() +
  guides(x = guide_axis(angle = 45)) +
  scale_y_log10() +
  theme_bw(base_size=14) + ylab("COND ((μS/cm))") +
   stat_summary(
    fun.data = function(x) data.frame(y = median(x), label = length(x)),
    geom = "text",
    position = position_dodge(width = 0.75),
    vjust = -0.5
  )
  
# Ignore warning for 540 rows removed - these are NAs in the Pred_ref column.
```

#### Combined Leaflet map of predictions

```{r combinded leaflet map of predictions}
leaflet_df <- left_join(boxplot_df, COND %>%  select(MLocID, Project1, Lat_DD, Long_DD), by = "MLocID") %>% 
  pivot_wider(names_from = Model, values_from = Prediction) %>% 
  relocate(Obs, .before= Pred_unadj)

map_all <- function(data) {
  leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "Basic Map") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  setView(lng = -120.5583, lat =44.0671, zoom =6.4) %>%
  addCircleMarkers(data = data, group = "Observed", lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal(data$Obs), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Observed COND: ", "</strong>", data$Obs))  %>% 
      addCircleMarkers(data = data, group = "Pred_unadj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal2(data$Pred_unadj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND UNADJUSTED (reduced model): ", "</strong>", data$Pred_unadj))  %>% 
     addCircleMarkers(data = data, group = "Pred_adj",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal3(data$Pred_adj), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND ADJUSTED: ", "</strong>", data$Pred_adj))  %>% 
     addCircleMarkers(data = data, group = "Pred_ref",lng = ~data$Long_DD, 
                     lat = ~data$Lat_DD, fillColor = pal4(data$Pred_ref), stroke = TRUE,
                   color = "#000", weight = 0.5, fillOpacity = 2, radius = 3, 
                   popup = paste0("<strong>", "MLocID: ","</strong>", data$MLocID, "<br>",
                                  "<strong>", "Station Description: ", "</strong>", data$StationDes, "<br>",
                                  "<strong>", "Project: ", "</strong>", data$Project1, "<br>",
                                  "<strong>", "Level 3 Ecoregion: ", "</strong>", data$L3Eco, "<br>",
                                  "<strong>", "Reference Status: ", "</strong>", data$Ref, "<br>",
                                  "<strong>", "Predicted COND REF ONLY: ", "</strong>", data$Pred_ref))  %>% 
    addLayersControl(
      baseGroups = c("Basic Map", "Terrain", "Satellite"),
      overlayGroups = c("Observed","Pred_unadj", "Pred_adj", "Pred_ref"),
      options = layersControlOptions(collapsed = FALSE)) %>% 
    addLegend(pal = pal, group = "Observed", values = data$Obs, position = "bottomright" , title = "Observed COND" ) %>% 
    addLegend(pal = pal2,group = "Pred_unadj",values = data$Obs, position = "bottomright" , title = "Predicted COND (UNADJUSTED)" ) %>%  # Keep values=data$Obs here (instead of data$pred) to keep the legend colors scales the same between obs and pred.
        addLegend(pal = pal3, group = "Pred_adj", values = data$Pred_adj, position = "bottomleft" , title = "Predicted COND (ADJUSTED)" ) %>% 
        addLegend(pal = pal4, group = "Pred_ref", values = data$Pred_adj, position = "bottomleft" , title = "Predicted COND (REF ONLY)" )    # Keep values=data$Pred_adj here (instead of data$Ref) to keep the legend colors scales the same between ref expectations.
}

pal <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal2 <- colorNumeric(palette = "Greens", 
  domain = c(0, max(leaflet_df$Obs)))
pal3 <- colorNumeric(palette = "Oranges", c(0, max(leaflet_df$Pred_unadj)))
pal4 <- colorNumeric(palette = "Oranges", domain = c(0, max(leaflet_df$Pred_unadj)), na.color = "transparent") 

map_all(leaflet_df)

# Save map as html
# map_all <- map_all(leaflet_df)
# saveWidget(map_all, file = "//deqlab1//biomon//projects//Biomon Redux//2026//Benchmarks exploratory figures//CONDide//map_all.html", selfcontained = TRUE, title = "Map_All")
```
 
#### Unadjusted predictions vs. adjusted human inflence 

```{r unadjusted predictions vs adj human influence}
leaflet_df$Ref <- factor(leaflet_df$Ref, levels = c("REFERENCE", "MODERATELY DISTURBED", "MOST DISTURBED"))

ggplot(leaflet_df, aes(x = Pred_unadj, y = Pred_adj, color = Ref, shape = Ref)) +
  geom_point(size = 1.5) + theme_bw(base_size = 15) +
  labs(
    x = "Reduced model predicted COND ((μS/cm))",
    y = "Predicted COND ((μS/cm)) adjusted\n for no human disturbance predictions",
    color = "Reference Status", shape = "Reference Status") + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```



# **7 - Make predictions on Oregon data**

## A. STATIONS - *FROM WATER CHEMISTRY BENCHMARKS SCRIPT*

```{r}
cond <- read_excel('~/BioMonORDEQ/Benchmarks/Water Chemistry/cal.val_chem.ref.wq_2026-05-14.xlsx') %>%  # Output file from ref benchmarks script 
  filter(Char_Name == 'Conductivity') %>% 
  select(c(MLocID, Project1, StationDes, Lat_DD, Long_DD, SampleStartDate, Result_Numeric_mod, L2Eco, L3Eco, EcoRegion4, HUC8, ReferenceSite, cal_val, COMID)) %>% 
  mutate(Conductivity = Result_Numeric_mod) %>% 
  select(-Result_Numeric_mod)

comids <- cond %>% 
  select(COMID) # Get just the COMIDs for next step
```

## B. WATERSHED PREDICTORS - *FROM EPA STREAMCAT*

```{r streamcat pull}
# # Run function to query ALL StreamCat data for selected COMIDs by subcategory
source('~/BioMonORDEQ/Benchmarks/Water Chemistry/SC_pull.R')

#Pull ALL StreamCat data by subcategory and merge
ag <- get_streamcat(comids = comids, type = "Agriculture")
clim <- get_streamcat(comids = comids, type = "Climate")
dams <- get_streamcat(comids = comids, type = "Dams")
flow <- get_streamcat(comids = comids, type = "Flow")
iwi <- get_streamcat(comids = comids, type = "IWI")
cover <- get_streamcat(comids = comids, type = "Land Cover")
litho <- get_streamcat(comids = comids, type = "Lithology")
mintox <- get_streamcat(comids = comids, type = "Mining_Toxics")
riparian <- get_streamcat(comids = comids, type = "Riparian Buffer")
roads <- get_streamcat(comids = comids, type = "Roads")
urb <- get_streamcat(comids = comids, type = "Urban")
fire <- get_streamcat(comids = comids, type = "Wildfire")
other <- get_streamcat(comids = comids, type = "Others")
riparian <- get_streamcat(comids = comids, type = "Riparian Buffer")

# MERGE INTO ONE TABLE
sc_all <- list(ag, clim, dams, flow, iwi, cover, litho, mintox, roads, urb, fire, other)
sc_all <- sc_all %>% reduce(full_join, by = 'COMID')

rm(list=c('ag', 'clim', 'dams', 'flow','iwi', 'cover', 'litho', 'mintox', 'riparian', 'roads', 'urb', 'fire', 'other'))
```

#### Erodible geology, NHD slope, stream power

```{r}
# ERODIBLE VS RESISTENT GEOLOGY (KEPT ERODIBLE, SINCE RESISTENT IS INVERSE OF ERODIBLE)
erod_resist_strmpow <- sc_all %>%
  mutate(PCT_EROD = PCTALLUVCOASTWS+PCTCARBRESIDWS+PCTCOASTCRSWS+PCTCOLLUVSEDWS+PCTEOLCRSWS+PCTEOLFINEWS+PCTGLACLAKECRSWS+PCTGLACLAKEFINEWS+PCTGLACTILCLAYWS+PCTGLACTILCRSWS+PCTGLACTILLOAMWS+PCTSALLAKEWS+PCTNONCARBRESIDWS)  %>% 
  mutate(PCT_RESIST = PCTALKINTRUVOLWS+PCTEXTRUVOLWS+PCTHYDRICWS+PCTSILICICWS) 

# SLOPE FROM NHD (TAKES 1-2 MIN)
# Note: do NOT push the downloaded file to GitHub as it's too large and will result in issues. Commit the ignore instead.
erod_resist_strmpow <- ORDEQBioassessment::get_NHD_info(erod_resist_strmpow) 

# WATERSHED AREA AND STREAM POWER
erod_resist_strmpow <- erod_resist_strmpow %>% 
  mutate(STRMPOW_CAT = totdasqkm + PRECIP9120WS * slope) %>%  # WSAREASQKM from SC, instead pulling totdasqkm from get_NHD (they are the same)
  select(COMID, PCT_EROD, STRMPOW_CAT, NHD_pSLOPE, SITE_TYPE) # Drop PCT_RESIST since it's the inverse of PCT_ERODE

# JOIN WITH OTHER PARAMETERS NOT FROM STREAMCAT
cond_vars <- sc_all %>% 
  inner_join(erod_resist_strmpow, by='COMID') %>% # Join StreamCat data with erod_resist_strmpow data calulated above
  inner_join(cond, by = 'COMID') # Join those with columns from AWQMS.

# RE-ORDER COLUMNS 
cond_vars <- cond_vars %>% 
  relocate(MLocID:Conductivity, .before = 'COMID') %>% 
  relocate(L2Eco:L3Eco, .after = SITE_TYPE) %>% 
  select(-c(HUC8, EcoRegion4)) 
```

## A. HANDLE NA VALUES

#### IDENTIFY COLUMNS WITH MISSING DATA

```{r}
# FIND OUT WHICH COLUMNS HAVE NAs
sort(sapply(cond_vars, function(x) sum(is.na(x))))

# THEN WRITE CODE FOR THEM IN ROUGH FIX SECTION BELOW.

# Note: Some predictors may have a large number of missing values (e.g., bankfull width, n = 68). Consider performing rough fix only for variables with ~10 or fewer missing values. This will avoid introducing error in the correlation matrix and variable selection phase.
```

#### PERFORM NA ROUGH FIX

```{r}
# PERFORM NA ROUGHFIX FUNCTION ON METRICS WITH ONLY A FEW MISSING VALUES (Leave metrics with >10 NAs [MAST2008, BMMI0809, WETTEDWITDTH, THALWEGDEPTH, BANKFULLWIDTH, BANKFULLDEPTH] until after the correlation table is built.)
cond_vars$CLAYWS  <- na.roughfix(cond_vars$CLAYWS ) # 1 NA
cond_vars$MSST2008 <- na.roughfix(cond_vars$ MSST2008) # 1 NA
cond_vars$MWST2008 <- na.roughfix(cond_vars$ MWST2008) # 11 NAs
cond_vars$NANIWS <- na.roughfix(cond_vars$NANIWS) # 1 NA
cond_vars$NSURPWS <- na.roughfix(cond_vars$NSURPWS) # 1 NA
cond_vars$OMWS <- na.roughfix(cond_vars$OMWS) # 1 NA
cond_vars$PCTIMPSLPHIGH2001WS <- na.roughfix(cond_vars$PCTIMPSLPHIGH2001WS) # 4 NAs
cond_vars$PERMWS <- na.roughfix(cond_vars$PERMWS) # 1 NA
cond_vars$RCKDEPWS <- na.roughfix(cond_vars$RCKDEPWS) # 1 NA
cond_vars$SANDWS <- na.roughfix(cond_vars$SANDWS) # 1 NA
cond_vars$SW_FLUXWS <- na.roughfix(cond_vars$SW_FLUXWS) # 2 NA
cond_vars$WTDEPWS <- na.roughfix(cond_vars$WTDEPWS) # 1 NA
```

#### Calculate percent forest metric
```{r}

```


#### Update predictor list

```{r}
# Use Excel formula (=TEXTJOIN(", ", TRUE, range:range)) to compile list from decisions sheet and paste below.
cond_vars_selected <- cond_vars %>% 
  select(c(MLocID:COMID, BFIWS, CANALDENSWS, CAOWS, CLAYWS, COMPSTRGTHWS, DAMDENSWS, ELEVWS, FE2O3WS, HYDRLCONDWS, K2OWS, KFFACTWS, MINEDENSWS, MSST2008, NA2OWS, NHD_pSLOPE, NPDESDENSWS, NWS, OMWS, PCT_EROD, PCTAGDRAINAGEWS, PCTAGSLPMID2001WS, PCTALKINTRUVOLWS, PCTALLUVCOASTWS, PCTBL2001WS, PCTBURNAREA2002WS, PCTCOLLUVSEDWS, PCTCONIF2001WS, PCTCROP2001WS, PCTDECID2001WS, PCTEOLCRSWS, PCTEOLFINEWS, PCTEXTRUVOLWS, PCTFRSTLOSS2002WS, PCTGLACLAKECRSWS, PCTGLACLAKEFINEWS, PCTGLACTILCRSWS, PCTGRS2001WS, PCTHBWET2001WS, PCTICE2001WS, PCTIMP2001WS, PCTINCVEGRESP2002WS, PCTNONAGINTRODMANAGVEGWS, PCTOW2001WS, PCTSALLAKEWS, PCTSHRB2001WS, PCTURBHI2001WS, PCTURBLO2001WS, PCTURBOP2001WS, PCTWATERWS, PESTIC1997WS, POPDEN2010WS, PRECIP9120WS, RCKDEPWS, RDCRSSLPWTDWS, RDCRSWS, RDDENSWS, ROCKNWS, SANDWS, SEPTICWS, STRMPOW_CAT, SUPERFUNDDENSWS, SW_FLUXWS, SWS, TMAX9120WS, TRIDENSWS, WATERINPUTWS, WETINDEXWS, WETTEDWIDTH, WTDEPWS, WWTPALLDENSWS, SITE_TYPE:L3Eco)) %>% 
  relocate(COMID, .after = cal_val) %>% 
  mutate(logCOND = log(Conductivity)) %>% 
  relocate(logCOND, .before= Conductivity)
```

```{r}
# HISTOGRAM
hist(cond_vars_selected$Conductivity, main = "", xlab = "Conductivity") #raw data
hist(cond_vars_selected$logCOND, main = "", xlab = "log(Conductivity)" )

# BOX PLOT & STRIP CHART
boxplot(cond_vars_selected$Conductivity, ylab = "Conductivity")
boxplot(cond_vars_selected$logCOND, ylab = "logCond")
```